---
title: "WM-2026-02-19-supervisor-meeting"
author: "WM"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=TRUE,echo=FALSE}
# Global chunk options
if(!require("pacman")) install.packages("pacman")
p_load(readxl, tidyverse, cowplot, Hmsc, cli,knitr)
opts_chunk$set(
	fig.align = "center",
	fig.height = 5,
	fig.width = 15,
	inclue=T,echo=FALSE,
	warning=FALSE, message=FALSE
)

```

```{r helper-functions,include=TRUE,echo=FALSE}
# HELPER FUNCTIONS  -------------------------------------------------------
merge_by_rownames <- function(x,y) {
  merge <- merge(x,y,by='row.names')
  rownames(merge) <- merge$Row.names
  merge[,!colnames(merge) %in% c('Row.names')]
}
# MAKING DFs FUNCTIONS  ----------------------------------------------------
make_master_df <- function(dir,pattern){
  pattern2match <- pattern
  
  matching_folders <- list.dirs(dir, recursive = FALSE, full.names = F)
  matching_folders <- matching_folders[grepl(pattern2match,
                                             basename(matching_folders))]
  model <- matching_folders[1]
  models_nums <- as.numeric(str_extract(matching_folders, "(?<=Atlas)\\d+"))
  
  outputs <- lapply(matching_folders,function(model){
    # read objects
    MF <- readRDS(file.path(dir,
                            model,
                            'Results',
                            sprintf('%sMF.rds',
                                    model)))
    VP <- readRDS(file.path(dir,
                            model,
                            'Results',
                            sprintf('%sVP.rds',
                                    model)))
    load(file.path(dir,
                   model,
                   'Models/Fitted',
                   sprintf('HPC_samples_0250_thin_100_chains_4.Rdata')))
    m <- fitted_model$posteriors
    return(m)
  })
  names(outputs) <- models_nums
  # get enviro data
  Xs <- lapply(outputs, function(df) df$XData)
  Ys <- lapply(outputs, function(df) df$Y)
  Designs <- lapply(outputs, function(df) df$studyDesign)
  ranLevels <- lapply(outputs, function(df) df$ranLevels)
  s <- lapply(ranLevels, function(df) df$s)
  i<-1
  merged_list <- lapply(seq_along(outputs), function(i) {
    
    # Get the pieces for this specific model
    X_df <- as.data.frame(Xs[[i]])
    Y_df <- as.data.frame(Ys[[i]])
    Des  <- as.data.frame(Designs[[i]])
    ses <- as.data.frame(s[[i]]$s)
    
    head(X_df[,1:5])
    head(Y_df[,1:5])
    head(Des)
    head(ses)
    # Merge them together
    # We use '0' to merge by rownames in base R
    
    # 1. Merge the Sample-level data (X, Y, and Design)
    # merge(..., by = 0) merges by row names
    combined_samples <- merge(X_df, Y_df, by = 0) %>%
      column_to_rownames("Row.names") %>%
      merge(Des, by = 0) %>%
      column_to_rownames("Row.names")
    
    # 2. Add the Spatial data (ses)
    # Since 'ses' uses site names as rownames, we merge it with the 'site' column 
    # we just got from 'Des'
    final_df <- combined_samples %>%
      rownames_to_column("sample_id") %>%
      # Merge with ses: by.x is our column 'site', by.y is rownames of 'ses'
      merge(ses, by.x = "site", by.y = 0, all.x = TRUE) %>%
      # Put the sample ID back as the rowname
      column_to_rownames("sample_id")
    
    # 3. Clean up (Optional)
    # If you have an "(Intercept)" column from X_df that you don't need:
    final_df <- final_df %>% select(-any_of("(Intercept)"))%>%
      mutate(atlas = sapply(strsplit(rownames(.), "_"), "[", 2)) %>%
      
      # 3. REORDER: Move metadata to the front
      relocate(site, atlas, X, Y) 
    
    head(final_df)
    
    return(final_df)
  })
  
  # 4. Collapse the list into one large Master Dataframe
  master_df <- bind_rows(merged_list) %>%
    # Optional: Convert atlas to a factor for better plotting later
    mutate(atlas = factor(atlas, levels = c("1", "2", "3")))
  
}
make_merged <- function(dir,pattern){
  pattern2match <- pattern
  matching_folders <- list.dirs(dir, recursive = FALSE, full.names = F)
  matching_folders <- matching_folders[grepl(pattern2match, basename(matching_folders))]
  model <- matching_folders[1]
  
  outputs <- lapply(matching_folders,function(model){
    # read objects
    MF <- readRDS(file.path(dir,
                            model,
                            'Results',
                            sprintf('%sMF.rds',
                                    model)))
    VP <- readRDS(file.path(dir,
                            model,
                            'Results',
                            sprintf('%sVP.rds',
                                    model)))
    load(file.path(dir,
                   model,
                   'Models/Fitted',
                   sprintf('HPC_samples_0250_thin_100_chains_4.Rdata')))
    m <- fitted_model$posteriors
    # get the atlas used
    X <- m$XData
    atlas <- unique(sapply(strsplit(rownames(X),'_'),'[',2))
    # get number of occurrences in that atlas
    if(!length(atlas)==1){
      stop()
    }
    occs <- data.frame(occs = colSums(m$Y,na.rm=T),
                       atlas = atlas)
    row.names(occs) <- colnames(m$Y)
    # get the trait date
    Tr <- m$TrData
    #transpose VP
    df <-VP$vals # get vals
    df <- as.data.frame(t(df)) # transpose into dataframe
    # merge VP with species info
    Tjur <- data.frame(tjur = MF$TjurR2)
    df_rel <- Tjur$tjur*df
    spNames <- m$spNames
    row.names(Tjur) <- spNames
    TrOccs <- merge_by_rownames(Tr,occs)
    TrOccsTjur <- merge_by_rownames(TrOccs,Tjur)
    TrOccsTjurVP <- merge_by_rownames(TrOccsTjur,df_rel)
    TrOccsTjurVP
  })
  merged <- do.call(rbind,outputs)
  
  # count species per guild
  guild_counts <- merged %>%
    group_by(foraging_guild_consensus) %>%
    summarise(n_species = n()/3) %>%
    mutate(Guild_counts = paste0(foraging_guild_consensus,' (n=',n_species,')')) %>%
    ungroup()
  
  #str(merged)
  merged$Guild_counts <- merged$foraging_guild_consensus
  merged$Guild_counts <- guild_counts$Guild_counts[match(merged$foraging_guild_consensus,
                                                         guild_counts$foraging_guild_consensus)]
  return(merged)
}
make_VP_effect_sizes <- function(dir,pattern){
  pattern2match <- pattern
  
  matching_folders <- list.dirs(dir, recursive = FALSE, full.names = F)
  matching_folders <- matching_folders[grepl(pattern2match, basename(matching_folders))]
  model <- matching_folders[1]
  
  outputs <- lapply(matching_folders,function(model){
    # read objects
    MF <- readRDS(file.path(dir,
                            model,
                            'Results',
                            sprintf('%sMF.rds',
                                    model)))
    VP <- readRDS(file.path(dir,
                            model,
                            'Results',
                            sprintf('%sVP.rds',
                                    model)))
    effect_sizes <- read_excel(file.path(dir,
                                         model,
                                         'Results',
                                         sprintf('%sparameter_estimates_Beta_.xlsx',model)),
                               sheet = "Posterior mean")
    effect_sigs <- read_excel(file.path(dir,
                                        model,
                                        'Results',
                                        sprintf('%sparameter_estimates_Beta_.xlsx',model)),
                              sheet = "Pr(x>0)")
    
    load(file.path(dir,
                   model,
                   'Models/Fitted',
                   sprintf('HPC_samples_0250_thin_100_chains_4.Rdata')))
    m <- fitted_model$posteriors
    # get the atlas used
    X <- m$XData
    atlas <- unique(sapply(strsplit(rownames(X),'_'),'[',2))
    # get number of occurrences in that atlas
    if(!length(atlas)==1){
      stop()
    }
    occs <- data.frame(occs = colSums(m$Y,na.rm=T),
                       atlas = atlas)
    row.names(occs) <- colnames(m$Y)
    # get the trait date
    Tr <- m$TrData
    #transpose VP
    df <-VP$vals # get vals
    df <- as.data.frame(t(df)) # transpose into dataframe
    
    # merge VP with species info
    # 1. Prepare metrics and relative df first
    metrics <- data.frame(tjur = MF$TjurR2, auc = MF$AUC, row.names = m$spNames)
    df_rel <- metrics$tjur * df
    
    # 2. The main pipe, adding the scaled VPs 
    TrOccsTjurVP <- Tr %>%
      rownames_to_column("species") %>%
      inner_join(rownames_to_column(occs, "species"), by = "species") %>%
      inner_join(rownames_to_column(metrics, "species"), by = "species") %>%
      inner_join(rownames_to_column(as.data.frame(df_rel), "species"), by = "species") %>%
      pivot_longer(
        cols = c(tmean_year, prec_year, hh, `Random: site`,
                 starts_with("perc_")),   
        names_to = "environmental_variable",
        values_to = "VP_scaled_by_Tjur"
      )
    
    # 1. Pivot Effect Sizes to long format
    long_effects <- effect_sizes %>%
      select(-`(Intercept)`) %>%
      pivot_longer(cols = -Species, names_to = "environmental_variable", values_to = "effect_size") %>%
      rename(species = Species)
    
    # 2. Pivot Significance to long format
    long_sigs <- effect_sigs %>%
      select(-`(Intercept)`) %>%
      pivot_longer(cols = -Species, names_to = "environmental_variable", values_to = "sig_val") %>%
      rename(species = Species)
    
    # 3. Join, Mask, and Merge with your main data
    final_df <- long_effects %>%
      left_join(long_sigs, by = c("species", "environmental_variable")) %>%
      # Apply the mask: Keep only if sig <= 0.05 OR sig >= 0.95
      mutate(effect_size = ifelse(sig_val > 0.05 & sig_val < 0.95, NA, effect_size)) %>%
      select(-sig_val)
    # Join to your previous master table (assuming row names are Species)
    
    combined <- TrOccsTjurVP %>% 
      left_join(final_df,by=c("species","environmental_variable"))
    
    combined
    # adding effect sizes 
  })
  merged <- do.call(rbind,outputs)
  return(merged)
}
# PLOTTING FUNCTIONS  -----------------------------------------------------
# Guild vs environment VP across species 
guild_environment_vp_plot <- function(dir, env_var, plot_title,avg_lines=FALSE){
  merged <- make_merged(dir, pattern = "2026-02-10")
  
  v2inspect <- sym(env_var)
  
  # 1. Calculate the mean per atlas for the lines
  atlas_means <- merged %>%
    group_by(atlas) %>%
    summarise(atlas_avg = mean(!!v2inspect, na.rm = TRUE), .groups = "drop")
  
  guild_means <- merged %>%
    group_by(Guild_counts, atlas) %>%
    summarise(!!v2inspect := mean(!!v2inspect, na.rm = TRUE), .groups = "drop")
    
  colors <- c('1' = 'darkblue', '2' = 'darkgreen', '3' = 'darkred')
  
  ggplot# 1. Start the plot object
  p <- ggplot(merged,
              aes(x = reorder(Guild_counts, !!v2inspect, FUN = mean, na.rm = TRUE),
                  y = !!v2inspect,
                  col = atlas, fill = atlas))
  
  # 2. Add lines conditionally (placed first to be behind points)
  if(avg_lines == TRUE){
    p <- p + geom_hline(data = atlas_means, 
                        aes(yintercept = atlas_avg, color = atlas),
                        linewidth = 0.4, linetype = "dashed", alpha = 0.7)
  }
  
  # 3. Add the rest of the layers
  p <- p + 
    geom_point(alpha = 0.5, size = 0.5) +
    geom_point(data = guild_means,
               aes(y = !!v2inspect, fill = atlas, color = atlas),
               size = 1, shape = 21, stroke = 1) +
    scale_color_manual(values = colors) +
    scale_fill_manual(values = colors) +
    scale_y_continuous(limits = (c(0, 0.175))) +
    labs(title = plot_title) +
    xlab(label = 'Guilds (number of species)') +
    ylab(label = 'Absolute variance explained') +
    coord_flip() +
    theme_minimal() +
    theme(axis.line = element_blank())
    
  return(p)
}
# Single species VP vs effect sizes 
species_vp_effect_plots <- function(dir,species, english_name, remove_site = T){
  df <- make_VP_effect_sizes(dir,pattern = "2026-02-10")
  subset <- df[df$species == species,]
  
  if(remove_site == T) {
    subset <- subset[subset$environmental_variable != "Random: site",]
  }
  
  base_colors <- c(
    "Temperature"     = "red",
    "Precipitation"   = "blue",
    "Land-use heterogeneity"   = "pink",
    "Urban (% coverage)"           = "grey",
    "Cropland (% coverage)"        = "yellow3", # Slightly darker so it doesn't vanish
    "Pasture (% coverage)"         = "orange",
    "Forest (% coverage)"          = "darkgreen",
    "Grass/Shrubland (% coverage)" = "lightgreen"
  )
  
  # --- STEP 1: CALCULATE SCALING ---
  # We need to map effect_size into the R2 range (0 to max VP).
  # Adjust 'coeff' if your effect sizes are very large or small.
  max_vp <- max(subset$VP_scaled_by_Tjur, na.rm = TRUE)
  max_effect <- max(abs(subset$effect_size), na.rm = TRUE)
  coeff <- max_vp / max_effect 
  
  subset_clean <- subset %>%
    mutate(
      environmental_variable = case_match(environmental_variable,
                                          "tmean_year"       ~ "Temperature",
                                          "prec_year"        ~ "Precipitation",
                                          "hh"               ~ "Land-use heterogeneity",
                                          "perc_urban"       ~ "Urban (% coverage)",
                                          "perc_cropland"    ~ "Cropland (% coverage)",
                                          "perc_pasture"     ~ "Pasture (% coverage)",
                                          "perc_forest"      ~ "Forest (% coverage)",
                                          "perc_grass_shrub" ~ "Grass/Shrubland (% coverage)",
                                          .default = environmental_variable),
      environmental_variable = factor(environmental_variable, levels = c(
        "Temperature", "Precipitation", "Land-use heterogeneity", 
        "Urban (% coverage)", "Cropland (% coverage)", 
        "Pasture (% coverage)", "Forest (% coverage)", "Grass/Shrubland (% coverage)"
      )),
      atlas = factor(atlas),
      # Scale the effect size to fit the R2 axis
      effect_scaled = effect_size * coeff
    )
  
  ggplot(data = subset_clean, aes(x = environmental_variable)) +
    # Bars for Variance (Primary Axis)
    geom_bar(aes(y = VP_scaled_by_Tjur, alpha = atlas, fill = environmental_variable), 
             stat = "identity", position = "dodge") +
    
    # Points/Lines for Effect Sizes (Secondary Axis)
    geom_point(aes(y = effect_scaled, group = atlas), 
               position = position_dodge(width = 0.9), color = "black", shape = 18, size = 3) +
    
    # --- STEP 2: DUAL AXIS ---
    scale_y_continuous(
      name = "Absolute variance explained (R²)",
      labels = scales::percent,
      sec.axis = sec_axis(~ . / coeff, name = "Effect Size (Pr>0.95)")
    ) +
    
    scale_fill_manual(values = base_colors, guide = "none") +
    scale_alpha_manual(values = c("1" = 0.4, "2" = 0.7, "3" = 1.0), 
                       name = "Atlas Period",
                       labels = c("1970s", "1990s", "2010s")) +
    labs(title = english_name, subtitle = "Bars = Variance Explained | Diamonds = Effect Size", x = NULL) +
    theme_minimal(base_size = 14) +
    theme(axis.text.x = element_text(angle = 40, hjust = 1),
          legend.position = "top")
}
# Species presences in environmental space 
species_environment_plots <- function(dir,species, english_name, env_vars,
                                      x_variable,y_variable) {
  
  master_df <- make_master_df(dir,pattern="2026-02-10")
  ### NOW WE GET TO PLOTTING 
  subs_names <- c('atlas', env_vars, species)
  subs <- master_df[, colnames(master_df) %in% subs_names]
  
  v1 <- sym(env_vars[1])
  v2 <- sym(env_vars[2])
  
  xlims <- master_df %>% pull(!!v1) %>% range(na.rm = TRUE)
  ylims <- master_df %>% pull(!!v2) %>% range(na.rm = TRUE)
  
  # A. Generate a "dummy" plot just to grab the shared legend
  # We make a simple ggplot with the same alpha mapping
  legend_plot <- ggplot(subs, aes(x = !!v1, y = !!v2, alpha = factor(!!sym(species)))) +
    geom_point() +
    scale_alpha_manual(values = c("0" = 0.1, "1" = 1.0), 
                       labels = c("Absence", "Presence"),
                       name = "Status") +
    theme_minimal() +
    theme(legend.position = "bottom")
  
  shared_legend <- get_legend(legend_plot)
  
  # B. Generate the 3 actual plots (with legend = FALSE)
  colors <- c('darkblue','darkgreen','darkred')
  plots <- lapply(c(1, 2, 3), function(x) {
    single_plot(x, subs, color = colors[x], env_vars, xlims, ylims, show_legend = FALSE,x_variable,y_variable)
  })
  
  # C. Combine into a grid
  plot_row <- plot_grid(plotlist = plots, ncol = 3, labels = "AUTO", align = 'h')
  
  # D. Add the legend at the bottom
  final_combined <- plot_grid(plot_row, shared_legend, ncol = 1, rel_heights = c(1, 0.1))
  
  # Add a super title (optional)
  title_theme <- ggdraw() + draw_label(paste(english_name), fontface='bold')
  
  return(plot_grid(title_theme, final_combined, ncol = 1, rel_heights = c(0.1, 1)))
}
single_plot <- function(atlas_nr, subs, color, env_vars, xlims, ylims, show_legend = FALSE,x_variable,y_variable) {
  
  subset_data <- subs[subs$atlas == atlas_nr, ]
  colnames(subset_data)[ncol(subset_data)] <- "presenceabsence"
  subset_data$presenceabsence <- factor(subset_data$presenceabsence, levels = c("0", "1"))
  
  v1 <- sym(env_vars[1])
  v2 <- sym(env_vars[2])
  
  # MAIN PLOT
  pmain <- ggplot(data = subset_data, aes(x = !!v1, y = !!v2, alpha = presenceabsence)) +
    geom_point(col = color) +
    scale_alpha_manual(values = c("0" = 0.1, "1" = 1.0), 
                       labels = c("Absence", "Presence"),
                       name = "Status") +
    scale_x_continuous(limits = xlims) + 
    scale_y_continuous(limits = ylims) +
    theme_minimal() +
    labs(title = paste0("Atlas Period: ", atlas_nr),
         x = x_variable,
         y = y_variable) +
    # Toggle legend based on function call
    theme(legend.position = if(show_legend) "bottom" else "none")
  
  # MARGINAL DENSITIES
  xdens <- axis_canvas(pmain, axis = "x") +
    geom_density(data = subset_data, aes(x = !!v1), color = color, linetype = 3) +
    geom_density(data = subset_data[subset_data$presenceabsence == 1, ], aes(x = !!v1), linetype = 1, color = color)
  
  ydens <- axis_canvas(pmain, axis = "y", coord_flip = TRUE) +
    geom_density(data = subset_data, aes(x = !!v2), linetype = 3, color = color) +
    geom_density(data = subset_data[subset_data$presenceabsence == 1, ], aes(x = !!v2), linetype = 1, color = color) +
    coord_flip()
  
  # COMBINE
  p1 <- insert_xaxis_grob(pmain, xdens, grid::unit(.2, "null"), position = "top")
  p2 <- insert_yaxis_grob(p1, ydens, grid::unit(.2, "null"), position = "right")
  
  return(p2)
}

# Species presences & environmental variables in spatial space
species_spatial_comparison <- function(dir, species, english_name, env_var,
                                       col_low, col_high, legend_name,
                                       relative = FALSE) { # New toggle
  
  master_df <- make_master_df(dir, pattern="2026-02-10")
  v_env <- sym(env_var)
  
  # 1. Prep data
  subs_cols <- c('atlas', 'X', 'Y', env_var, species) 
  subs <- master_df[, colnames(master_df) %in% subs_cols]
  
  # 2. Handle Relative Transformation
  if(relative) {
    subs <- subs %>%
      group_by(atlas) %>%
      mutate(!!v_env := !!v_env - mean(!!v_env, na.rm = TRUE)) %>%
      ungroup()
    
    # Update legend name to reflect the change
    legend_name <- paste(legend_name, "deviation from mean for respective atlas")
  }
  
  # 3. Global limits for consistent axes and colors
  xlims <- range(subs$X, na.rm = TRUE)
  ylims <- range(subs$Y, na.rm = TRUE)
  
  # For relative plots, we often want the scale to be symmetric around 0
  if(relative) {
    max_val <- max(abs(range(subs %>% pull(!!v_env), na.rm = TRUE)))
    env_lims <- c(-max_val, max_val)
  } else {
    env_lims <- subs %>% pull(!!v_env) %>% range(na.rm = TRUE)
  }
  
  # 4. Create a list for each Atlas containing TOP and BOTTOM plots
  all_plots <- lapply(c(1, 2, 3), function(x) {
    subset_data <- subs[subs$atlas == x, ]
    colnames(subset_data)[ncol(subset_data)] <- "presenceabsence"
    
    p_top <- single_spatial_plot(subset_data, v_env, xlims, ylims, env_lims, 
                                 title = paste0("Atlas ", x, ": Presence"), 
                                 use_alpha = TRUE,
                                 col_low, col_high, legend_name)
    
    p_bottom <- single_spatial_plot(subset_data, v_env, xlims, ylims, env_lims, 
                                    title = paste0("Atlas ", x, ": Full Gradient"), 
                                    use_alpha = FALSE,
                                    col_low, col_high, legend_name)
    
    list(top = p_top, bottom = p_bottom)
  })
  
  # 5. Extraction and Grid Assembly
  top_row <- lapply(all_plots, `[[`, "top")
  bottom_row <- lapply(all_plots, `[[`, "bottom")
  
  final_grid <- plot_grid(
    plot_grid(plotlist = top_row, ncol = 3, align = 'h'),
    plot_grid(plotlist = bottom_row, ncol = 3, align = 'h'),
    ncol = 1, rel_heights = c(1, 1)
  )
  
  # Create legend
  dummy_plot <- ggplot(subs, aes(x=X, y=Y, color=!!v_env)) + 
    geom_point() +
    scale_color_gradient2(low = col_low, mid = "white", high = col_high, 
                          midpoint = 0, name = legend_name)
  
  # If not relative, stick to standard gradient; if relative, use gradient2 for a 0-midpoint
  if(!relative) {
    dummy_plot <- ggplot(subs, aes(x=X, y=Y, color=!!v_env)) + 
      geom_point() +
      scale_color_gradient(low = col_low, high = col_high, name = legend_name)
  }

  leg <- get_legend(dummy_plot + theme(legend.position = "bottom"))
  
  title <- ggdraw() + draw_label(english_name, fontface = 'bold', x = 0.5, size = 18)
  
  layout <- plot_grid(title, final_grid, leg, ncol = 1, rel_heights = c(0.1, 1, 0.1))  
  return(layout)
}

single_spatial_plot <- function(subset_data, v_env, xlims, ylims, env_lims, title, use_alpha,
                                col_low, col_high, legend_name) {
  
  if(use_alpha) {
    subset_data <- subset_data %>% arrange(presenceabsence)
    alpha_scale <- scale_alpha_manual(values = c("0" = 0.05, "1" = 1.0), guide = "none")
  } else {
    alpha_scale <- scale_alpha_manual(values = c("0" = 1.0, "1" = 1.0), guide = "none")
  }
  
  # Determine scale type: if limits are centered on 0, use a diverging scale
  is_relative <- env_lims[1] < 0 && env_lims[2] > 0
  
  color_scale <- if(is_relative) {
    scale_color_gradient2(limits = env_lims, low = col_low, mid = "white", high = col_high, midpoint = 0)
  } else {
    scale_color_gradient(limits = env_lims, low = col_low, high = col_high)
  }
  
  ggplot(subset_data, aes(x = X, y = Y)) +
    geom_point(aes(color = !!v_env, alpha = as.factor(presenceabsence)), size = 0.8) +
    alpha_scale +
    color_scale +
    coord_fixed(xlim = xlims, ylim = ylims) +
    theme_minimal() +
    theme(
      axis.text = element_blank(), axis.title = element_blank(),
      panel.grid = element_blank(), legend.position = "none"
    ) +
    labs(subtitle = title) 
}
```

Hey everyone,

I'm trying out this format as an agenda for a meeting as it allows me to easily express my thoughts / ask questions supported by figures (e.g. from raw data or model results). It saves me quite some hassle as I can code / type in one document, and any updates to code/figures immediately get saved in a file that I can share with you, without me having to reshuffle slides, figures etc. If this format doesn't work for you, let me know and I'll find an alternative.  

# Meeting agenda 
  - Update
  - Administrative questions for Carsten:
    - Can I keep working on Mjölnir throughout the rest of the PhD? Or indefinitely until I don't need it anymore and then sign off? 
    - Can I apply for the research stay abroad from Copenhagen to try and get money for the relocation towards here? It's much much more than Adelaide gives me (almost 10x!). 
  - Chapter 1:
    - Predictions from Atlas 1 to Atlas 2/3 are alright, but predictions from Atlas 2/3 to 1 are no better than random. 
    - Does the inclusion of temperature / precipitation even make sense? An oyster-catching example.

# Chapter 1
### Predictions forward / backward in time

In the plot below you see three models, each trained on one of the atlases. Each dot represents one species, The X and Y axes compare explanatory power (metrics derived from training data) with predictive power (metrics derived from testing on validation data). Black dots show predictive power from 5-fold cross-validation on the same atlas, red and blue dots show predictive power from testing on a different atlas. 

There's a quite striking pattern that the first atlas predicts 'okay' in atlas 2 and atlas 3, and atlas 2 and atlas 3 predict 'okay' on each-other but predict basically random in atlas 1. 

I've already discussed this a bit with Kasper. If I can speak on his behalf, he thinks this is driven by the fact that Atlas 1 has more variance explained by the spatial component (as we'll see below), and this is spatial component is then 'carrying the weight' for predictions in later atlases (if I understood Kasper correctly). I think this is possible and will be testing this, but I personally find the differences in accuracies too big to be explained by such 'little' differences in variance explained.

To be quite honest, I can't really explain this properly yet. I'm happy to discuss what this could mean.

```{r}

# plotting function
plot_CV_gg = function(type, modelled_atlas, cMF, cMFCV, tMF) {
  testing_atlases <- list(
    '1' = c(2,3),
    '2' = c(1,3),
    '3' = c(1,2)
  )
  
  tested_atlases <- testing_atlases[[as.character(modelled_atlas)]]
  
  # Set limits and vline/hline values
  limit = switch(type,
                 "RMSE"   = c(max(c(cMF[[type]], cMFCV[[type]]), na.rm = TRUE), 
                              max(c(cMF[[type]], cMFCV[[type]]), na.rm = TRUE), 0, 0),
                 "AUC"    = c(0, 1, 0.5, 0.5),
                 "O.AUC"  = c(0, 1, 0.5, 0.5),
                 "TjurR2" = c(0, 1, 0, 0), # Added TjurR2 limits
                 c(1, 1, 0, 0))
  
  # Determine coordinate limits
  x_min <- if(type %in% c("AUC", "O.AUC", "TjurR2")) limit[1] else -limit[1]
  y_min <- x_min
  
  # Create exact legend labels
  lab_cv    <- sprintf('CV Atlas %d', modelled_atlas)
  lab_test1 <- sprintf('Test Atlas %d', tested_atlases[1])
  lab_test2 <- sprintf('Test Atlas %d', tested_atlases[2])
  
  # Structure data for ggplot (long format)
  df_cv <- data.frame(
    Explanatory = cMF[[type]],
    Predictive  = cMFCV[[type]],
    Evaluation  = lab_cv
  )
  df_t1 <- data.frame(
    Explanatory = cMF[[type]],
    Predictive  = tMF[[1]][[type]],
    Evaluation  = lab_test1
  )
  df_t2 <- data.frame(
    Explanatory = cMF[[type]],
    Predictive  = tMF[[2]][[type]],
    Evaluation  = lab_test2
  )
  
  plot_df <- bind_rows(df_cv, df_t1, df_t2)
  
  title_str <- sprintf("%s - Atlas %d",
                       type,
                       modelled_atlas)
  subtitle_str <- sprintf("Training = %.2f, Cross-validation = %.2f\n Testing: Atlas %d = %.2f, Atlas %d = %.2f",
                       mean(cMF[[type]], na.rm = TRUE),
                       mean(cMFCV[[type]], na.rm = TRUE),
                       tested_atlases[1],
                       mean(tMF[[1]][[type]], na.rm = TRUE),
                       tested_atlases[2],
                       mean(tMF[[2]][[type]], na.rm = TRUE))
  
  # Exact color mapping matching your original script
  col_mapping <- setNames(c('black', 'red', 'blue'), c(lab_cv, lab_test1, lab_test2))
  
  # Generate ggplot
  p <- ggplot(plot_df, aes(x = Explanatory, y = Predictive, color = Evaluation)) +
    geom_point(alpha = 0.6, size = 1.5) +
    geom_abline(slope = 1, intercept = 0, linetype = "solid", color = "black", linewidth = 0.5) +
    geom_vline(xintercept = limit[3], linetype = "solid", color = "black", linewidth = 0.5) +
    geom_hline(yintercept = limit[4], linetype = "solid", color = "black", linewidth = 0.5) +
    scale_color_manual(values = col_mapping) +
    coord_cartesian(xlim = c(x_min, limit[2]), ylim = c(y_min, limit[2])) +
    labs(x = "Explanatory power", y = "Predictive power", 
         title = title_str,subtitle = subtitle_str) +
    theme_bw() +
    theme(
      legend.position = "top",
      legend.title = element_blank(),
      plot.title = element_text(hjust = 0.5)
    )
  
  return(p)
}

# 2. Set up directories and the target nested list structure
pattern2match <- "2026-02-10"
matching_folders <- list.dirs('HmscOutputs', recursive = FALSE, full.names = FALSE)
matching_folders <- matching_folders[grepl(pattern2match, basename(matching_folders))]

# Define the metrics and initialize an empty nested list: outputs[[metric]][[atlas]]
metrics2plot <- c('AUC', 'TjurR2', 'RMSE')
# Initialize storage
outputs <- sapply(metrics2plot, function(x) list(), simplify = FALSE)
summary_df <- data.frame()
raw_results_df <- data.frame() # NEW: To store raw data for boxplots

testing_atlases_map <- list('1' = c(2,3), '2' = c(1,3), '3' = c(1,2))

# 3. Outer loop: Use a standard 'for' loop to populate the nested list dynamically
for(folders2match in matching_folders) {
  models_description = folders2match
  
  localDir  = "./HmscOutputs"
  ModelDir  = file.path(localDir, sprintf("%s/Models/Fitted", models_description))
  TempDir   = file.path(localDir, sprintf("%s/Models/Temp", models_description))
  ResultDir = file.path(localDir, sprintf("%s/Results", models_description))
  TestDir   = file.path(localDir, sprintf("%s/Tests", models_description))
  
  # Extract atlas number dynamically from the folder string
  atlas_num <- as.numeric(str_extract(models_description, "(?<=Atlas)\\d+"))
  
  samples_list = c(250)
  thin_list = c(100)
  nChains = 4
  nfolds = 5
  
  thin = thin_list[1]
  samples = samples_list[1]
  
  # Check if models exist
  filename = file.path(ModelDir, sprintf("MF_samples_%.4d_thin_%.2d_chains_%.1d_nfolds_%.1d.rdata",
                                         samples, thin, nChains, nfolds))
  testfilename = file.path(TestDir, sprintf("Preds/AtlasPreds_%s_HPC_samples_%.4d_thin_%.2d_chains_%.1d.Rdata",
                                            models_description, samples, thin, nChains))
  
  if(!file.exists(filename) || !file.exists(testfilename)){
    cli_alert_danger(sprintf("Missing files for Atlas %d. Skipping...", atlas_num))
    next 
  }
  
  load(testfilename)
  load(filename)
  
  extract_fit <- function(tests, atlas_name) {
    tests[[atlas_name]]$fit_test
  }
  tMF <- lapply(names(tests), function(test) extract_fit(tests, test))
  
  # Identify which atlases were tested
  tested_atlases <- testing_atlases_map[[as.character(atlas_num)]]
  
for(metric in metrics2plot) {
    # 1. Generate and store the scatter plot
    p <- plot_CV_gg(type = metric, modelled_atlas = atlas_num, cMF = MF, cMFCV = MFCV, tMF = tMF)
    outputs[[metric]][[as.character(atlas_num)]] <- p
    
    # 2. COLLECT RAW DATA FOR BOXPLOT (NEW)
    # We create a long-format data frame of every species' value
    tmp_raw <- data.frame(
      Metric = metric,
      Modelled_Atlas = paste("Atlas", atlas_num),
      Training = as.numeric(MF[[metric]]),
      CV = as.numeric(MFCV[[metric]]),
      Test_A = as.numeric(tMF[[1]][[metric]]),
      Test_B = as.numeric(tMF[[2]][[metric]])
    ) %>%
      pivot_longer(cols = c(Training, CV, Test_A, Test_B), 
                   names_to = "Evaluation", values_to = "Value")
    
    raw_results_df <- bind_rows(raw_results_df, tmp_raw)

    # 3. Extract means for the summary table
    # ... [Keep your train_mean, cv_mean, and summary_df logic same] ...
    train_mean <- mean(MF[[metric]], na.rm = TRUE)
    cv_mean    <- mean(MFCV[[metric]], na.rm = TRUE)
    t1_mean    <- mean(tMF[[1]][[metric]], na.rm = TRUE)
    t2_mean    <- mean(tMF[[2]][[metric]], na.rm = TRUE)
    
    test_results <- c(NA, NA, NA)
    test_results[tested_atlases[1]] <- t1_mean
    test_results[tested_atlases[2]] <- t2_mean
    
    row_data <- data.frame(
      Metric = metric, Modelled_Atlas = atlas_num,
      Training = train_mean, Cross_Validation = cv_mean,
      Test_Atlas_1 = test_results[1], Test_Atlas_2 = test_results[2], Test_Atlas_3 = test_results[3]
    )
    summary_df <- bind_rows(summary_df, row_data)
  }
}

### PLOTTING 
# AUC - species points 
# Using do.call to dynamically grab all 3 atlas plots for AUC
grid_auc <- plot_grid(plotlist = outputs[['AUC']], nrow = 1)
title <- ggdraw() + draw_label("Scatter Plots: Explanatory vs Predictive (AUC)", fontface='bold')
plot_grid(title, grid_auc, ncol=1, rel_heights=c(0.1, 1))

# Metrics - Boxplots
# Re-order the Metrics
# This ensures AUC is the first panel, then TjurR2, then RMSE
raw_results_df$Metric <- factor(raw_results_df$Metric, 
                                levels = c("AUC", "TjurR2", "RMSE"))

# Re-order the X-axis (Evaluation)
# This forces Training -> CV -> Test_A -> Test_B
raw_results_df$Evaluation <- factor(raw_results_df$Evaluation, 
                                    levels = c("Training", "CV", "Test_A", "Test_B"))

# Ensure Modelled_Atlas is a factor for the alpha mapping
raw_results_df$Modelled_Atlas <- as.factor(raw_results_df$Modelled_Atlas)


### --- 2. GENERATE BOXPLOTS --- ###

boxplot_viz <- ggplot(raw_results_df, aes(x = Evaluation, y = Value, fill = Evaluation, alpha = Modelled_Atlas)) +
  geom_boxplot(outlier.alpha = 0.3) +
  # Using facet_wrap to keep the specified Metric order
  facet_wrap(~Metric, scales = "free_y", ncol = 3) +
  scale_fill_manual(values = c("Training"="black", "CV"="grey80", "Test_A"="red", "Test_B"="blue")) +
  # Adjust alpha range so the boxes aren't too transparent
  scale_alpha_manual(values = c("Atlas 1" = 0.4, "Atlas 2" = 0.7, "Atlas 3" = 1.0)) + 
  theme_minimal() +
  labs(title = "Comparing Training fit, CV, and transferability to other Atlases",
       alpha = "Training Atlas",
       fill = "Evaluation Type") +
  theme(
    legend.position = "bottom", 
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.text = element_text(face = "bold", size = 12)
  )

# Render the plots in order
print(boxplot_viz)

### --- 3. THE SUMMARY TABLE --- ###
# We also factor the summary_df so the table follows the same order as the plots
# summary_df$Metric <- factor(summary_df$Metric, levels = c("AUC", "TjurR2", "RMSE"))
# 
# summary_df <- summary_df %>%
#   arrange(Metric, Modelled_Atlas)
# 
# kable(
#   summary_df, 
#   digits = 2,
#   col.names = c("Metric", "Modelled Atlas", "Training", "Cross Validation", 
#                 "Test Atlas 1", "Test Atlas 2", "Test Atlas 3"),
#   align = "c",
#   caption = "Average accuracy metrics"
# )


```

### Variance explained by land-use / climate
These models are trained on climate / land-use data. The variance explained by these different components looks like this. The top row shows the total variance explained by both environmental and random (spatial) effects. The bottom row excludes those spatial effects to look at patterns of environmental variables and their relative / absolute variance explained through time. 

In terms of absolute variance explained, temperature, precipition and % forest cover do the most work explaining the data. Let's look into those a bit further. 

```{r}
pattern2match <- "2026-02-10"
matching_folders <- list.dirs('HmscOutputs', recursive = FALSE, full.names = FALSE)
matching_folders <- matching_folders[grepl(pattern2match, basename(matching_folders))]
model_path <- matching_folders[1]
# 1. Load and process both versions (Scaled and Unscaled)
VPs_combined <- lapply(matching_folders, function(model_path) {
  atlas_num <- as.numeric(str_extract(model_path, "(?<=Atlas)\\d+"))
  
  # Load the VP CSV
  file_to_load <- file.path('HmscOutputs', model_path, 'Results', 
                            sprintf('%sparameter_estimates_VP_.csv', model_path))
  
  df <- read.csv(file_to_load) %>% column_to_rownames(var='X')
  
  # Extract TjurR2 for scaling
  r2_values <- as.numeric(df["TjurR2", ])
  
  # Version A: Proportional (Unscaled) - just remove TjurR2 row
  df_unscaled <- df %>% filter(rownames(.) != "TjurR2")
  
  # Version B: Raw Variance (Scaled) - multiply columns by TjurR2
  df_scaled <- df_unscaled %>%
    map2_dfc(r2_values, ~ .x * .y)
  # Restore rownames lost by map2_dfc
  rownames(df_scaled) <- rownames(df_unscaled)
  
  # Calculate averages for both
  list(
    unscaled = data.frame(variable = rownames(df_unscaled), 
                          average = rowMeans(df_unscaled, na.rm = TRUE),
                          type = "Relative Variance Explained"),
    scaled   = data.frame(variable = rownames(df_scaled), 
                          average = rowMeans(df_scaled, na.rm = TRUE),
                          type = "Absolute Variance Explained")
  )
})

# 2. Combine into one master plot data frame
plot_data <- lapply(seq_along(VPs_combined), function(i) {
  atlas_id <- as.character(str_extract(matching_folders[i], "(?<=Atlas)\\d+"))
  bind_rows(VPs_combined[[i]]$unscaled, VPs_combined[[i]]$scaled) %>%
    mutate(Atlas = atlas_id)
}) %>% bind_rows()

# 3. Clean and Factorize
base_colors <- c(
  "Temperature" = "firebrick3", "Precipitation" = "dodgerblue3",
  "Land-use heterogeneity" = "maroon4", "Urban (% coverage)" = "seashell4",
  "Cropland (% coverage)" = "goldenrod1", "Pasture (% coverage)" = "orange",
  "Forest (% coverage)" = "springgreen4", "Grass/Shrubland (% coverage)" = "palegreen3",
  "Site (Spatial Random Effect)" = "cornsilk2"

)

# Clean the names and remove the TjurR2 variable itself (since it's not a bar)
plot_data_base <- plot_data %>%
  filter(variable != "TjurR2") %>%
  mutate(
    variable_clean = case_match(variable,
                                "tmean_year"       ~ "Temperature",
                                "prec_year"        ~ "Precipitation",
                                "hh"               ~ "Land-use heterogeneity",
                                "perc_urban"       ~ "Urban (% coverage)",
                                "perc_cropland"    ~ "Cropland (% coverage)",
                                "perc_pasture"     ~ "Pasture (% coverage)",
                                "perc_forest"      ~ "Forest (% coverage)",
                                "perc_grass_shrub" ~ "Grass/Shrubland (% coverage)",
                                "Random: site"     ~ "Site (Spatial Random Effect)",
                                .default = variable),
    variable_clean = factor(variable_clean, levels = names(base_colors)),
    atlas = factor(Atlas),
    type = factor(type, levels = c("Relative Variance Explained", 
                                   "Absolute Variance Explained"))
  )



# We want Atlas on the X-axis to see the total model composition
plot_top <- ggplot(plot_data_base, aes(x = atlas, y = average, fill = variable_clean)) +
  geom_bar(stat = "identity", position = position_stack(reverse=T), color = "white", linewidth = 0.1) +
  facet_wrap(~type, scales = "fixed") + # Same axis for comparison of total R2
  scale_fill_manual(values = base_colors,
                    guide = guide_legend(reverse = T)) +
  scale_y_continuous(labels = scales::percent) +
  theme_minimal() +
  labs(title = "Total Model Composition (Environmental + Site Effects)",
       x = "Atlas Period", y = "Variance", fill = "Variable") +
  theme(legend.position = "right", panel.grid.minor = element_blank())

# --- BOTTOM ROW: ENVIRONMENTAL ZOOM ---
# We exclude Site and switch Variable to the X-axis
plot_bottom <- ggplot(plot_data_base %>% filter(variable_clean != "Site (Spatial Random Effect)"), 
                      aes(x = variable_clean, y = average, fill = variable_clean, alpha = atlas)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.8)) +
  facet_wrap(~type, scales = "free_y") + # free_y is essential here!
  scale_fill_manual(values = base_colors, guide = "none") +
  scale_alpha_manual(values = c("1" = 0.4, "2" = 0.7, "3" = 1.0), 
                     name = "Atlas Period", labels = c("1970s", "1990s", "2010s")) +
  scale_y_continuous(labels = scales::percent) +
  theme_minimal() +
  labs(title = "Detailed Environmental Drivers (Site Effect Excluded)",
       x = NULL, y = "Variance") +
  theme(axis.text.x = element_text(angle = 40, hjust = 1),
        legend.position = "right")
### --- COMBINE INTO THE FINAL LAYOUT ---
# Align the plots vertically
final_plot <- plot_grid(
  plot_top, 
  plot_bottom, 
  ncol = 1, 
  rel_heights = c(1, 1.2), # Give bottom plot slightly more room for X-axis labels
  align = 'v', 
  axis = 'lr'
)

plot_top
plot_bottom
```

### Temperature 
When we look at a bit closer detail at variance explained by temperature across species, we find that plovers on average have the highest variance explained by temperature, across all the guilds. 

```{r model-fit-by-guild-absolute,fig.height = 5, fig.width = 7.5}
### Set directory to end up in 'HmscOutputs' folder. 
path2HmscOutputs <- './HmscOutputs'
species <- 'Haematopus_ostralegus'
english_name <- 'Oystercatcher'
env_var <- 'tmean_year'
env_vars <- c('tmean_year','prec_year')
title <- 'Annual temperature'

guild_environment_vp_plot(dir = path2HmscOutputs,
                          env_var = env_var,
                          plot_title = title,
                          avg_lines = FALSE)


```

### Oystercatcher 
We can look a bit more in detail at one of these species. The oystercatcher is one of the plovers with very high variance explained across all models. 

This plot shows the variance explained for this species (across the atlases) and the significant effect sizes of the different variables in that atlas period. What's quite striking is that all land-use variables included here have a negative effect, and the temperature has a positive effect. That's pretty cool, but let's look at the raw data for this species. 

```{r oystercatcher-vp-effects,fig.height = 5, fig.width = 10}
# ### LOOK INTO VARIANCE EXPLAINED BY DIFFERENT VARIABLES AND ESTIMATES OF EFFECT SIZES ACROSS PERIODS FOR A SINGLE SPECIES 
species_vp_effect_plots(dir = path2HmscOutputs,
                        species = species,
                        english_name = english_name,
                        remove_site=T)
# # you'll get some errors with this one, that's correct, there are some insignificant effect sizes plotted as NAs and that 
# # throws and error
# 
# ### LOOK AT ENVIRONMENTAL SPACE OCCUPIED BY A SPECIES 
# # here you can choose two variables to look at that environmental space 
# species_environment_plots(dir = path2HmscOutputs,
#                           species=species,
#                           english_name=english_name,
#                           env_vars=env_vars)
# ### LOOK AT THIS OCCUPIED ENVIRONMENTAL SPACE ON A MAP
# # you you choose two colors, i like purple yellow for the stark contrast 
# species_spatial_comparison(dir = path2HmscOutputs,
#                            species=species,
#                            english_name=english_name, 
#                            env_var=env_var,
#                            col_low = 'purple',col_high = 'yellow',
#                            legend_name = title)

```

Some things are clear from here - the climatic space (temperature, precipitation) is completely disjunct from one atlas to the next. The temperature in atlas 3 does not overlap with the temperature in the previous atlases at all. So the fact that this species occurs in all three atlases already signals that temperature is not 'really' influencing this species' distributions in a way. But there is a clear pattern of the species occupping warmer grid cells rather than colder grid cells. So what is driving this? 

```{r oystercatcher-environmental space,fig.height = 5, fig.width = 10}
# ### LOOK INTO VARIANCE EXPLAINED BY DIFFERENT VARIABLES AND ESTIMATES OF 
species_environment_plots(dir = path2HmscOutputs,
                          species=species,
                          english_name=english_name,
                          env_vars=env_vars,
                          x_variable = 'Temperature',
                          y_variable = 'Precipitation')
# ### LOOK AT THIS OCCUPIED ENVIRONMENTAL SPACE ON A MAP
# # you you choose two colors, i like purple yellow for the stark contrast 
# species_spatial_comparison(dir = path2HmscOutputs,
#                            species=species,
#                            english_name=english_name, 
#                            env_var=env_var,
#                            col_low = 'purple',col_high = 'yellow',
#                            legend_name = title)

```

Quite likely the coastline! As we can see on the plots below, the coastlines are consistently warmer than than the middle of the country. This happens to be the locations that the oystercatchers inhabit. Overtime, they continuously move land-inwards, which maybe explains the decreasing variance explained / effect sizes that we saw two plots ago.

This reasoning maybe applies to other shore/wading birds as well, which explains why most of the guilds with highest variance explained by temperature are shore/wading birds. 

If this is the case, it seems that what we're modelling is not a species relationship to the environment, but a species' preferred habitat to the environment. If so, does it even make sense to include these variables in the model? 

Let's discuss! 

```{r oystercatcher-environment in space,fig.height = 5, fig.width = 10}

### LOOK AT THIS OCCUPIED ENVIRONMENTAL SPACE ON A MAP
# you you choose two colors, i like purple yellow for the stark contrast
species_spatial_comparison(dir = path2HmscOutputs,
                           species=species,
                           english_name=english_name,
                           env_var=env_var,
                           col_low = 'dodgerblue3',col_high = 'firebrick3',
                           legend_name = title,
                           relative=T)

```

### Forest coverage  
I'll rerun this for the % coverage of forest because there's a really nice relationship with some birds! 

```{r model-fit-by-guild-forest,fig.height = 5, fig.width = 7.5}
### Set directory to end up in 'HmscOutputs' folder. 
path2HmscOutputs <- './HmscOutputs'
species <- 'Caprimulgus_europaeus'
english_name <- 'European nightjar'
env_var <- 'perc_forest'
env_vars <- c('perc_forest','perc_cropland')
title <- '% forest'

guild_environment_vp_plot(dir = path2HmscOutputs,
                          env_var = env_var,
                          plot_title = title,
                          avg_lines = FALSE)


```

### Nightjar 
We can look a bit more in detail at the nightjar. % coverage of forest seems to consistently have a profound positive effect on nightjars, explaining a lot of variation with high effect sizes. Higher coverage of croplands have a generally negative effect on the species.

```{r nightjar-vp-effects,fig.height = 5, fig.width = 10}
# ### LOOK INTO VARIANCE EXPLAINED BY DIFFERENT VARIABLES AND ESTIMATES OF EFFECT SIZES ACROSS PERIODS FOR A SINGLE SPECIES 
species_vp_effect_plots(dir = path2HmscOutputs,
                        species = species,
                        english_name = english_name,
                        remove_site=T)
# # you'll get some errors with this one, that's correct, there are some insignificant effect sizes plotted as NAs and that 
# # throws and error
# 
# ### LOOK AT ENVIRONMENTAL SPACE OCCUPIED BY A SPECIES 
# # here you can choose two variables to look at that environmental space 
# species_environment_plots(dir = path2HmscOutputs,
#                           species=species,
#                           english_name=english_name,
#                           env_vars=env_vars)
# ### LOOK AT THIS OCCUPIED ENVIRONMENTAL SPACE ON A MAP
# # you you choose two colors, i like purple yellow for the stark contrast 
# species_spatial_comparison(dir = path2HmscOutputs,
#                            species=species,
#                            english_name=english_name, 
#                            env_var=env_var,
#                            col_low = 'purple',col_high = 'yellow',
#                            legend_name = title)

```

This is reflected in the environmental space also, mostly occupying grids with some % coverage of forest, and low % coverage of croplands. 

```{r nightjar-environmental space,fig.height = 5, fig.width = 10}
# ### LOOK INTO VARIANCE EXPLAINED BY DIFFERENT VARIABLES AND ESTIMATES OF 
species_environment_plots(dir = path2HmscOutputs,
                          species=species,
                          english_name=english_name,
                          env_vars=env_vars,
                          x_variable = 'Forest',
                          y_variable = 'Cropland')
# ### LOOK AT THIS OCCUPIED ENVIRONMENTAL SPACE ON A MAP
# # you you choose two colors, i like purple yellow for the stark contrast 
# species_spatial_comparison(dir = path2HmscOutputs,
#                            species=species,
#                            english_name=english_name, 
#                            env_var=env_var,
#                            col_low = 'purple',col_high = 'yellow',
#                            legend_name = title)

```

I really like these plots for this, as you can nicely see the 'pockets' of high % coverage of forest in each atlas, which very nicely match with the presences of the species. 

```{r nightjar-environment in space,fig.height = 5, fig.width = 10}

### LOOK AT THIS OCCUPIED ENVIRONMENTAL SPACE ON A MAP
# you you choose two colors, i like purple yellow for the stark contrast
species_spatial_comparison(dir = path2HmscOutputs,
                           species=species,
                           english_name=english_name,
                           env_var=env_var,
                           col_low = 'gold',col_high = 'forestgreen',
                           legend_name = title,
                           relative=F)

```



