---
title: "Modelled trait composition of 1970s avian bioregion clusters"
author: "WM"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  results = "hide",
  warning = FALSE,
  message = FALSE,
  fig.align = "center"
)
knitr::opts_knit$set(root.dir = normalizePath("../.."))
```

## Aim

This notebook summarizes which kinds of species make up the 1970s modelled
avian bioregion clusters. These are the Atlas 1 reference clusters from the
avian bioregion PCA analysis, and the composition estimates use HMSC modelled
species probabilities only.

For each 1970s cluster, predicted species probabilities are summed within
thermal-affinity, migratory-strategy, and foraging-guild groups. These sums are
then divided by the total expected richness among species with known values for
that trait family. The reported percentages therefore describe the expected
trait composition of the modelled 1970s community, not raw species counts.

## Load Model Outputs And Traits

```{r load-data}
if (!require("pacman")) install.packages("pacman")
pacman::p_load(
  tidyverse, Hmsc, scales, patchwork, RColorBrewer, knitr
)

source(file.path("support_scripts", "figure_data_helpers.R"))

base_dir <- "./HmscOutputs"
pattern <- "2026-03-13"
bioregion_assignment_path <- file.path(
  "notebooks", "exploratory", "outputs", "bioregion-pca",
  paste0(pattern, "-bioregion-assignments.csv")
)
out_dir <- file.path("notebooks", "exploratory", "outputs", "bioregion-trait-composition")
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

year_lookup <- c("1" = "1970s", "2" = "1990s", "3" = "2010s")
focal_atlas <- "1"
focal_period <- year_lookup[[focal_atlas]]
output_prefix <- paste0(pattern, "-", focal_period)
thermal_levels <- c("very cold", "cold", "medium", "warm", "very warm")
migration_levels <- c(
  "long-distance",
  "short-and long-distance",
  "short-distance",
  "sedentary and short-distance",
  "sedentary"
)
thermal_palette <- setNames(
  colorRampPalette(c("blue4", "darkorchid", "red2"))(length(thermal_levels)),
  thermal_levels
)
migration_palette <- setNames(
  colorRampPalette(c("#542788", "#f2e5ff"))(length(migration_levels)),
  migration_levels
)
cluster2_shift_focal_comparison <- "2010s minus 1970s"

predictor_names <- function(df) {
  c(
    "tmean_breeding",
    "prec_breeding",
    "hh",
    grep("^perc_", colnames(df), value = TRUE)
  )
}

var_labels <- c(
  "tmean_breeding" = "Temperature",
  "prec_breeding" = "Precipitation",
  "hh" = "Land-use heterogeneity",
  "perc_urban" = "Urban",
  "perc_cropland" = "Cropland",
  "perc_pasture" = "Pasture",
  "perc_forest" = "Forest",
  "perc_grass_shrub" = "Grass/Shrubland"
)

environment_colours <- c(
  "Temperature" = "firebrick3",
  "Precipitation" = "dodgerblue3",
  "Land-use heterogeneity" = "orchid3",
  "Urban" = "#4d4d4d",
  "Cropland" = "goldenrod1",
  "Pasture" = "darkorange",
  "Forest" = "springgreen4",
  "Grass/Shrubland" = "springgreen2"
)

matching_folders <- figure_model_folders(pattern = pattern, base_dir = base_dir)
model_nums <- as.character(atlas_numbers(matching_folders))
mods <- load_hmsc_posteriors(matching_folders, base_dir = base_dir)
predsY <- load_or_compute_site_predictions(mods, matching_folders, base_dir = base_dir)
names(mods) <- names(predsY) <- model_nums
predsY_all <- predsY
predsY <- predsY[focal_atlas]

bioregion_assignments_all <- read_csv(
  bioregion_assignment_path,
  show_col_types = FALSE
) |>
  mutate(
    atlas = as.character(atlas),
    period = recode(atlas, !!!year_lookup),
    bioregion = factor(bioregion, levels = sort(unique(bioregion)))
  )

bioregion_assignments <- bioregion_assignments_all |>
  filter(atlas == focal_atlas) |>
  select(atlas, period, survey, site, X, Y, bioregion)

bioregion_palette <- setNames(
  brewer.pal(max(3, length(levels(bioregion_assignments$bioregion))), "Set2")[seq_along(levels(bioregion_assignments$bioregion))],
  levels(bioregion_assignments$bioregion)
)

load("Data/preprocessed_data_min_occs_is_5_coverage_is_good_and_average.RData")

thermal_groups <- function(sti) {
  cut(
    sti,
    breaks = quantile(sti, seq(0, 1, 0.2), na.rm = TRUE),
    include.lowest = TRUE,
    labels = thermal_levels
  )
}

traits <- Tr |>
  as.data.frame() |>
  rownames_to_column("species") |>
  select(species, Migration_a3_DOF, foraging_guild_consensus, species_thermal_index) |>
  mutate(
    thermal_group = thermal_groups(species_thermal_index),
    thermal_group = factor(thermal_group, levels = thermal_levels)
  )

trait_lookup <- bind_rows(
  traits |>
    filter(!is.na(thermal_group)) |>
    transmute(
      species,
      trait_family = "Thermal affinity",
      trait_value = as.character(thermal_group)
    ),
  traits |>
    filter(!is.na(Migration_a3_DOF), Migration_a3_DOF != "") |>
    transmute(
      species,
      trait_family = "Migratory strategy",
      trait_value = Migration_a3_DOF
    ),
  traits |>
    filter(!is.na(foraging_guild_consensus), foraging_guild_consensus != "") |>
    transmute(
      species,
      trait_family = "Foraging guild",
      trait_value = foraging_guild_consensus
    )
) |>
  mutate(
    trait_family = factor(
      trait_family,
      levels = c("Thermal affinity", "Migratory strategy", "Foraging guild")
    )
  )

trait_group_sizes <- trait_lookup |>
  distinct(trait_family, trait_value, species) |>
  count(trait_family, trait_value, name = "n_species_in_trait_group")

thermal_group_summary <- traits |>
  filter(!is.na(thermal_group)) |>
  count(thermal_group, name = "n_species") |>
  complete(
    thermal_group = factor(thermal_levels, levels = thermal_levels),
    fill = list(n_species = 0)
  )
```

## Modelled Cluster Composition

```{r composition-functions}
prediction_frame <- function(pred_y, atlas) {
  pred_y |>
    as.data.frame(check.names = FALSE) |>
    rownames_to_column("survey") |>
    as_tibble() |>
    pivot_longer(
      -survey,
      names_to = "species",
      values_to = "predicted_probability"
    ) |>
    mutate(atlas = as.character(atlas), .before = 1)
}

make_trait_palette <- function(values, palette = "Set2") {
  values <- sort(unique(values))
  n_values <- length(values)
  if (n_values <= 8) {
    colours <- brewer.pal(max(3, n_values), palette)[seq_len(n_values)]
  } else {
    colours <- colorRampPalette(brewer.pal(8, palette))(n_values)
  }
  setNames(colours, values)
}
```

```{r compute-composition}
prediction_long <- imap_dfr(predsY, prediction_frame)

prediction_assignment_checks <- prediction_long |>
  distinct(atlas, survey) |>
  left_join(
    bioregion_assignments |> distinct(atlas, survey, bioregion),
    by = c("atlas", "survey")
  ) |>
  group_by(atlas) |>
  summarise(
    n_prediction_sites = n(),
    n_sites_without_cluster = sum(is.na(bioregion)),
    .groups = "drop"
  )

species_trait_checks <- tibble(
  trait_family = c("Thermal affinity", "Migratory strategy", "Foraging guild"),
  n_model_species = length(colnames(predsY[[1]])),
  n_species_with_trait = c(
    sum(colnames(predsY[[1]]) %in% (trait_lookup |> filter(trait_family == "Thermal affinity") |> pull(species))),
    sum(colnames(predsY[[1]]) %in% (trait_lookup |> filter(trait_family == "Migratory strategy") |> pull(species))),
    sum(colnames(predsY[[1]]) %in% (trait_lookup |> filter(trait_family == "Foraging guild") |> pull(species)))
  )
) |>
  mutate(n_species_missing_trait = n_model_species - n_species_with_trait)

prediction_with_clusters <- prediction_long |>
  inner_join(
    bioregion_assignments,
    by = c("atlas", "survey")
  )

composition_numerators <- prediction_with_clusters |>
  inner_join(trait_lookup, by = "species", relationship = "many-to-many") |>
  group_by(atlas, period, bioregion, trait_family, trait_value) |>
  summarise(
    expected_group_richness = sum(predicted_probability, na.rm = TRUE),
    mean_probability = mean(predicted_probability, na.rm = TRUE),
    n_species = n_distinct(species),
    n_sites = n_distinct(survey),
    .groups = "drop"
  )

composition_denominators <- composition_numerators |>
  group_by(atlas, period, bioregion, trait_family) |>
  summarise(
    expected_total_richness_known_trait = sum(expected_group_richness, na.rm = TRUE),
    .groups = "drop"
  )

cluster_trait_composition <- composition_numerators |>
  left_join(
    composition_denominators,
    by = c("atlas", "period", "bioregion", "trait_family")
  ) |>
  mutate(
    proportion = expected_group_richness / expected_total_richness_known_trait,
    percent = 100 * proportion
  ) |>
  arrange(trait_family, period, bioregion, desc(proportion))

composition_sum_checks <- cluster_trait_composition |>
  group_by(atlas, period, bioregion, trait_family) |>
  summarise(
    n_trait_groups = n(),
    percent_sum = sum(percent, na.rm = TRUE),
    max_abs_deviation_from_100 = abs(percent_sum - 100),
    expected_total_richness_known_trait = first(expected_total_richness_known_trait),
    .groups = "drop"
  )

cluster_trait_dominant_groups <- cluster_trait_composition |>
  group_by(atlas, period, bioregion, trait_family) |>
  slice_max(proportion, n = 1, with_ties = FALSE) |>
  ungroup() |>
  select(
    atlas, period, bioregion, trait_family, dominant_trait_group = trait_value,
    percent, expected_group_richness, expected_total_richness_known_trait,
    n_species, n_sites
  )

species_expected_cluster_richness <- prediction_with_clusters |>
  group_by(atlas, period, bioregion, species) |>
  summarise(
    expected_occurrence_sum = sum(predicted_probability, na.rm = TRUE),
    mean_predicted_probability = mean(predicted_probability, na.rm = TRUE),
    n_sites = n_distinct(survey),
    .groups = "drop"
  ) |>
  group_by(atlas, period, bioregion) |>
  mutate(
    expected_occurrence_share = expected_occurrence_sum / sum(expected_occurrence_sum, na.rm = TRUE),
    rank_in_cluster = dense_rank(desc(expected_occurrence_sum))
  ) |>
  ungroup()

cluster_top_modelled_species <- species_expected_cluster_richness |>
  filter(rank_in_cluster <= 15) |>
  left_join(
    traits |>
      select(
        species,
        thermal_group,
        migratory_strategy = Migration_a3_DOF,
        foraging_guild = foraging_guild_consensus,
        species_thermal_index
      ),
    by = "species"
  ) |>
  arrange(period, bioregion, rank_in_cluster)

cluster_trait_composition_checks <- bind_rows(
  composition_sum_checks |>
    mutate(check_type = "Composition percentages") |>
    rename(check_group = trait_family) |>
    select(check_type, atlas, period, bioregion, check_group, everything()),
  prediction_assignment_checks |>
    mutate(
      check_type = "Prediction-to-cluster join",
      period = focal_period,
      bioregion = NA,
      check_group = "All sites",
      percent_sum = NA_real_,
      max_abs_deviation_from_100 = NA_real_,
      expected_total_richness_known_trait = NA_real_,
      n_trait_groups = NA_integer_
    ) |>
    rename(n_sites_without_cluster = n_sites_without_cluster) |>
    select(check_type, atlas, period, bioregion, check_group, everything()),
  species_trait_checks |>
    mutate(
      check_type = "Species trait coverage",
      atlas = NA_character_,
      period = NA_character_,
      bioregion = NA,
      check_group = trait_family,
      percent_sum = NA_real_,
      max_abs_deviation_from_100 = NA_real_,
      expected_total_richness_known_trait = NA_real_,
      n_trait_groups = NA_integer_
    ) |>
    select(check_type, atlas, period, bioregion, check_group, everything())
  )
```

## Diagnostics

```{r diagnostics, results='asis'}
kable(
  thermal_group_summary,
  caption = "Thermal-affinity group sizes from STI quintiles."
)

kable(
  prediction_assignment_checks,
  caption = "Checks that model prediction rows join to bioregion cluster assignments."
)

kable(
  species_trait_checks,
  caption = "Modelled species with available trait metadata."
)

kable(
  composition_sum_checks |>
    mutate(percent_sum = round(percent_sum, 4),
           max_abs_deviation_from_100 = round(max_abs_deviation_from_100, 6)),
  caption = "Checks that trait percentages sum to 100 within each 1970s cluster-trait family."
)
```

## Dominant Trait Groups

```{r dominant-groups, results='asis'}
kable(
  cluster_trait_dominant_groups |>
    mutate(percent = percent(percent / 100, accuracy = 0.1)) |>
    select(bioregion, trait_family, dominant_trait_group, percent, n_species, n_sites),
  caption = "Dominant model-weighted trait group within each 1970s cluster."
)
```

## Thermal And Migratory Composition

```{r thermal-migration-plot, fig.width=8.5, fig.height=7}
plot_stacked_trait <- function(df, trait_family_label, palette) {
  plot_df <- df |>
    filter(trait_family == trait_family_label) |>
    mutate(
      trait_value = if (trait_family_label == "Thermal affinity") {
        factor(trait_value, levels = thermal_levels)
      } else {
        factor(trait_value, levels = names(palette))
      }
    )

  ggplot(plot_df, aes(x = bioregion, y = proportion, fill = trait_value)) +
    geom_col(width = 0.72, colour = "white", linewidth = 0.25) +
    scale_y_continuous(labels = percent_format(accuracy = 1), expand = expansion(mult = c(0, 0.02))) +
    scale_fill_manual(values = palette, drop = FALSE, name = NULL) +
    labs(
      title = trait_family_label,
      x = NULL,
      y = "Expected community composition"
    ) +
    theme_minimal(base_size = 11) +
    theme(
      panel.grid.major.x = element_blank(),
      panel.grid.minor = element_blank(),
      axis.text.x = element_text(angle = 35, hjust = 1),
      legend.position = "bottom",
      plot.title = element_text(face = "bold")
    )
}

p_thermal <- plot_stacked_trait(cluster_trait_composition, "Thermal affinity", thermal_palette)
p_migration <- plot_stacked_trait(cluster_trait_composition, "Migratory strategy", migration_palette)

thermal_migration_composition_plot <- p_thermal / p_migration +
  plot_annotation(
    title = "Model-weighted thermal and migratory composition of 1970s avian clusters",
    subtitle = "Percentages are expected richness shares within each Atlas 1 reference cluster."
  )

thermal_migration_composition_plot
```

## Foraging Guild Composition

```{r foraging-guild-heatmap, fig.width=8.5, fig.height=9.5}
foraging_order <- cluster_trait_composition |>
  filter(trait_family == "Foraging guild") |>
  group_by(trait_value) |>
  summarise(mean_percent = mean(percent, na.rm = TRUE), .groups = "drop") |>
  arrange(mean_percent) |>
  pull(trait_value)

foraging_heatmap_df <- cluster_trait_composition |>
  filter(trait_family == "Foraging guild") |>
  mutate(
    trait_value = factor(trait_value, levels = foraging_order),
    bioregion = factor(bioregion, levels = levels(bioregion_assignments$bioregion))
  )

foraging_guild_composition_heatmap <- ggplot(
  foraging_heatmap_df,
  aes(x = bioregion, y = trait_value, fill = proportion)
) +
  geom_tile(colour = "white", linewidth = 0.25) +
  scale_fill_distiller(
    palette = "YlGnBu",
    direction = 1,
    labels = percent_format(accuracy = 1),
    name = "Expected\nshare"
  ) +
  labs(
    title = "Model-weighted foraging-guild composition of 1970s avian clusters",
    subtitle = "Cells show expected richness share within each Atlas 1 reference cluster.",
    x = NULL,
    y = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(face = "bold")
  )

foraging_guild_composition_heatmap
```

## Atlas 1 Environmental Profiles

This section compares the environmental context of the three Atlas 1 reference
clusters. Values are centred on the national Atlas 1 median for each predictor
and scaled by the national Atlas 1 standard deviation, so positive values mean
that a cluster is above the national median for that variable and negative
values mean that it is below the national median.

```{r atlas1-environment-profiles}
reference_environment_predictors <- predictor_names(mods[[focal_atlas]]$XData)

reference_environment_values <- mods[[focal_atlas]]$XData |>
  as_tibble(rownames = "survey") |>
  select(survey, all_of(reference_environment_predictors)) |>
  inner_join(
    bioregion_assignments |> select(survey, site, bioregion),
    by = "survey"
  ) |>
  pivot_longer(
    cols = all_of(reference_environment_predictors),
    names_to = "variable",
    values_to = "value_raw"
  ) |>
  mutate(
    variable_label = recode(variable, !!!var_labels),
    variable_label = factor(variable_label, levels = names(environment_colours))
  )

reference_environment_national <- reference_environment_values |>
  group_by(variable, variable_label) |>
  summarise(
    national_mean = mean(value_raw, na.rm = TRUE),
    national_median = median(value_raw, na.rm = TRUE),
    national_sd = sd(value_raw, na.rm = TRUE),
    .groups = "drop"
  )

reference_environment_values <- reference_environment_values |>
  left_join(reference_environment_national, by = c("variable", "variable_label")) |>
  mutate(
    value_relative_to_national = if_else(
      national_sd > 0,
      (value_raw - national_median) / national_sd,
      NA_real_
    )
  )

reference_environment_plot_data <- bind_rows(
  reference_environment_values |>
    mutate(
      group = "All Atlas 1 cells",
      group_type = "National distribution"
    ),
  reference_environment_values |>
    mutate(
      group = as.character(bioregion),
      group_type = "Reference cluster"
    )
) |>
  mutate(
    group = factor(group, levels = c("All Atlas 1 cells", names(bioregion_palette))),
    group_type = factor(group_type, levels = c("National distribution", "Reference cluster"))
  )

reference_environment_summary <- reference_environment_plot_data |>
  group_by(group, group_type, bioregion, variable, variable_label) |>
  summarise(
    n_sites = n_distinct(site),
    mean_raw = mean(value_raw, na.rm = TRUE),
    median_raw = median(value_raw, na.rm = TRUE),
    q025_raw = unname(quantile(value_raw, 0.025, na.rm = TRUE)),
    q975_raw = unname(quantile(value_raw, 0.975, na.rm = TRUE)),
    mean_relative = mean(value_relative_to_national, na.rm = TRUE),
    median_relative = median(value_relative_to_national, na.rm = TRUE),
    q25_relative = unname(quantile(value_relative_to_national, 0.25, na.rm = TRUE)),
    q75_relative = unname(quantile(value_relative_to_national, 0.75, na.rm = TRUE)),
    q025_relative = unname(quantile(value_relative_to_national, 0.025, na.rm = TRUE)),
    q975_relative = unname(quantile(value_relative_to_national, 0.975, na.rm = TRUE)),
    .groups = "drop"
  )

reference_safe_wilcox_two_sample <- function(x, y) {
  x <- x[is.finite(x)]
  y <- y[is.finite(y)]
  if (length(x) < 2 || length(y) < 2 || length(unique(c(x, y))) < 2) {
    return(NA_real_)
  }
  suppressWarnings(wilcox.test(x, y, exact = FALSE)$p.value)
}

reference_safe_cliffs_delta <- function(x, y) {
  x <- x[is.finite(x)]
  y <- y[is.finite(y)]
  if (length(x) < 1 || length(y) < 1) {
    return(NA_real_)
  }
  ranks <- rank(c(x, y), ties.method = "average")
  n_x <- length(x)
  n_y <- length(y)
  rank_sum_x <- sum(ranks[seq_len(n_x)])
  u_statistic <- rank_sum_x - n_x * (n_x + 1) / 2
  2 * u_statistic / (n_x * n_y) - 1
}

reference_safe_bootstrap_median_difference <- function(x, y, n_boot = 2000) {
  x <- x[is.finite(x)]
  y <- y[is.finite(y)]
  if (length(x) < 2 || length(y) < 2) {
    return(tibble(
      median_difference_ci_low = NA_real_,
      median_difference_ci_high = NA_real_
    ))
  }
  bootstrap_differences <- replicate(
    n_boot,
    median(sample(x, length(x), replace = TRUE), na.rm = TRUE) -
      median(sample(y, length(y), replace = TRUE), na.rm = TRUE)
  )
  tibble(
    median_difference_ci_low = unname(quantile(bootstrap_differences, 0.025, na.rm = TRUE)),
    median_difference_ci_high = unname(quantile(bootstrap_differences, 0.975, na.rm = TRUE))
  )
}

set.seed(20260505)
reference_environment_tests <- reference_environment_values |>
  group_by(bioregion, variable, variable_label) |>
  summarise(
    n_cluster = n_distinct(site),
    n_other_cells = n_distinct(reference_environment_values$site[
      reference_environment_values$variable == first(variable) &
        reference_environment_values$bioregion != first(bioregion)
    ]),
    median_cluster = median(value_relative_to_national, na.rm = TRUE),
    median_other_cells = median(
      reference_environment_values$value_relative_to_national[
        reference_environment_values$variable == first(variable) &
          reference_environment_values$bioregion != first(bioregion)
      ],
      na.rm = TRUE
    ),
    median_difference_cluster_minus_other = median_cluster - median_other_cells,
    p_cluster_vs_other = reference_safe_wilcox_two_sample(
      value_relative_to_national,
      reference_environment_values$value_relative_to_national[
        reference_environment_values$variable == first(variable) &
          reference_environment_values$bioregion != first(bioregion)
      ]
    ),
    cliffs_delta_cluster_vs_other = reference_safe_cliffs_delta(
      value_relative_to_national,
      reference_environment_values$value_relative_to_national[
        reference_environment_values$variable == first(variable) &
          reference_environment_values$bioregion != first(bioregion)
      ]
    ),
    bootstrap = list(reference_safe_bootstrap_median_difference(
      value_relative_to_national,
      reference_environment_values$value_relative_to_national[
        reference_environment_values$variable == first(variable) &
          reference_environment_values$bioregion != first(bioregion)
      ]
    )),
    .groups = "drop"
  ) |>
  unnest(bootstrap) |>
  mutate(
    p_cluster_vs_other_adj = p.adjust(p_cluster_vs_other, method = "BH"),
    bootstrap_ci_excludes_zero = replace_na(
      median_difference_ci_low > 0 | median_difference_ci_high < 0,
      FALSE
    ),
    non_trivial_effect_size = replace_na(abs(cliffs_delta_cluster_vs_other) >= 0.147, FALSE),
    highlighted = replace_na(p_cluster_vs_other_adj < 0.05, FALSE) &
      bootstrap_ci_excludes_zero &
      non_trivial_effect_size,
    significance_status = if_else(
      highlighted,
      "Different from other Atlas 1 cells",
      "Not different from other Atlas 1 cells"
    )
  )

reference_environment_summary <- reference_environment_summary |>
  left_join(
    reference_environment_tests |>
      select(
        bioregion, variable, n_cluster, n_other_cells,
        median_difference_cluster_minus_other,
        median_difference_ci_low, median_difference_ci_high,
        p_cluster_vs_other, p_cluster_vs_other_adj,
        cliffs_delta_cluster_vs_other,
        bootstrap_ci_excludes_zero, non_trivial_effect_size,
        highlighted, significance_status
      ),
    by = c("bioregion", "variable")
  ) |>
  mutate(
    highlighted = if_else(group_type == "Reference cluster", replace_na(highlighted, FALSE), FALSE),
    significance_status = case_when(
      group_type == "Reference cluster" & highlighted ~ "Different from other Atlas 1 cells",
      group_type == "Reference cluster" ~ "Not different from other Atlas 1 cells",
      TRUE ~ "National distribution"
    )
  )
```

```{r atlas1-environment-profile-plot, fig.width=10.5, fig.height=6.3}
reference_environment_national_summary <- reference_environment_summary |>
  filter(group == "All Atlas 1 cells")

reference_environment_cluster_summary <- reference_environment_summary |>
  filter(group != "All Atlas 1 cells") |>
  mutate(
    significance_status = factor(
      significance_status,
      levels = c(
        "Different from other Atlas 1 cells",
        "Not different from other Atlas 1 cells"
      )
    )
  )

reference_environment_profile_plot <- ggplot() +
  geom_linerange(
    data = reference_environment_national_summary,
    aes(x = variable_label, ymin = q025_relative, ymax = q975_relative),
    colour = "grey82",
    linewidth = 1.1
  ) +
  geom_linerange(
    data = reference_environment_national_summary,
    aes(x = variable_label, ymin = q25_relative, ymax = q75_relative),
    colour = "grey62",
    linewidth = 3.2
  ) +
  geom_point(
    data = reference_environment_national_summary,
    aes(x = variable_label, y = median_relative),
    colour = "grey25",
    size = 1.7
  ) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey35", linewidth = 0.35) +
  geom_pointrange(
    data = reference_environment_cluster_summary,
    aes(
      x = variable_label,
      y = median_relative,
      ymin = q25_relative,
      ymax = q75_relative,
      colour = group,
      alpha = significance_status
    ),
    position = position_dodge(width = 0.62),
    linewidth = 0.65,
    fatten = 2.2
  ) +
  coord_flip() +
  scale_colour_manual(values = bioregion_palette, drop = FALSE, name = "Atlas 1 cluster") +
  scale_alpha_manual(
    values = c(
      "Different from other Atlas 1 cells" = 1,
      "Not different from other Atlas 1 cells" = 0.24
    ),
    guide = "none"
  ) +
  labs(
    title = "Environmental profiles of Atlas 1 reference clusters",
    subtitle = str_wrap(
      "Grey intervals show all Atlas 1 Danish cells centred on the national median. Coloured points show cluster medians and interquartile ranges; full opacity marks clusters that differ from all other Atlas 1 cells after BH-adjusted Wilcoxon testing, bootstrap support, and a non-trivial Cliff's delta.",
      width = 100
    ),
    x = NULL,
    y = "Environmental value relative to national Atlas 1 median (SD units)"
  ) +
  theme_minimal(base_size = 10) +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "top",
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(size = 10, lineheight = 1.05)
  )

reference_environment_profile_plot
```

```{r atlas1-environment-profile-table, results='asis'}
kable(
  reference_environment_tests |>
    filter(highlighted) |>
    mutate(
      median_difference_cluster_minus_other = round(median_difference_cluster_minus_other, 3),
      median_difference_ci_low = round(median_difference_ci_low, 3),
      median_difference_ci_high = round(median_difference_ci_high, 3),
      p_cluster_vs_other_adj = signif(p_cluster_vs_other_adj, 3),
      cliffs_delta_cluster_vs_other = round(cliffs_delta_cluster_vs_other, 3)
    ) |>
    select(
      bioregion, variable_label, n_cluster, n_other_cells,
      median_difference_cluster_minus_other,
      median_difference_ci_low, median_difference_ci_high,
      p_cluster_vs_other_adj, cliffs_delta_cluster_vs_other
    ),
  caption = "Atlas 1 environmental predictors where a cluster differs from all other Atlas 1 cells."
)
```

## Drivers Of Shifts Into Cluster 2

This section asks what changed in the sites that were part of `Cluster 1` or
`Cluster 3` in the 1970s, but were assigned to `Cluster 2` in a later atlas
period. The analysis is paired by grid cell: for each focal site and species,
the 1970s predicted probability is subtracted from the later predicted
probability. Positive values indicate modelled increases, while negative values
indicate modelled losses.

Trait-driver plots are shown as a size-normalized version. For each trait group,
the summed expected-richness delta is divided by the number of modelled species
in that group. This asks whether the average species in a trait group increased
or decreased, rather than giving larger trait groups more weight simply because
they contain more species.

```{r cluster2-shift-drivers}
reference_species <- colnames(predsY_all[[focal_atlas]])
mismatched_species <- names(predsY_all)[map_lgl(predsY_all, ~ !identical(colnames(.x), reference_species))]
if (length(mismatched_species) > 0) {
  stop("Species columns do not align for atlas: ", paste(mismatched_species, collapse = ", "), call. = FALSE)
}

assignments_wide <- bioregion_assignments_all |>
  select(site, atlas, survey, bioregion, assignment_margin, low_confidence) |>
  pivot_wider(
    names_from = atlas,
    values_from = c(survey, bioregion, assignment_margin, low_confidence),
    names_sep = "_"
  )

make_focal_transition <- function(source_cluster, target_atlas) {
  target_period <- year_lookup[[target_atlas]]

  assignments_wide |>
    filter(
      bioregion_1 == source_cluster,
      .data[[paste0("bioregion_", target_atlas)]] == "Cluster 2"
    ) |>
    transmute(
      comparison = paste0(target_period, " minus 1970s"),
      source_cluster = source_cluster,
      target_cluster = "Cluster 2",
      transition = paste(source_cluster, "to Cluster 2"),
      target_atlas = target_atlas,
      target_period = target_period,
      site,
      survey_from = survey_1,
      survey_to = .data[[paste0("survey_", target_atlas)]],
      assignment_margin_target = .data[[paste0("assignment_margin_", target_atlas)]],
      low_confidence_target = .data[[paste0("low_confidence_", target_atlas)]]
    )
}

cluster2_shift_focal_sites <- bind_rows(
  make_focal_transition("Cluster 1", "2"),
  make_focal_transition("Cluster 3", "2"),
  make_focal_transition("Cluster 1", "3"),
  make_focal_transition("Cluster 3", "3")
) |>
  mutate(
    transition = factor(transition, levels = c("Cluster 1 to Cluster 2", "Cluster 3 to Cluster 2")),
    comparison = factor(comparison, levels = c("1990s minus 1970s", "2010s minus 1970s"))
  )

cluster2_shift_site_counts <- cluster2_shift_focal_sites |>
  group_by(comparison, transition) |>
  summarise(
    n_sites = n(),
    n_low_confidence = sum(low_confidence_target, na.rm = TRUE),
    pct_low_confidence = mean(low_confidence_target, na.rm = TRUE),
    median_assignment_margin = median(assignment_margin_target, na.rm = TRUE),
    .groups = "drop"
  )

environment_predictors <- predictor_names(mods[[focal_atlas]]$XData)
environment_predictors <- environment_predictors[environment_predictors %in% colnames(mods[["3"]]$XData)]

environment_frame <- function(model, atlas) {
  model$XData |>
    as_tibble(rownames = "survey") |>
    select(survey, all_of(environment_predictors)) |>
    mutate(atlas = atlas, .before = 1)
}

environment_from <- environment_frame(mods[[focal_atlas]], focal_atlas) |>
  rename_with(~ paste0(.x, "_from"), all_of(environment_predictors))

environment_to <- environment_frame(mods[["3"]], "3") |>
  rename_with(~ paste0(.x, "_to"), all_of(environment_predictors))

environment_matched_sites <- assignments_wide |>
  filter(!is.na(survey_1), !is.na(survey_3)) |>
  select(site, survey_from = survey_1, survey_to = survey_3, bioregion_1, bioregion_3) |>
  left_join(environment_from, by = c("survey_from" = "survey")) |>
  left_join(environment_to, by = c("survey_to" = "survey")) |>
  pivot_longer(
    cols = matches(paste0("^(", paste(environment_predictors, collapse = "|"), ")_(from|to)$")),
    names_to = c("variable", "timepoint"),
    names_pattern = "(.+)_(from|to)",
    values_to = "value"
  ) |>
  pivot_wider(names_from = timepoint, values_from = value) |>
  mutate(
    delta_raw = to - from,
    variable_label = recode(variable, !!!var_labels)
  )

environment_delta_reference <- environment_matched_sites |>
  group_by(variable, variable_label) |>
  summarise(
    national_mean_delta = mean(delta_raw, na.rm = TRUE),
    national_median_delta = median(delta_raw, na.rm = TRUE),
    national_sd_delta = sd(delta_raw, na.rm = TRUE),
    .groups = "drop"
  )

environment_matched_sites <- environment_matched_sites |>
  left_join(environment_delta_reference, by = c("variable", "variable_label")) |>
  mutate(
    delta_relative_to_national = if_else(
      national_sd_delta > 0,
      (delta_raw - national_median_delta) / national_sd_delta,
      NA_real_
    )
  )

environment_retained_comparison_groups <- assignments_wide |>
  filter(
    (bioregion_1 == "Cluster 1" & bioregion_3 %in% c("Cluster 1", "Cluster 2")) |
      (bioregion_1 == "Cluster 3" & bioregion_3 %in% c("Cluster 3", "Cluster 2"))
  ) |>
  transmute(
    site,
    source_cluster = bioregion_1,
    target_cluster = bioregion_3,
    group_status = if_else(
      target_cluster == "Cluster 2",
      "Shifted into Cluster 2",
      "Retained source cluster"
    ),
    group = case_when(
      source_cluster == "Cluster 1" & target_cluster == "Cluster 1" ~ "Cluster 1 to Cluster 1",
      source_cluster == "Cluster 1" & target_cluster == "Cluster 2" ~ "Cluster 1 to Cluster 2",
      source_cluster == "Cluster 3" & target_cluster == "Cluster 3" ~ "Cluster 3 to Cluster 3",
      source_cluster == "Cluster 3" & target_cluster == "Cluster 2" ~ "Cluster 3 to Cluster 2"
    )
  ) |>
  distinct()

environment_retained_comparison_counts <- environment_retained_comparison_groups |>
  count(source_cluster, group_status, group, name = "n_sites")

environment_delta_plot_data <- bind_rows(
  environment_matched_sites |>
    mutate(
      source_cluster = NA_character_,
      target_cluster = NA_character_,
      group_status = "National distribution",
      group = "All matched cells"
    ),
  environment_matched_sites |>
    inner_join(environment_retained_comparison_groups, by = "site")
) |>
  mutate(
    group = factor(
      group,
      levels = c(
        "All matched cells",
        "Cluster 1 to Cluster 1", "Cluster 1 to Cluster 2",
        "Cluster 3 to Cluster 3", "Cluster 3 to Cluster 2"
      )
    ),
    group_status = factor(
      group_status,
      levels = c("National distribution", "Retained source cluster", "Shifted into Cluster 2")
    ),
    source_cluster = factor(source_cluster, levels = c("Cluster 1", "Cluster 3")),
    variable_label = factor(variable_label, levels = names(environment_colours))
  )

environment_delta_summary <- environment_delta_plot_data |>
  group_by(source_cluster, target_cluster, group_status, group, variable, variable_label) |>
  summarise(
    n_sites = n_distinct(site),
    mean_delta_raw = mean(delta_raw, na.rm = TRUE),
    median_delta_raw = median(delta_raw, na.rm = TRUE),
    q025_delta_raw = quantile(delta_raw, 0.025, na.rm = TRUE),
    q975_delta_raw = quantile(delta_raw, 0.975, na.rm = TRUE),
    mean_delta_relative = mean(delta_relative_to_national, na.rm = TRUE),
    median_delta_relative = median(delta_relative_to_national, na.rm = TRUE),
    q25_delta_relative = quantile(delta_relative_to_national, 0.25, na.rm = TRUE),
    q75_delta_relative = quantile(delta_relative_to_national, 0.75, na.rm = TRUE),
    q025_delta_relative = quantile(delta_relative_to_national, 0.025, na.rm = TRUE),
    q975_delta_relative = quantile(delta_relative_to_national, 0.975, na.rm = TRUE),
    .groups = "drop"
  )

safe_wilcox_one_sample <- function(x, mu = 0) {
  x <- x[is.finite(x)]
  if (length(x) < 2 || length(unique(x)) < 2) {
    return(NA_real_)
  }
  suppressWarnings(wilcox.test(x, mu = mu, exact = FALSE)$p.value)
}

safe_wilcox_two_sample <- function(x, y) {
  x <- x[is.finite(x)]
  y <- y[is.finite(y)]
  if (length(x) < 2 || length(y) < 2 || length(unique(c(x, y))) < 2) {
    return(NA_real_)
  }
  suppressWarnings(wilcox.test(x, y, exact = FALSE)$p.value)
}

safe_cliffs_delta <- function(x, y) {
  x <- x[is.finite(x)]
  y <- y[is.finite(y)]
  if (length(x) < 1 || length(y) < 1) {
    return(NA_real_)
  }
  ranks <- rank(c(x, y), ties.method = "average")
  n_x <- length(x)
  n_y <- length(y)
  rank_sum_x <- sum(ranks[seq_len(n_x)])
  u_statistic <- rank_sum_x - n_x * (n_x + 1) / 2
  2 * u_statistic / (n_x * n_y) - 1
}

safe_bootstrap_median_difference <- function(x, y, n_boot = 2000) {
  x <- x[is.finite(x)]
  y <- y[is.finite(y)]
  if (length(x) < 2 || length(y) < 2) {
    return(tibble(
      median_difference_ci_low = NA_real_,
      median_difference_ci_high = NA_real_
    ))
  }
  bootstrap_differences <- replicate(
    n_boot,
    median(sample(x, length(x), replace = TRUE), na.rm = TRUE) -
      median(sample(y, length(y), replace = TRUE), na.rm = TRUE)
  )
  tibble(
    median_difference_ci_low = unname(quantile(bootstrap_differences, 0.025, na.rm = TRUE)),
    median_difference_ci_high = unname(quantile(bootstrap_differences, 0.975, na.rm = TRUE))
  )
}

set.seed(20260505)
environment_delta_tests <- environment_delta_plot_data |>
  filter(group_status %in% c("Retained source cluster", "Shifted into Cluster 2")) |>
  group_by(source_cluster, variable, variable_label) |>
  summarise(
    n_changed = n_distinct(site[group_status == "Shifted into Cluster 2"]),
    n_retained = n_distinct(site[group_status == "Retained source cluster"]),
    median_changed = median(delta_relative_to_national[group_status == "Shifted into Cluster 2"], na.rm = TRUE),
    median_retained = median(delta_relative_to_national[group_status == "Retained source cluster"], na.rm = TRUE),
    median_difference_changed_minus_retained = median_changed - median_retained,
    p_changed_vs_retained = safe_wilcox_two_sample(
      delta_relative_to_national[group_status == "Shifted into Cluster 2"],
      delta_relative_to_national[group_status == "Retained source cluster"]
    ),
    cliffs_delta_changed_vs_retained = safe_cliffs_delta(
      delta_relative_to_national[group_status == "Shifted into Cluster 2"],
      delta_relative_to_national[group_status == "Retained source cluster"]
    ),
    bootstrap = list(safe_bootstrap_median_difference(
      delta_relative_to_national[group_status == "Shifted into Cluster 2"],
      delta_relative_to_national[group_status == "Retained source cluster"]
    )),
    .groups = "drop"
  ) |>
  unnest(bootstrap) |>
  mutate(
    p_changed_vs_retained_adj = p.adjust(p_changed_vs_retained, method = "BH"),
    sample_size_ratio = pmax(n_changed, n_retained) / pmax(pmin(n_changed, n_retained), 1),
    bootstrap_ci_excludes_zero = replace_na(
      median_difference_ci_low > 0 | median_difference_ci_high < 0,
      FALSE
    ),
    non_trivial_effect_size = replace_na(abs(cliffs_delta_changed_vs_retained) >= 0.147, FALSE),
    significant_changed_vs_retained = replace_na(p_changed_vs_retained_adj < 0.05, FALSE) &
      bootstrap_ci_excludes_zero,
    highlighted = significant_changed_vs_retained & non_trivial_effect_size,
    significance_status = if_else(
      highlighted,
      "Different from retained cells",
      "Not different from retained cells"
    )
  )

environment_delta_summary <- environment_delta_summary |>
  left_join(
    environment_delta_tests |>
      select(
        source_cluster, variable, n_changed, n_retained,
        median_difference_changed_minus_retained,
        median_difference_ci_low, median_difference_ci_high,
        p_changed_vs_retained, p_changed_vs_retained_adj,
        cliffs_delta_changed_vs_retained, sample_size_ratio,
        bootstrap_ci_excludes_zero, non_trivial_effect_size,
        significant_changed_vs_retained, highlighted, significance_status
      ),
    by = c("source_cluster", "variable")
  ) |>
  mutate(
    highlighted = if_else(group_status == "Shifted into Cluster 2", replace_na(highlighted, FALSE), FALSE),
    significance_status = case_when(
      group_status == "Shifted into Cluster 2" & highlighted ~ "Different from retained cells",
      group_status == "Shifted into Cluster 2" ~ "Not different from retained cells",
      group_status == "Retained source cluster" ~ "Retained source-cluster cells",
      TRUE ~ "National distribution"
    )
  )

make_species_delta_frame <- function(focal_df) {
  target_atlas <- unique(focal_df$target_atlas)
  if (length(target_atlas) != 1) {
    stop("Expected one target atlas per focal transition group.", call. = FALSE)
  }

  missing_from <- setdiff(focal_df$survey_from, rownames(predsY_all[[focal_atlas]]))
  missing_to <- setdiff(focal_df$survey_to, rownames(predsY_all[[target_atlas]]))
  if (length(missing_from) > 0 || length(missing_to) > 0) {
    stop("Missing prediction rows for focal transition sites.", call. = FALSE)
  }

  from_pred <- predsY_all[[focal_atlas]][focal_df$survey_from, reference_species, drop = FALSE]
  to_pred <- predsY_all[[target_atlas]][focal_df$survey_to, reference_species, drop = FALSE]

  from_long <- as_tibble(from_pred, .name_repair = "minimal") |>
    mutate(site = focal_df$site, .before = 1) |>
    pivot_longer(-site, names_to = "species", values_to = "probability_from")

  to_long <- as_tibble(to_pred, .name_repair = "minimal") |>
    mutate(site = focal_df$site, .before = 1) |>
    pivot_longer(-site, names_to = "species", values_to = "probability_to")

  focal_df |>
    select(
      comparison, transition, source_cluster, target_cluster, target_atlas,
      target_period, site, low_confidence_target
    ) |>
    left_join(from_long, by = "site", relationship = "one-to-many") |>
    left_join(to_long, by = c("site", "species")) |>
    mutate(
      probability_delta = probability_to - probability_from,
      probability_gain = pmax(probability_delta, 0),
      probability_loss = pmin(probability_delta, 0)
    )
}

cluster2_shift_species_site_deltas <- cluster2_shift_focal_sites |>
  group_split(comparison, transition) |>
  map_dfr(make_species_delta_frame)

species_trait_table <- traits |>
  select(
    species,
    thermal_group,
    migratory_strategy = Migration_a3_DOF,
    foraging_guild = foraging_guild_consensus,
    species_thermal_index
  )

cluster2_shift_species_deltas <- cluster2_shift_species_site_deltas |>
  group_by(comparison, transition, source_cluster, target_cluster, species) |>
  summarise(
    n_sites = n(),
    mean_probability_from = mean(probability_from, na.rm = TRUE),
    mean_probability_to = mean(probability_to, na.rm = TRUE),
    mean_probability_delta = mean(probability_delta, na.rm = TRUE),
    median_probability_delta = median(probability_delta, na.rm = TRUE),
    q10_probability_delta = quantile(probability_delta, 0.10, na.rm = TRUE),
    q90_probability_delta = quantile(probability_delta, 0.90, na.rm = TRUE),
    mean_probability_gain = mean(probability_gain, na.rm = TRUE),
    mean_probability_loss = mean(probability_loss, na.rm = TRUE),
    prop_sites_increase = mean(probability_delta > 0, na.rm = TRUE),
    .groups = "drop"
  ) |>
  left_join(species_trait_table, by = "species") |>
  mutate(
    driver_direction = if_else(mean_probability_delta >= 0, "Gain", "Loss"),
    species_label = str_replace_all(species, "_", " ")
  )

cluster2_shift_top_species_drivers <- bind_rows(
  cluster2_shift_species_deltas |>
    filter(mean_probability_delta < 0) |>
    group_by(comparison, transition) |>
    slice_min(mean_probability_delta, n = 12, with_ties = FALSE) |>
    ungroup(),
  cluster2_shift_species_deltas |>
    filter(mean_probability_delta > 0) |>
    group_by(comparison, transition) |>
    slice_max(mean_probability_delta, n = 12, with_ties = FALSE) |>
    ungroup()
) |>
  arrange(comparison, transition, mean_probability_delta)

cluster2_shift_site_trait_deltas <- cluster2_shift_species_site_deltas |>
  inner_join(trait_lookup, by = "species", relationship = "many-to-many") |>
  group_by(comparison, transition, source_cluster, target_cluster, site, trait_family, trait_value) |>
  summarise(
    expected_richness_delta = sum(probability_delta, na.rm = TRUE),
    expected_richness_gain = sum(probability_gain, na.rm = TRUE),
    expected_richness_loss = sum(probability_loss, na.rm = TRUE),
    .groups = "drop"
  )

cluster2_shift_trait_deltas <- cluster2_shift_site_trait_deltas |>
  group_by(comparison, transition, source_cluster, target_cluster, trait_family, trait_value) |>
  summarise(
    n_sites = n_distinct(site),
    mean_expected_richness_delta = mean(expected_richness_delta, na.rm = TRUE),
    median_expected_richness_delta = median(expected_richness_delta, na.rm = TRUE),
    q10_expected_richness_delta = quantile(expected_richness_delta, 0.10, na.rm = TRUE),
    q90_expected_richness_delta = quantile(expected_richness_delta, 0.90, na.rm = TRUE),
    mean_expected_richness_gain = mean(expected_richness_gain, na.rm = TRUE),
    mean_expected_richness_loss = mean(expected_richness_loss, na.rm = TRUE),
    prop_sites_increase = mean(expected_richness_delta > 0, na.rm = TRUE),
    .groups = "drop"
  ) |>
  left_join(trait_group_sizes, by = c("trait_family", "trait_value")) |>
  mutate(
    mean_expected_richness_delta_per_species = mean_expected_richness_delta / n_species_in_trait_group,
    median_expected_richness_delta_per_species = median_expected_richness_delta / n_species_in_trait_group,
    q10_expected_richness_delta_per_species = q10_expected_richness_delta / n_species_in_trait_group,
    q90_expected_richness_delta_per_species = q90_expected_richness_delta / n_species_in_trait_group,
    mean_expected_richness_gain_per_species = mean_expected_richness_gain / n_species_in_trait_group,
    mean_expected_richness_loss_per_species = mean_expected_richness_loss / n_species_in_trait_group
  ) |>
  group_by(comparison, transition, trait_family) |>
  mutate(
    total_absolute_trait_change = sum(abs(mean_expected_richness_delta), na.rm = TRUE),
    total_absolute_trait_change_per_species = sum(abs(mean_expected_richness_delta_per_species), na.rm = TRUE),
    absolute_change_share = ifelse(
      total_absolute_trait_change > 0,
      abs(mean_expected_richness_delta) / total_absolute_trait_change,
      NA_real_
    ),
    absolute_change_share_per_species = ifelse(
      total_absolute_trait_change_per_species > 0,
      abs(mean_expected_richness_delta_per_species) / total_absolute_trait_change_per_species,
      NA_real_
    ),
    driver_direction = if_else(mean_expected_richness_delta >= 0, "Gain", "Loss"),
    driver_direction_per_species = if_else(mean_expected_richness_delta_per_species >= 0, "Gain", "Loss")
  ) |>
  ungroup()

cluster2_shift_trait_drivers <- bind_rows(
  cluster2_shift_trait_deltas |>
    filter(mean_expected_richness_delta < 0) |>
    group_by(comparison, transition, trait_family) |>
    slice_min(mean_expected_richness_delta, n = 5, with_ties = FALSE) |>
    ungroup(),
  cluster2_shift_trait_deltas |>
    filter(mean_expected_richness_delta > 0) |>
    group_by(comparison, transition, trait_family) |>
    slice_max(mean_expected_richness_delta, n = 5, with_ties = FALSE) |>
    ungroup()
)

cluster2_shift_trait_drivers_size_normalized <- bind_rows(
  cluster2_shift_trait_deltas |>
    filter(mean_expected_richness_delta_per_species < 0) |>
    group_by(comparison, transition, trait_family) |>
    slice_min(mean_expected_richness_delta_per_species, n = 5, with_ties = FALSE) |>
    ungroup(),
  cluster2_shift_trait_deltas |>
    filter(mean_expected_richness_delta_per_species > 0) |>
    group_by(comparison, transition, trait_family) |>
    slice_max(mean_expected_richness_delta_per_species, n = 5, with_ties = FALSE) |>
    ungroup()
)

cluster2_shift_checks <- bind_rows(
  cluster2_shift_site_counts |>
    mutate(
      check_type = "Focal transition counts",
      expected_n_sites = case_when(
        comparison == "1990s minus 1970s" & transition == "Cluster 1 to Cluster 2" ~ 142L,
        comparison == "1990s minus 1970s" & transition == "Cluster 3 to Cluster 2" ~ 219L,
        comparison == "2010s minus 1970s" & transition == "Cluster 1 to Cluster 2" ~ 245L,
        comparison == "2010s minus 1970s" & transition == "Cluster 3 to Cluster 2" ~ 392L,
        TRUE ~ NA_integer_
      ),
      passed = n_sites == expected_n_sites
    ) |>
    select(check_type, comparison, transition, n_sites, expected_n_sites, passed, everything()),
  cluster2_shift_focal_sites |>
    group_by(comparison, transition, target_atlas) |>
    summarise(
      check_type = "Prediction row matches",
      n_sites = n(),
      n_missing_from = sum(!survey_from %in% rownames(predsY_all[[focal_atlas]])),
      n_missing_to = sum(!survey_to %in% rownames(predsY_all[[unique(target_atlas)]])),
      passed = n_missing_from == 0 & n_missing_to == 0,
      .groups = "drop"
    ) |>
    mutate(expected_n_sites = NA_integer_) |>
    select(check_type, comparison, transition, n_sites, expected_n_sites, passed, everything()),
  tibble(
    check_type = "Species metadata coverage",
    comparison = NA,
    transition = NA,
    n_sites = NA_integer_,
    expected_n_sites = NA_integer_,
    passed = all(reference_species %in% species_trait_table$species),
    n_model_species = length(reference_species),
    n_species_missing_metadata = sum(!reference_species %in% species_trait_table$species)
  )
)
```

```{r cluster2-shift-tables, results='asis'}
kable(
  cluster2_shift_site_counts |>
    mutate(
      pct_low_confidence = percent(pct_low_confidence, accuracy = 0.1),
      median_assignment_margin = round(median_assignment_margin, 3)
    ),
  caption = "Focal sites that shifted from Cluster 1 or Cluster 3 into Cluster 2."
)

kable(
  cluster2_shift_trait_drivers_size_normalized |>
    filter(comparison == cluster2_shift_focal_comparison) |>
    mutate(
      n_species_in_trait_group = as.integer(n_species_in_trait_group),
      mean_expected_richness_delta = round(mean_expected_richness_delta, 3),
      mean_expected_richness_delta_per_species = round(mean_expected_richness_delta_per_species, 4),
      mean_expected_richness_loss_per_species = round(mean_expected_richness_loss_per_species, 4),
      mean_expected_richness_gain_per_species = round(mean_expected_richness_gain_per_species, 4),
      absolute_change_share_per_species = percent(absolute_change_share_per_species, accuracy = 0.1),
      prop_sites_increase = percent(prop_sites_increase, accuracy = 1)
    ) |>
    select(
      comparison, transition, trait_family, driver_direction_per_species, trait_value,
      n_species_in_trait_group, mean_expected_richness_delta,
      mean_expected_richness_delta_per_species,
      mean_expected_richness_loss_per_species,
      mean_expected_richness_gain_per_species,
      absolute_change_share_per_species, prop_sites_increase
    ),
  caption = "Strongest size-normalized trait-group losses and gains among sites shifting into Cluster 2 by the 2010s."
)
```

```{r cluster2-shift-environment-plot, fig.width=10.5, fig.height=6.5}
environment_national_summary <- environment_delta_summary |>
  filter(group == "All matched cells") |>
  mutate(
    national_direction = case_when(
      mean_delta_raw > 0 ~ "Increasing nationally",
      mean_delta_raw < 0 ~ "Decreasing nationally",
      TRUE ~ "No national mean change"
    ),
    national_direction = factor(
      national_direction,
      levels = c("Increasing nationally", "Decreasing nationally", "No national mean change")
    )
  )

environment_focal_summary <- environment_delta_summary |>
  filter(group != "All matched cells") |>
  mutate(
    significance_status = factor(
      significance_status,
      levels = c(
        "Different from retained cells",
        "Not different from retained cells",
        "Retained source-cluster cells"
      )
    )
  )

cluster2_shift_environment_plot <- ggplot() +
  geom_linerange(
    data = environment_national_summary,
    aes(
      x = variable_label,
      ymin = q025_delta_relative,
      ymax = q975_delta_relative,
      colour = national_direction
    ),
    linewidth = 1.1,
    alpha = 0.28
  ) +
  geom_linerange(
    data = environment_national_summary,
    aes(
      x = variable_label,
      ymin = q25_delta_relative,
      ymax = q75_delta_relative,
      colour = national_direction
    ),
    linewidth = 3.2,
    alpha = 0.68
  ) +
  geom_point(
    data = environment_national_summary,
    aes(x = variable_label, y = median_delta_relative),
    colour = "grey25",
    size = 1.7
  ) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey35", linewidth = 0.35) +
  geom_pointrange(
    data = environment_focal_summary,
    aes(
      x = variable_label,
      y = median_delta_relative,
      ymin = q25_delta_relative,
      ymax = q75_delta_relative,
      colour = group,
      alpha = significance_status
    ),
    position = position_dodge(width = 0.58),
    linewidth = 0.65,
    fatten = 2.2
  ) +
  coord_flip() +
  scale_colour_manual(
    values = c(
      "Increasing nationally" = "#3b88c3",
      "Decreasing nationally" = "#c65a32",
      "No national mean change" = "grey70",
      "Cluster 1 to Cluster 1" = "#8dd3c7",
      "Cluster 1 to Cluster 2" = "#1b9e77",
      "Cluster 3 to Cluster 3" = "#bebada",
      "Cluster 3 to Cluster 2" = "#7570b3"
    ),
    breaks = c(
      "Increasing nationally",
      "Decreasing nationally",
      "Cluster 1 to Cluster 1",
      "Cluster 1 to Cluster 2",
      "Cluster 3 to Cluster 3",
      "Cluster 3 to Cluster 2"
    ),
    name = NULL
  ) +
  scale_alpha_manual(
    values = c(
      "Different from retained cells" = 1,
      "Not different from retained cells" = 0.22,
      "Retained source-cluster cells" = 0.46
    ),
    guide = "none"
  ) +
  labs(
    title = "Environmental context of cells shifting into Cluster 2",
    subtitle = str_wrap(
      "National intervals are centred so each median Danish change equals zero; blue/orange bars show whether the mean national change increased or decreased. Shifted cells are tested against cells that started in the same cluster and remained there; full opacity requires BH-adjusted Wilcoxon support, a bootstrap median-difference interval excluding zero, and a non-trivial Cliff's delta.",
      width = 95
    ),
    x = NULL,
    y = "Environmental change anomaly from national median (SD units)"
  ) +
  theme_minimal(base_size = 10) +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "top",
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(size = 10, lineheight = 1.05)
  )

cluster2_shift_environment_plot
```

```{r cluster2-shift-thermal-migration-plot, fig.width=10.5, fig.height=6.5}
thermal_migration_driver_df <- cluster2_shift_trait_deltas |>
  filter(
    comparison == cluster2_shift_focal_comparison,
    trait_family %in% c("Thermal affinity", "Migratory strategy")
  ) |>
  mutate(
    trait_value = factor(trait_value, levels = c(thermal_levels, rev(migration_levels))),
    trait_colour_group = as.character(trait_value)
  )

cluster2_shift_thermal_migration_plot <- ggplot(
  thermal_migration_driver_df,
  aes(
    x = mean_expected_richness_delta_per_species,
    y = trait_value,
    fill = trait_colour_group
  )
) +
  geom_vline(xintercept = 0, colour = "grey35", linewidth = 0.35) +
  geom_col(width = 0.72, colour = "white", linewidth = 0.2) +
  facet_grid(trait_family ~ transition, scales = "free_y", space = "free_y") +
  scale_fill_manual(
    values = c(thermal_palette, migration_palette),
    breaks = c(rev(thermal_levels), migration_levels),
    name = NULL
  ) +
  labs(
    title = "Size-normalized thermal and migratory drivers of shifts into Cluster 2 by the 2010s",
    subtitle = "Bars show mean expected-richness delta per species in each trait group.",
    x = "Mean expected richness delta per species",
    y = NULL
  ) +
  theme_minimal(base_size = 10) +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    strip.text = element_text(face = "bold"),
    legend.position = "top"
  )

cluster2_shift_thermal_migration_plot
```

```{r cluster2-shift-foraging-plot, fig.width=12.5, fig.height=8.5}
aquatic_wetland_guilds <- c(
  "Aquatic pursuers", "Dabbling ducks", "Dippers", "Diving ducks",
  "Grazing waterfowl", "Gulls", "Kingfishers", "Marsh warblers",
  "Plovers", "Rails", "Reedlings", "Scolopacids", "Terns", "Wading birds"
)

foraging_driver_order <- cluster2_shift_trait_deltas |>
  filter(
    comparison == cluster2_shift_focal_comparison,
    trait_family == "Foraging guild"
  ) |>
  group_by(trait_value) |>
  summarise(mean_delta = mean(mean_expected_richness_delta_per_species, na.rm = TRUE), .groups = "drop") |>
  arrange(mean_delta) |>
  pull(trait_value)

foraging_driver_df <- cluster2_shift_trait_deltas |>
  filter(
    comparison == cluster2_shift_focal_comparison,
    trait_family == "Foraging guild"
  ) |>
  mutate(
    trait_value = factor(trait_value, levels = foraging_driver_order),
    transition = factor(transition, levels = c("Cluster 1 to Cluster 2", "Cluster 3 to Cluster 2"))
  )

guild_driver_annotations <- species_trait_table |>
  filter(!is.na(foraging_guild), foraging_guild != "") |>
  mutate(
    foraging_habitat = if_else(
      foraging_guild %in% aquatic_wetland_guilds,
      "Aquatic/wetland",
      "Terrestrial/aerial"
    ),
    is_long_distance = str_detect(replace_na(migratory_strategy, ""), "long-distance")
  ) |>
  group_by(foraging_guild) |>
  summarise(
    trait_value = first(foraging_guild),
    foraging_habitat = if_else(
      mean(foraging_habitat == "Aquatic/wetland", na.rm = TRUE) >= 0.5,
      "Aquatic/wetland",
      "Terrestrial/aerial"
    ),
    n_species = n(),
    .groups = "drop"
  ) |>
  filter(trait_value %in% foraging_driver_order) |>
  mutate(
    trait_value = factor(trait_value, levels = foraging_driver_order),
    foraging_habitat = factor(foraging_habitat, levels = c("Aquatic/wetland", "Terrestrial/aerial"))
  )

guild_migration_composition <- species_trait_table |>
  filter(
    !is.na(foraging_guild), foraging_guild != "",
    !is.na(migratory_strategy), migratory_strategy != "",
    foraging_guild %in% foraging_driver_order
  ) |>
  count(foraging_guild, migratory_strategy, name = "n_species") |>
  group_by(foraging_guild) |>
  mutate(proportion = n_species / sum(n_species, na.rm = TRUE)) |>
  ungroup() |>
  complete(
    foraging_guild = foraging_driver_order,
    migratory_strategy = migration_levels,
    fill = list(n_species = 0, proportion = 0)
  ) |>
  mutate(
    trait_value = factor(foraging_guild, levels = foraging_driver_order),
    migratory_strategy = factor(migratory_strategy, levels = migration_levels)
  )

overall_mean_sti <- mean(species_trait_table$species_thermal_index, na.rm = TRUE)

guild_sti_summary <- species_trait_table |>
  filter(
    !is.na(foraging_guild), foraging_guild != "",
    !is.na(species_thermal_index),
    foraging_guild %in% foraging_driver_order
  ) |>
  group_by(foraging_guild) |>
  summarise(
    trait_value = first(foraging_guild),
    mean_sti = mean(species_thermal_index, na.rm = TRUE),
    sti_q025 = quantile(species_thermal_index, 0.025, na.rm = TRUE),
    sti_q975 = quantile(species_thermal_index, 0.975, na.rm = TRUE),
    n_species = n(),
    .groups = "drop"
  ) |>
  mutate(
    trait_value = factor(trait_value, levels = foraging_driver_order)
  )

guild_habitat_strip <- ggplot(
  guild_driver_annotations,
  aes(x = "Habitat", y = trait_value, fill = foraging_habitat)
) +
  geom_tile(colour = "white", linewidth = 0.25) +
  scale_fill_manual(
    values = c("Aquatic/wetland" = "#2b8cbe", "Terrestrial/aerial" = "#7f8f3a"),
    name = "Guild habitat"
  ) +
  labs(x = NULL, y = NULL) +
  theme_minimal(base_size = 10) +
  theme(
    panel.grid = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.ticks = element_blank(),
    legend.position = "bottom"
  )

guild_migration_bar <- ggplot(
  guild_migration_composition,
  aes(x = proportion, y = trait_value, fill = migratory_strategy)
) +
  geom_col(width = 0.72, colour = "white", linewidth = 0.15) +
  scale_x_continuous(
    labels = percent_format(accuracy = 1),
    breaks = c(0, 0.5, 1),
    limits = c(0, 1),
    expand = expansion(mult = c(0, 0))
  ) +
  scale_fill_manual(
    values = migration_palette,
    breaks = migration_levels,
    drop = FALSE,
    name = "Migratory strategy"
  ) +
  labs(x = "Migration", y = NULL) +
  guides(fill = guide_legend(nrow = 2, byrow = TRUE)) +
  theme_minimal(base_size = 10) +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    plot.margin = margin(5.5, 14, 5.5, 5.5),
    legend.position = "bottom"
  )

guild_sti_forest_plot <- ggplot(
  guild_sti_summary,
  aes(y = trait_value)
) +
  geom_vline(xintercept = overall_mean_sti, colour = "grey35", linewidth = 0.35, linetype = "dashed") +
  geom_segment(
    aes(x = sti_q025, xend = sti_q975, yend = trait_value),
    colour = "grey45",
    linewidth = 0.55
  ) +
  geom_point(aes(x = mean_sti, colour = mean_sti), size = 2.4) +
  scale_x_continuous(
    expand = expansion(mult = c(0.04, 0.04))
  ) +
  scale_colour_gradientn(
    colours = c("blue4", "darkorchid", "red2"),
    name = "Mean STI"
  ) +
  labs(x = "STI", y = NULL) +
  theme_minimal(base_size = 10) +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    plot.margin = margin(5.5, 14, 5.5, 5.5),
    legend.position = "bottom"
  )

cluster2_shift_foraging_heatmap <- ggplot(
  foraging_driver_df,
  aes(x = transition, y = trait_value, fill = mean_expected_richness_delta_per_species)
) +
  geom_tile(colour = "white", linewidth = 0.25) +
  scale_fill_gradient2(
    low = "#d95f02",
    mid = "white",
    high = "#1b9e77",
    midpoint = 0,
    name = "Mean expected\nrichness delta\nper species"
  ) +
  labs(
    x = NULL,
    y = NULL
  ) +
  theme_minimal(base_size = 10) +
  theme(
    panel.grid = element_blank(),
    axis.text.x = element_text(angle = 35, hjust = 1),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    plot.margin = margin(5.5, 5.5, 5.5, 5.5)
  )

cluster2_shift_foraging_plot <- (
  guild_habitat_strip + guild_migration_bar + guild_sti_forest_plot + cluster2_shift_foraging_heatmap +
    plot_layout(widths = c(1.05, 2.1, 2.1, 3.4), guides = "collect")
) +
  plot_annotation(
    title = "Foraging-guild drivers of shifts into Cluster 2 by the 2010s",
    subtitle = "The STI forest plot shows guild mean and central 95% species interval; the dashed line is the overall mean STI."
  ) &
  theme(
    legend.position = "right",
    legend.text = element_text(size = 9),
    legend.title = element_text(size = 10),
    plot.title = element_text(face = "bold")
  )

cluster2_shift_foraging_plot
```

## Export Results

```{r export-results}
write_csv(
  cluster_trait_composition,
  file.path(out_dir, paste0(output_prefix, "-cluster-trait-composition.csv"))
)
write_csv(
  cluster_trait_dominant_groups,
  file.path(out_dir, paste0(output_prefix, "-cluster-trait-composition-dominant-groups.csv"))
)
write_csv(
  cluster_top_modelled_species,
  file.path(out_dir, paste0(output_prefix, "-cluster-top-modelled-species.csv"))
)
write_csv(
  cluster_trait_composition_checks,
  file.path(out_dir, paste0(output_prefix, "-cluster-trait-composition-checks.csv"))
)
write_csv(
  reference_environment_values,
  file.path(out_dir, paste0(output_prefix, "-reference-cluster-environment-values.csv"))
)
write_csv(
  reference_environment_summary,
  file.path(out_dir, paste0(output_prefix, "-reference-cluster-environment-summary.csv"))
)
write_csv(
  reference_environment_tests,
  file.path(out_dir, paste0(output_prefix, "-reference-cluster-environment-tests.csv"))
)
write_csv(
  cluster2_shift_focal_sites,
  file.path(out_dir, paste0(pattern, "-cluster2-shift-focal-sites.csv"))
)
write_csv(
  cluster2_shift_species_deltas,
  file.path(out_dir, paste0(pattern, "-cluster2-shift-species-deltas.csv"))
)
write_csv(
  cluster2_shift_top_species_drivers,
  file.path(out_dir, paste0(pattern, "-cluster2-shift-top-species-drivers.csv"))
)
write_csv(
  cluster2_shift_trait_deltas,
  file.path(out_dir, paste0(pattern, "-cluster2-shift-trait-deltas.csv"))
)
write_csv(
  cluster2_shift_trait_deltas |>
    select(
      comparison, transition, source_cluster, target_cluster, trait_family,
      trait_value, n_species_in_trait_group,
      mean_expected_richness_delta_per_species,
      median_expected_richness_delta_per_species,
      q10_expected_richness_delta_per_species,
      q90_expected_richness_delta_per_species,
      mean_expected_richness_gain_per_species,
      mean_expected_richness_loss_per_species,
      absolute_change_share_per_species,
      prop_sites_increase,
      everything()
    ),
  file.path(out_dir, paste0(pattern, "-cluster2-shift-trait-deltas-size-normalized.csv"))
)
write_csv(
  environment_matched_sites,
  file.path(out_dir, paste0(pattern, "-cluster2-shift-environment-deltas.csv"))
)
write_csv(
  environment_retained_comparison_counts,
  file.path(out_dir, paste0(pattern, "-cluster2-shift-environment-retained-comparison-counts.csv"))
)
write_csv(
  environment_delta_summary,
  file.path(out_dir, paste0(pattern, "-cluster2-shift-environment-delta-summary.csv"))
)
write_csv(
  environment_delta_tests,
  file.path(out_dir, paste0(pattern, "-cluster2-shift-environment-delta-tests.csv"))
)
write_csv(
  cluster2_shift_checks,
  file.path(out_dir, paste0(pattern, "-cluster2-shift-checks.csv"))
)

ggsave(
  file.path(out_dir, paste0(output_prefix, "-cluster-thermal-migration-composition.png")),
  thermal_migration_composition_plot,
  width = 8.5,
  height = 7,
  units = "in",
  dpi = 300,
  bg = "white"
)
ggsave(
  file.path(out_dir, paste0(output_prefix, "-cluster-foraging-guild-composition-heatmap.png")),
  foraging_guild_composition_heatmap,
  width = 8.5,
  height = 9.5,
  units = "in",
  dpi = 300,
  bg = "white"
)
ggsave(
  file.path(out_dir, paste0(output_prefix, "-reference-cluster-environment-profile.png")),
  reference_environment_profile_plot,
  width = 10.5,
  height = 6.3,
  units = "in",
  dpi = 300,
  bg = "white"
)
obsolete_species_plot_path <- file.path(out_dir, paste0(pattern, "-cluster2-shift-species-drivers.png"))
if (file.exists(obsolete_species_plot_path)) {
  unlink(obsolete_species_plot_path)
}

ggsave(
  file.path(out_dir, paste0(pattern, "-cluster2-shift-thermal-migration-drivers-size-normalized.png")),
  cluster2_shift_thermal_migration_plot,
  width = 10.5,
  height = 6.5,
  units = "in",
  dpi = 300,
  bg = "white"
)
ggsave(
  file.path(out_dir, paste0(pattern, "-cluster2-shift-environment-context.png")),
  cluster2_shift_environment_plot,
  width = 10.5,
  height = 6.5,
  units = "in",
  dpi = 300,
  bg = "white"
)
ggsave(
  file.path(out_dir, paste0(pattern, "-cluster2-shift-foraging-guild-drivers-size-normalized.png")),
  cluster2_shift_foraging_plot,
  width = 16,
  height = 8.8,
  units = "in",
  dpi = 300,
  bg = "white"
)
```
