---
title: "1970s avian bioregion cluster species and environmental driver profiles"
author: "WM"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  results = "hide",
  warning = FALSE,
  message = FALSE,
  fig.align = "center"
)
knitr::opts_knit$set(root.dir = normalizePath("../.."))
```

## Aim

This notebook asks which species characterize the 1970s modelled avian
bioregion clusters, and which environmental variables are most important for
those species.

The clusters are the Atlas 1 reference clusters from the avian bioregion PCA.
Species are not assigned to clusters with a hard cutoff. Instead, each species
gets a cluster-specific weight equal to its share of total modelled expected
occurrence in that 1970s cluster. Environmental driver summaries then use those
species weights to summarize Atlas 1 scaled variance partitioning (VP) and
supported Atlas 1 Beta effects.

## Load Data

```{r load-data}
if (!require("pacman")) install.packages("pacman")
pacman::p_load(
  tidyverse, scales, patchwork, knitr
)

source(file.path("support_scripts", "figure_data_helpers.R"))

base_dir <- "./HmscOutputs"
pattern <- "2026-03-13"
focal_atlas <- "1"
focal_period <- "1970s"

bioregion_assignment_path <- file.path(
  "notebooks", "exploratory", "outputs", "bioregion-pca",
  paste0(pattern, "-bioregion-assignments.csv")
)
trait_metadata_path <- file.path(
  "Data", "preprocessed_data_min_occs_is_5_coverage_is_good_and_average.RData"
)
out_dir <- file.path(
  "notebooks", "exploratory", "outputs", "bioregion-cluster-vp-effects"
)
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

output_prefix <- paste0(pattern, "-", focal_period)

matching_folders <- figure_model_folders(pattern = pattern, base_dir = base_dir)
model_nums <- as.character(atlas_numbers(matching_folders))
names(matching_folders) <- model_nums
atlas1_folder <- matching_folders[[focal_atlas]]

prediction_path <- file.path(
  base_dir,
  atlas1_folder,
  "Results", "Preds", "predicted-values.rdata"
)
if (!file.exists(prediction_path)) {
  stop("Missing fitted-site prediction file: ", prediction_path, call. = FALSE)
}
preds_atlas1 <- readRDS(prediction_path)

bioregion_assignments <- read_csv(
  bioregion_assignment_path,
  show_col_types = FALSE
) |>
  mutate(atlas = as.character(.data$atlas)) |>
  filter(.data$atlas == focal_atlas)

load(trait_metadata_path)
if (!exists("Tr")) {
  stop("Expected object `Tr` in trait metadata file.", call. = FALSE)
}
```

```{r shared-definitions}
cluster_levels <- paste("Cluster", 1:3)
cluster_recode <- c(
  "Cluster 1" = "Cluster 1",
  "Cluster 3" = "Cluster 2",
  "Cluster 2" = "Cluster 3"
)
relabel_clusters <- function(x) {
  factor(unname(cluster_recode[as.character(x)]), levels = cluster_levels)
}

thermal_levels <- c("very cold", "cold", "medium", "warm", "very warm")
thermal_colours <- c("#313695", "#74ADD1", "#FEE090", "#F46D43", "#9E0142")
thermal_palette <- setNames(thermal_colours, thermal_levels)

environment_order <- c(
  "Temperature",
  "Precipitation",
  "Land-use heterogeneity",
  "Urban (% coverage)",
  "Cropland (% coverage)",
  "Pasture (% coverage)",
  "Forest (% coverage)",
  "Grass/Shrubland (% coverage)"
)
var_labels <- c(
  "tmean_breeding" = "Temperature",
  "prec_breeding" = "Precipitation",
  "hh" = "Land-use heterogeneity",
  "perc_urban" = "Urban (% coverage)",
  "perc_cropland" = "Cropland (% coverage)",
  "perc_pasture" = "Pasture (% coverage)",
  "perc_forest" = "Forest (% coverage)",
  "perc_grass_shrub" = "Grass/Shrubland (% coverage)"
)
environment_colours <- c(
  "Temperature" = "firebrick3",
  "Precipitation" = "dodgerblue3",
  "Land-use heterogeneity" = "orchid3",
  "Urban (% coverage)" = "snow3",
  "Cropland (% coverage)" = "goldenrod1",
  "Pasture (% coverage)" = "darkorange",
  "Forest (% coverage)" = "springgreen4",
  "Grass/Shrubland (% coverage)" = "springgreen2"
)

thermal_groups <- function(sti) {
  cut(
    sti,
    breaks = quantile(sti, seq(0, 1, 0.2), na.rm = TRUE),
    include.lowest = TRUE,
    labels = thermal_levels
  )
}

species_trait_table <- Tr |>
  as.data.frame() |>
  rownames_to_column("species") |>
  transmute(
    species,
    thermal_group = thermal_groups(.data$species_thermal_index),
    migratory_strategy = .data$Migration_a3_DOF,
    foraging_guild = .data$foraging_guild_consensus,
    species_thermal_index = .data$species_thermal_index
  ) |>
  mutate(
    thermal_group = factor(.data$thermal_group, levels = thermal_levels)
  )
```

## Methods

For each 1970s cluster, predicted species probabilities were summed across all
grid cells in the cluster. A species weight is the species' expected occurrence
sum divided by the total expected occurrence summed over all species in that
cluster. The weights therefore describe the modelled composition of each
cluster.

Scaled VP values are Atlas 1 variable contributions multiplied by species Tjur
R2, following the Fig. 2 workflow. Beta effects are used only where posterior
support is high enough for `read_parameter_effects()` to retain the estimate.
Unsupported effects are treated as unknown direction, not as zero effects.

The waffle chart combines these two pieces of information. For every supported
`species x environmental variable` Beta association, the contribution is:

`species expected occurrence share in the 1970s cluster x species scaled VP for
that environmental variable`

The resulting values are summed by `Cluster x Beta sign x environmental
variable`. Each waffle is then normalized within a `Cluster x Beta sign`
combination, so the tiles show the relative composition of weighted supported
environmental dependency among positive or negative species associations.

The PCA uses the same species-level dependency profiles as the waffle chart,
but retains the raw multivariate structure before aggregation. For each
`Cluster x species`, each environmental variable is split into positive and
negative supported-dependency features. Cluster 3 species profiles are used to
fit the reference PCA, and Cluster 1 and Cluster 2 species profiles are then
projected into that same Cluster 3 environmental-dependency space.

## Data Preparation

```{r species-weights}
pred_surveys <- rownames(preds_atlas1)

assignments_atlas1 <- bioregion_assignments |>
  mutate(
    bioregion_original = .data$bioregion,
    bioregion = relabel_clusters(.data$bioregion)
  ) |>
  select(survey, site, X, Y, bioregion_original, bioregion)

site_match_summary <- tibble(
  check = c(
    "Prediction rows",
    "Assignment rows",
    "Shared surveys",
    "Prediction rows missing assignment",
    "Assignment rows missing prediction"
  ),
  value = c(
    length(pred_surveys),
    nrow(assignments_atlas1),
    length(intersect(pred_surveys, assignments_atlas1$survey)),
    length(setdiff(pred_surveys, assignments_atlas1$survey)),
    length(setdiff(assignments_atlas1$survey, pred_surveys))
  )
)

if (site_match_summary$value[site_match_summary$check == "Prediction rows missing assignment"] > 0 ||
    site_match_summary$value[site_match_summary$check == "Assignment rows missing prediction"] > 0) {
  stop("Atlas 1 prediction rows and bioregion assignments do not align.", call. = FALSE)
}

assignments_ordered <- assignments_atlas1 |>
  slice(match(pred_surveys, .data$survey))

cluster_species_sums <- rowsum(
  preds_atlas1,
  group = assignments_ordered$bioregion,
  reorder = FALSE
)
n_sites_by_cluster <- assignments_ordered |>
  count(.data$bioregion, name = "n_sites")

cluster_totals <- rowSums(cluster_species_sums, na.rm = TRUE)

species_weights <- cluster_species_sums |>
  as.data.frame() |>
  rownames_to_column("bioregion") |>
  pivot_longer(
    cols = -bioregion,
    names_to = "species",
    values_to = "expected_occurrence_sum"
  ) |>
  mutate(
    bioregion = factor(.data$bioregion, levels = cluster_levels),
    expected_occurrence_share = .data$expected_occurrence_sum / cluster_totals[as.character(.data$bioregion)]
  ) |>
  left_join(n_sites_by_cluster, by = "bioregion") |>
  mutate(
    mean_predicted_probability = .data$expected_occurrence_sum / .data$n_sites
  ) |>
  group_by(.data$bioregion) |>
  arrange(desc(.data$expected_occurrence_share), .by_group = TRUE) |>
  mutate(rank_in_cluster = row_number()) |>
  ungroup() |>
  left_join(species_trait_table, by = "species") |>
  mutate(
    species_label = str_replace_all(.data$species, "_", " ")
  )

top_weighted_species <- species_weights |>
  group_by(.data$bioregion) |>
  slice_max(
    order_by = .data$expected_occurrence_share,
    n = 20,
    with_ties = FALSE
  ) |>
  ungroup()

species_weight_checks <- species_weights |>
  group_by(.data$bioregion) |>
  summarise(
    n_species = n(),
    n_sites = first(.data$n_sites),
    weight_sum = sum(.data$expected_occurrence_share, na.rm = TRUE),
    weight_sum_deviation = abs(.data$weight_sum - 1),
    .groups = "drop"
  )
```

```{r vp-and-beta}
vp_scaled <- load_vp_estimates(matching_folders, base_dir = base_dir, scaled = TRUE)
vp_atlas1 <- vp_scaled[[focal_atlas]]

vp_long <- vp_atlas1 |>
  as.data.frame() |>
  rownames_to_column("variable") |>
  filter(.data$variable %in% names(var_labels)) |>
  pivot_longer(
    cols = -variable,
    names_to = "species",
    values_to = "scaled_vp"
  ) |>
  mutate(
    variable_label = unname(var_labels[.data$variable]),
    variable_label = factor(.data$variable_label, levels = environment_order)
  )

beta_atlas1 <- read_parameter_effects(pattern, effect = "Beta", base_dir = base_dir) |>
  filter(
    .data$atlas == as.numeric(focal_atlas),
    .data$variable %in% names(var_labels)
  ) |>
  mutate(
    supported = !is.na(.data$effect_size),
    effect_sign = case_when(
      .data$effect_size > 0 ~ 1,
      .data$effect_size < 0 ~ -1,
      TRUE ~ NA_real_
    ),
    effect_direction = case_when(
      .data$effect_size > 0 ~ "Positive",
      .data$effect_size < 0 ~ "Negative",
      TRUE ~ "Unsupported"
    )
  )

cluster_species_vp_effects <- species_weights |>
  select(
    bioregion,
    species,
    species_label,
    species_weight = expected_occurrence_share,
    rank_in_cluster,
    thermal_group,
    migratory_strategy,
    foraging_guild
  ) |>
  inner_join(vp_long, by = "species", relationship = "many-to-many") |>
  left_join(
    beta_atlas1 |>
      select(species, variable, effect_size, supported, effect_sign, effect_direction),
    by = c("species", "variable")
  ) |>
  mutate(
    supported = replace_na(.data$supported, FALSE),
    effect_direction = replace_na(.data$effect_direction, "Unsupported"),
    weighted_scaled_vp = .data$species_weight * .data$scaled_vp,
    weighted_supported_vp = if_else(
      .data$supported,
      .data$weighted_scaled_vp,
      0
    ),
    weighted_signed_vp = if_else(
      .data$supported,
      .data$weighted_scaled_vp * .data$effect_sign,
      0
    )
  )
```

## Weighted Species-Level Driver Waffles

This plot keeps the analysis close to individual species, but summarizes the
species-level information as proportional waffles rather than a large species
table. All species in each 1970s cluster are included. For each supported
`species x environmental variable` Beta association, the contribution is the
species' expected occurrence share in that cluster multiplied by the species'
scaled VP for that environmental variable. The waffle therefore asks: among the
modelled species that make up a cluster, which environmental dependencies are
most represented among positive and negative supported associations?

Each waffle is normalized within a `Cluster x Beta sign` combination. A tile is
approximately 1% of the weighted supported dependency in that waffle. The tile
colours show the environmental variables.

```{r species-driver-cloud-data}
cluster_driver_cloud_data <- cluster_species_vp_effects |>
  filter(.data$supported) |>
  mutate(
    effect_direction = factor(
      .data$effect_direction,
      levels = c("Positive", "Negative")
    ),
    effect_sign_label = case_when(
      .data$effect_direction == "Positive" ~ "+",
      .data$effect_direction == "Negative" ~ "-",
      TRUE ~ NA_character_
    )
  )

cluster_driver_cloud_counts <- species_weights |>
  group_by(.data$bioregion) |>
  summarise(
    n_species = n_distinct(.data$species),
    total_expected_occurrence_share = sum(.data$expected_occurrence_share),
    .groups = "drop"
  )

cluster_driver_waffle_summary <- cluster_driver_cloud_data |>
  group_by(.data$bioregion, .data$effect_direction, .data$variable_label) |>
  summarise(
    n_supported_species = n_distinct(.data$species),
    weighted_dependency = sum(.data$weighted_supported_vp, na.rm = TRUE),
    mean_species_weight = mean(.data$species_weight, na.rm = TRUE),
    .groups = "drop"
  ) |>
  complete(
    bioregion = factor(cluster_levels, levels = cluster_levels),
    effect_direction = factor(c("Positive", "Negative"), levels = c("Positive", "Negative")),
    variable_label = factor(environment_order, levels = environment_order),
    fill = list(
      n_supported_species = 0,
      weighted_dependency = 0,
      mean_species_weight = NA_real_
    )
  ) |>
  group_by(.data$bioregion, .data$effect_direction) |>
  mutate(
    total_weighted_dependency = sum(.data$weighted_dependency, na.rm = TRUE),
    dependency_share = if_else(
      .data$total_weighted_dependency > 0,
      .data$weighted_dependency / .data$total_weighted_dependency,
      0
    )
  ) |>
  ungroup()

allocate_waffle_tiles <- function(data, n_tiles = 100) {
  data <- data |>
    arrange(.data$variable_label) |>
    mutate(
      raw_tiles = .data$dependency_share * n_tiles,
      tile_n = floor(.data$raw_tiles),
      remainder = .data$raw_tiles - .data$tile_n
    )

  missing_tiles <- n_tiles - sum(data$tile_n, na.rm = TRUE)
  if (missing_tiles > 0 && sum(data$dependency_share, na.rm = TRUE) > 0) {
    add_index <- data |>
      arrange(desc(.data$remainder), desc(.data$dependency_share)) |>
      slice_head(n = missing_tiles) |>
      mutate(add_tile = 1) |>
      select(variable_label, add_tile)

    data <- data |>
      left_join(add_index, by = "variable_label") |>
      mutate(tile_n = .data$tile_n + replace_na(.data$add_tile, 0)) |>
      select(-add_tile)
  }

  data |>
    filter(.data$tile_n > 0) |>
    uncount(.data$tile_n, .id = "tile_within_variable") |>
    arrange(.data$variable_label, .data$tile_within_variable) |>
    mutate(
      tile_id = row_number(),
      tile_col = ((.data$tile_id - 1) %% 10) + 1,
      tile_row = 10 - ((.data$tile_id - 1) %/% 10)
    )
}

cluster_driver_waffle_tiles <- cluster_driver_waffle_summary |>
  group_by(.data$bioregion, .data$effect_direction) |>
  group_modify(~allocate_waffle_tiles(.x, n_tiles = 100)) |>
  ungroup() |>
  mutate(
    effect_direction = factor(
      .data$effect_direction,
      levels = c("Positive", "Negative")
    ),
    variable_label = factor(.data$variable_label, levels = environment_order)
  )
```

```{r species-driver-cloud-plot}
cluster_species_driver_cloud_plot <- cluster_driver_waffle_tiles |>
  ggplot(aes(
    x = .data$tile_col,
    y = .data$tile_row,
    fill = .data$variable_label
  )) +
  geom_tile(
    width = 0.92,
    height = 0.92,
    colour = "white",
    linewidth = 0.2
  ) +
  facet_grid(
    effect_direction ~ bioregion
  ) +
  scale_fill_manual(
    values = environment_colours,
    breaks = environment_order,
    drop = FALSE,
    name = "Environmental variable"
  ) +
  coord_equal(clip = "off") +
  labs(
    title = NULL,
    subtitle = NULL,
    x = NULL,
    y = NULL
  ) +
  scale_x_continuous(expand = expansion(mult = 0), breaks = NULL) +
  scale_y_continuous(expand = expansion(mult = 0), breaks = NULL) +
  theme_minimal(base_size = 10) +
  theme(
    panel.grid = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text = element_blank(),
    strip.text = element_text(face = "bold"),
    panel.spacing = unit(0.55, "lines"),
    legend.position = "right",
    legend.direction = 'vertical',
    legend.key.size = unit(0.35, "cm")
  )

```

## Cluster 3 Reference PCA

```{r cluster3-reference-pca, fig.width=10.5, fig.height=11.6}
dependency_feature_order <- as.vector(outer(
  environment_order,
  c("Positive", "Negative"),
  paste,
  sep = " | "
))

species_dependency_profiles <- cluster_species_vp_effects |>
  mutate(
    effect_direction = factor(
      .data$effect_direction,
      levels = c("Positive", "Negative", "Unsupported")
    ),
    dependency_feature = paste(.data$variable_label, .data$effect_direction, sep = " | "),
    dependency_value = if_else(
      .data$supported,
      .data$weighted_supported_vp,
      0
    )
  ) |>
  filter(.data$effect_direction %in% c("Positive", "Negative")) |>
  select(
    bioregion,
    species,
    species_label,
    species_weight,
    dependency_feature,
    dependency_value
  ) |>
  complete(
    bioregion = factor(cluster_levels, levels = cluster_levels),
    species,
    dependency_feature = dependency_feature_order,
    fill = list(dependency_value = 0)
  ) |>
  left_join(
    species_weights |>
      select(bioregion, species, species_label, species_weight = expected_occurrence_share),
    by = c("bioregion", "species")
  ) |>
  mutate(
    species_label = coalesce(.data$species_label.x, .data$species_label.y),
    species_weight = coalesce(.data$species_weight.x, .data$species_weight.y)
  ) |>
  select(
    bioregion,
    species,
    species_label,
    species_weight,
    dependency_feature,
    dependency_value
  ) |>
  distinct()

species_dependency_matrix <- species_dependency_profiles |>
  mutate(
    dependency_feature = factor(
      .data$dependency_feature,
      levels = dependency_feature_order
    )
  ) |>
  pivot_wider(
    id_cols = c(bioregion, species, species_label, species_weight),
    names_from = dependency_feature,
    values_from = dependency_value,
    values_fill = 0
  )

dependency_feature_cols <- intersect(dependency_feature_order, names(species_dependency_matrix))
cluster3_matrix <- species_dependency_matrix |>
  filter(.data$bioregion == "Cluster 3") |>
  select(all_of(dependency_feature_cols)) |>
  as.matrix()

feature_sds <- apply(cluster3_matrix, 2, sd, na.rm = TRUE)
pca_feature_cols <- names(feature_sds)[is.finite(feature_sds) & feature_sds > 0]

cluster3_dependency_pca <- prcomp(
  cluster3_matrix[, pca_feature_cols, drop = FALSE],
  center = TRUE,
  scale. = TRUE
)

pca_scores <- predict(
  cluster3_dependency_pca,
  newdata = species_dependency_matrix |>
    select(all_of(pca_feature_cols)) |>
    as.matrix()
) |>
  as.data.frame() |>
  bind_cols(
    species_dependency_matrix |>
      select(bioregion, species, species_label, species_weight)
  ) |>
  mutate(
    bioregion = factor(.data$bioregion, levels = cluster_levels),
    is_reference_cluster = .data$bioregion == "Cluster 3"
  )

pca_variance <- summary(cluster3_dependency_pca)$importance["Proportion of Variance", ]
pca_axis_labels <- c(
  paste0("PC1 (", percent(pca_variance[1], accuracy = 0.1), ")"),
  paste0("PC2 (", percent(pca_variance[2], accuracy = 0.1), ")")
)

pca_loadings <- as.data.frame(cluster3_dependency_pca$rotation) |>
  rownames_to_column("dependency_feature") |>
  separate(
    .data$dependency_feature,
    into = c("variable_label", "effect_direction"),
    sep = " \\| ",
    remove = FALSE
  ) |>
  mutate(
    variable_label = factor(.data$variable_label, levels = environment_order),
    effect_direction = factor(.data$effect_direction, levels = c("Positive", "Negative")),
    loading_strength = sqrt(.data$PC1^2 + .data$PC2^2),
    loading_label = str_replace_all(
      as.character(.data$variable_label),
      c(
        " \\(% coverage\\)" = "",
        "Land-use heterogeneity" = "Heterogeneity",
        "Grass/Shrubland" = "Grass/shrub"
      )
    ),
    loading_label = paste0(
      .data$loading_label,
      if_else(.data$effect_direction == "Positive", " +", " -")
    )
  )

pca_centroids <- pca_scores |>
  group_by(.data$bioregion) |>
  summarise(
    PC1 = mean(.data$PC1, na.rm = TRUE),
    PC2 = mean(.data$PC2, na.rm = TRUE),
    .groups = "drop"
  )

pca_x_limits <- quantile(pca_scores$PC1, c(0.01, 0.985), na.rm = TRUE)
pca_y_limits <- quantile(pca_scores$PC2, c(0.01, 0.99), na.rm = TRUE)

cluster_palette <- c(
  "Cluster 1" = "#76B7D8",
  "Cluster 2" = "#F6E58D",
  "Cluster 3" = "#C49A00"
)

cluster3_reference_pca_plot <- pca_scores |>
  ggplot(aes(
    x = .data$PC1,
    y = .data$PC2,
    colour = .data$bioregion
  )) +
  geom_hline(yintercept = 0, colour = "grey85", linewidth = 0.3) +
  geom_vline(xintercept = 0, colour = "grey85", linewidth = 0.3) +
  geom_point(
    aes(size = .data$species_weight, alpha = .data$is_reference_cluster),
    shape = 16
  ) +
  geom_point(
    data = pca_centroids,
    aes(x = .data$PC1, y = .data$PC2, fill = .data$bioregion),
    inherit.aes = FALSE,
    shape = 21,
    colour = "grey15",
    size = 5,
    stroke = 0.55
  ) +
  geom_text(
    data = pca_centroids,
    aes(x = .data$PC1, y = .data$PC2, label = .data$bioregion),
    inherit.aes = FALSE,
    position = position_nudge(
      x = c(0.45, -0.4, -0.45),
      y = c(0.28, -0.32, 0.32)
    ),
    size = 3.3,
    fontface = "bold"
  ) +
  scale_colour_manual(values = cluster_palette, name = "Cluster") +
  scale_fill_manual(values = cluster_palette, guide = "none") +
  scale_alpha_manual(
    values = c("TRUE" = 0.75, "FALSE" = 0.32),
    guide = "none"
  ) +
  scale_size_continuous(
    range = c(0.8, 4.5),
    labels = percent_format(accuracy = 0.1),
    name = "Expected occurrence\nshare"
  ) +
  labs(
    title = NULL,
    subtitle = NULL,
    x = pca_axis_labels[[1]],
    y = pca_axis_labels[[2]]
  ) +
  coord_cartesian(
    xlim = pca_x_limits,
    ylim = pca_y_limits,
    clip = "off"
  ) +
  theme_minimal(base_size = 10) +
  theme(
    panel.grid.minor = element_blank(),
    legend.position = "bottom"
  )

pca_loading_limit <- max(abs(c(pca_loadings$PC1, pca_loadings$PC2)), na.rm = TRUE) * 1.18

pca_loading_plot <- ggplot(pca_loadings, aes(x = .data$PC1, y = .data$PC2)) +
  geom_hline(yintercept = 0, colour = "grey85", linewidth = 0.3) +
  geom_vline(xintercept = 0, colour = "grey85", linewidth = 0.3) +
  geom_segment(
    aes(
      x = 0,
      y = 0,
      xend = .data$PC1,
      yend = .data$PC2,
      colour = .data$variable_label,
      linetype = .data$effect_direction
    ),
    linewidth = 0.45,
    arrow = arrow(length = unit(0.08, "in"))
  ) +
  geom_point(
    data = filter(pca_loadings, .data$effect_direction == "Positive"),
    aes(size = .data$loading_strength, colour = .data$variable_label, fill = .data$variable_label),
    shape = 21,
    stroke = 0.5
  ) +
  geom_point(
    data = filter(pca_loadings, .data$effect_direction == "Negative"),
    aes(size = .data$loading_strength, colour = .data$variable_label),
    shape = 21,
    fill = "white",
    stroke = 0.8
  ) +
  geom_text(
    aes(label = .data$loading_label),
    size = 2.65,
    hjust = if_else(pca_loadings$PC1 >= 0, -0.08, 1.08),
    vjust = if_else(pca_loadings$PC2 >= 0, -0.15, 1.15),
    check_overlap = TRUE
  ) +
  scale_colour_manual(values = environment_colours, guide = "none") +
  scale_fill_manual(values = environment_colours, guide = "none") +
  scale_linetype_manual(
    values = c("Positive" = "solid", "Negative" = "dotted"),
    guide = "none"
  ) +
  scale_size_continuous(range = c(2, 7), guide = "none") +
  coord_equal(
    xlim = c(-pca_loading_limit, pca_loading_limit),
    ylim = c(-pca_loading_limit, pca_loading_limit),
    clip = "off"
  ) +
  labs(
    title = "Environmental loadings",
    subtitle = "Solid/filled = positive Beta feature; dotted/open = negative Beta feature.",
    x = pca_axis_labels[[1]],
    y = pca_axis_labels[[2]]
  ) +
  theme_minimal(base_size = 10) +
  theme(
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold"),
    plot.margin = margin(5.5, 18, 5.5, 5.5)
  )

pca_row_plot <- (cluster3_reference_pca_plot + theme(legend.position = "none")) |
  pca_loading_plot +
  plot_layout(widths = c(1, 0.72))

cluster_driver_composite_plot <- cluster_species_driver_cloud_plot /
  pca_row_plot +
  plot_layout(heights = c(1.2, 1), guides = "keep")

cluster_driver_composite_plot
```

## Checks And Exports

```{r checks-and-exports}
species_alignment_checks <- tibble(
  check = c(
    "Prediction species",
    "VP species",
    "Beta species",
    "Prediction species in VP",
    "Prediction species in Beta",
    "Prediction species with trait rows"
  ),
  value = c(
    ncol(preds_atlas1),
    n_distinct(vp_long$species),
    n_distinct(beta_atlas1$species),
    length(intersect(colnames(preds_atlas1), vp_long$species)),
    length(intersect(colnames(preds_atlas1), beta_atlas1$species)),
    length(intersect(colnames(preds_atlas1), species_trait_table$species))
  )
)

variable_checks <- tibble(
  check = c(
    "Expected environmental variables",
    "VP environmental variables",
    "Driver waffle clusters",
    "Driver waffle species",
    "Driver waffle supported species-variable rows",
    "Driver waffle tiles",
    "PCA reference features",
    "PCA projected species profiles"
  ),
  value = c(
    length(environment_order),
    n_distinct(vp_long$variable_label),
    n_distinct(cluster_driver_cloud_data$bioregion),
    n_distinct(paste(cluster_driver_cloud_data$bioregion, cluster_driver_cloud_data$species)),
    nrow(cluster_driver_cloud_data),
    nrow(cluster_driver_waffle_tiles),
    length(pca_feature_cols),
    nrow(pca_scores)
  )
)

cluster_vp_effect_checks <- bind_rows(
  site_match_summary,
  species_weight_checks |>
    transmute(
      check = paste0(as.character(.data$bioregion), " species weight sum"),
      value = .data$weight_sum
    ),
  species_alignment_checks,
  variable_checks
)

write_csv(
  species_weights,
  file.path(out_dir, paste0(output_prefix, "-cluster-species-weights.csv"))
)
write_csv(
  cluster_driver_cloud_data,
  file.path(out_dir, paste0(output_prefix, "-cluster-species-driver-cloud-data.csv"))
)
write_csv(
  cluster_driver_waffle_summary,
  file.path(out_dir, paste0(output_prefix, "-cluster-species-driver-waffle-summary.csv"))
)
write_csv(
  cluster_driver_waffle_tiles,
  file.path(out_dir, paste0(output_prefix, "-cluster-species-driver-waffle-tiles.csv"))
)
write_csv(
  pca_scores,
  file.path(out_dir, paste0(output_prefix, "-cluster3-reference-dependency-pca-scores.csv"))
)
write_csv(
  as.data.frame(cluster3_dependency_pca$rotation) |>
    rownames_to_column("dependency_feature"),
  file.path(out_dir, paste0(output_prefix, "-cluster3-reference-dependency-pca-loadings.csv"))
)
write_csv(
  pca_loadings,
  file.path(out_dir, paste0(output_prefix, "-cluster3-reference-dependency-pca-loadings-annotated.csv"))
)
write_csv(
  cluster_vp_effect_checks,
  file.path(out_dir, paste0(output_prefix, "-cluster-vp-effect-checks.csv"))
)

ggsave(
  file.path(out_dir, paste0(output_prefix, "-cluster-species-driver-cloud.png")),
  cluster_species_driver_cloud_plot,
  width = 10.2,
  height = 6.7,
  units = "in",
  dpi = 300,
  bg = "white"
)
ggsave(
  file.path(out_dir, paste0(output_prefix, "-cluster-species-driver-waffle.png")),
  cluster_species_driver_cloud_plot,
  width = 10.2,
  height = 6.7,
  units = "in",
  dpi = 300,
  bg = "white"
)
ggsave(
  file.path(out_dir, paste0(output_prefix, "-cluster3-reference-dependency-pca.png")),
  cluster3_reference_pca_plot,
  width = 10.2,
  height = 5.7,
  units = "in",
  dpi = 300,
  bg = "white"
)
ggsave(
  file.path(out_dir, paste0(output_prefix, "-cluster3-reference-dependency-pca-loadings.png")),
  pca_loading_plot,
  width = 6.2,
  height = 5.7,
  units = "in",
  dpi = 300,
  bg = "white"
)
ggsave(
  file.path(out_dir, paste0(output_prefix, "-cluster-species-driver-waffle-pca-composite.png")),
  cluster_driver_composite_plot,
  width = 13,
  height = 12,
  units = "in",
  dpi = 300,
  bg = "white"
)
```
