---
title: "Modelled avian bioregion and environmental-response PCA"
author: "WM"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  results = "hide",
  warning = FALSE,
  message = FALSE,
  fig.align = "center"
)
knitr::opts_knit$set(root.dir = normalizePath("../.."))
```

## Aim

This notebook focuses on modelled avian community structure through time. It
contains two linked analyses:

- modelled bioregion shifts: define Atlas 1 reference community types from HMSC
  predicted species probability vectors, then assign Atlas 2 and Atlas 3 grid
  cells to those same reference types
- species environmental-response PCA: ordinate species by their HMSC
  environmental beta parameters, using Atlas 1 as the reference response space

Both analyses use HMSC modelled outputs only.

## Load Model Outputs

```{r load-models}
if (!require("pacman")) install.packages("pacman")
pacman::p_load(
  tidyverse, Hmsc, sf, vegan, patchwork, RColorBrewer, cowplot, cluster
)

source(file.path("support_scripts", "figure_data_helpers.R"))

base_dir <- "./HmscOutputs"
pattern <- "2026-03-13"

matching_folders <- figure_model_folders(pattern = pattern, base_dir = base_dir)
models_nums <- atlas_numbers(matching_folders)

mods <- load_hmsc_posteriors(matching_folders, base_dir = base_dir)
designs <- load_hmsc_study_designs(mods)
predsY <- load_or_compute_site_predictions(mods, matching_folders, base_dir = base_dir)

names(mods) <- names(designs) <- names(predsY) <- models_nums

year_lookup <- c("1" = "1970s", "2" = "1990s", "3" = "2010s")
reference_atlas <- "1"
```

```{r shared-definitions}
predictor_names <- function(df) {
  c(
    "tmean_breeding",
    "prec_breeding",
    "hh",
    grep("^perc_", colnames(df), value = TRUE)
  )
}

var_labels <- c(
  "tmean_breeding" = "Temperature",
  "prec_breeding" = "Precipitation",
  "hh" = "Land-use heterogeneity",
  "perc_urban" = "Urban (% coverage)",
  "perc_cropland" = "Cropland (% coverage)",
  "perc_pasture" = "Pasture (% coverage)",
  "perc_forest" = "Forest (% coverage)",
  "perc_grass_shrub" = "Grass/Shrubland (% coverage)"
)

driver_colours <- c(
  "Temperature" = "firebrick3",
  "Precipitation" = "dodgerblue3",
  "Land-use heterogeneity" = "orchid3",
  "Urban (% coverage)" = "#4d4d4d",
  "Cropland (% coverage)" = "goldenrod1",
  "Pasture (% coverage)" = "darkorange",
  "Forest (% coverage)" = "springgreen4",
  "Grass/Shrubland (% coverage)" = "springgreen2",
  "No significant driver" = "#d9d9d9"
)

mainland_bbox <- st_bbox(
  c(xmin = 400000, xmax = 750000, ymin = 6000000, ymax = 6450000),
  crs = st_crs(25832)
)
bornholm_bbox <- st_bbox(
  c(xmin = 855000, xmax = 905000, ymin = 6100000, ymax = 6160000),
  crs = st_crs(25832)
)

mainland_width <- as.numeric(mainland_bbox["xmax"] - mainland_bbox["xmin"])
mainland_height <- as.numeric(mainland_bbox["ymax"] - mainland_bbox["ymin"])
bornholm_width <- as.numeric(bornholm_bbox["xmax"] - bornholm_bbox["xmin"])
bornholm_height <- as.numeric(bornholm_bbox["ymax"] - bornholm_bbox["ymin"])

inset_w <- bornholm_width / mainland_width
inset_h <- bornholm_height / mainland_height
```

## Modelled Avian Bioregion Shifts

This analysis treats Atlas 1 as a reference community map. It first defines
discrete clusters from Atlas 1 modelled species probability vectors, then
assigns Atlas 2 and Atlas 3 communities to those same reference community
types. A retained cell is compositionally closest to the same Atlas 1 cluster;
a shifted cell is closer to another Atlas 1 cluster.

```{r bioregion-functions}
assert_species_alignment <- function(predictions) {
  reference_species <- colnames(predictions[[1]])
  mismatched <- names(predictions)[map_lgl(predictions, ~ !identical(colnames(.x), reference_species))]

  if (length(mismatched) > 0) {
    stop("Species columns do not align for atlas: ", paste(mismatched, collapse = ", "), call. = FALSE)
  }

  tibble(
    atlas = names(predictions),
    n_sites = map_int(predictions, nrow),
    n_species = map_int(predictions, ncol)
  )
}

hellinger_transform <- function(x) {
  row_total <- rowSums(x, na.rm = TRUE)
  rel <- sweep(x, 1, row_total, "/")
  rel[!is.finite(as.matrix(rel))] <- 0
  sqrt(rel)
}

bioregion_input_frame <- function(pred_y, design, atlas) {
  design_aligned <- design[match(rownames(pred_y), design$survey), ]

  tibble(
    atlas = atlas,
    period = year_lookup[[as.character(atlas)]],
    survey = rownames(pred_y)
  ) |>
    bind_cols(design_aligned |> select(site, X, Y))
}

choose_reference_bioregions <- function(reference_hellinger, k_values = 3:8) {
  reference_dist <- vegan::vegdist(reference_hellinger, method = "bray")

  k_diagnostics <- map_dfr(k_values, function(k) {
    fit <- cluster::pam(reference_dist, k = k, diss = TRUE)
    cluster_sizes <- table(fit$clustering)

    tibble(
      k = k,
      mean_silhouette = fit$silinfo$avg.width,
      min_cluster_n = min(cluster_sizes),
      min_cluster_prop = min(cluster_sizes) / length(fit$clustering)
    )
  })

  chosen_k <- k_diagnostics |>
    filter(min_cluster_prop >= 0.03) |>
    slice_max(mean_silhouette, n = 1, with_ties = FALSE) |>
    pull(k)

  if (length(chosen_k) == 0) {
    chosen_k <- k_diagnostics |>
      slice_max(mean_silhouette, n = 1, with_ties = FALSE) |>
      pull(k)
  }

  chosen_fit <- cluster::pam(reference_dist, k = chosen_k, diss = TRUE)

  list(
    reference_dist = reference_dist,
    k_diagnostics = k_diagnostics,
    chosen_k = chosen_k,
    fit = chosen_fit
  )
}

bray_to_medoids <- function(query_matrix, medoid_matrix) {
  map_dfc(seq_len(nrow(medoid_matrix)), function(i) {
    medoid <- medoid_matrix[i, ]
    numerator <- rowSums(abs(sweep(query_matrix, 2, medoid, "-")), na.rm = TRUE)
    denominator <- rowSums(sweep(query_matrix, 2, medoid, "+"), na.rm = TRUE)
    tibble(!!paste0("Cluster ", i) := ifelse(denominator == 0, 0, numerator / denominator))
  }) |>
    as_tibble()
}

assign_to_reference_bioregions <- function(query_hellinger, site_frame, medoid_matrix) {
  distance_tbl <- bray_to_medoids(query_hellinger, medoid_matrix)
  distance_matrix <- as.matrix(distance_tbl)
  nearest_idx <- max.col(-distance_matrix, ties.method = "first")
  sorted_distances <- t(apply(distance_matrix, 1, sort))

  site_frame |>
    mutate(
      bioregion = factor(paste0("Cluster ", nearest_idx), levels = paste0("Cluster ", seq_len(nrow(medoid_matrix)))),
      nearest_distance = distance_matrix[cbind(seq_len(nrow(distance_matrix)), nearest_idx)],
      second_distance = sorted_distances[, 2],
      assignment_margin = second_distance - nearest_distance,
      low_confidence = assignment_margin < quantile(assignment_margin, 0.1, na.rm = TRUE)
    )
}

summarise_bioregion_transitions <- function(assignments_wide, target_col, target_period) {
  assignments_wide |>
    filter(!is.na(reference_bioregion), !is.na(.data[[target_col]])) |>
    group_by(reference_bioregion) |>
    summarise(
      target_period = target_period,
      n_reference_sites = n(),
      n_retained = sum(.data[[target_col]] == reference_bioregion),
      pct_retained = n_retained / n_reference_sites,
      n_reassigned = n_reference_sites - n_retained,
      pct_reassigned = n_reassigned / n_reference_sites,
      dominant_destination = names(sort(table(.data[[target_col]][.data[[target_col]] != reference_bioregion]), decreasing = TRUE))[1],
      dominant_destination = ifelse(is.na(dominant_destination), "None", dominant_destination),
      mean_assignment_margin = mean(.data[[paste0("assignment_margin_", target_period)]], na.rm = TRUE),
      .groups = "drop"
    )
}

bioregion_theme <- function(legend_position = "bottom") {
  theme_minimal(base_size = 10) +
    theme(
      legend.position = legend_position,
      legend.title = element_text(face = "bold"),
      axis.text = element_blank(),
      axis.title = element_blank(),
      panel.grid = element_blank(),
      plot.title = element_text(face = "bold", hjust = 0.5),
      plot.background = element_rect(fill = "white", colour = NA),
      plot.margin = margin(0, 0, 0, 0)
    )
}

plot_bioregion_base <- function(df, fill_col, palette, title = NULL,
                                bbox = mainland_bbox, show_legend = FALSE,
                                border = FALSE) {
  ggplot(df) +
    geom_point(aes(x = X, y = Y, colour = .data[[fill_col]]), size = 1.25, alpha = 0.95) +
    scale_colour_manual(values = palette, drop = FALSE, name = "Cluster") +
    coord_fixed(
      xlim = c(bbox["xmin"], bbox["xmax"]),
      ylim = c(bbox["ymin"], bbox["ymax"]),
      expand = FALSE
    ) +
    bioregion_theme(if (show_legend) "bottom" else "none") +
    theme(panel.border = if (border) element_rect(colour = "grey35", fill = NA, linewidth = 0.45) else element_blank()) +
    labs(title = title)
}

plot_bioregion_map <- function(df, fill_col, palette, title) {
  p_main <- plot_bioregion_base(df, fill_col, palette, title = title, bbox = mainland_bbox)
  p_inset <- plot_bioregion_base(df, fill_col, palette, bbox = bornholm_bbox, border = TRUE) +
    theme_void() +
    theme(
      legend.position = "none",
      plot.background = element_rect(fill = "white", colour = NA),
      panel.border = element_rect(colour = "grey35", fill = NA, linewidth = 0.7)
    )

  ggdraw(p_main) +
    draw_plot(
      p_inset,
      x = 0.66,
      y = 0.60,
      width = 0.24,
      height = 0.22
    ) +
    draw_label("Bornholm", x = 0.78, y = 0.84, size = 7, colour = "grey30") |>
    as_grob()
}

bioregion_legend <- function(df, fill_col, palette) {
  plot_bioregion_base(df, fill_col, palette, show_legend = TRUE) |>
    get_legend()
}

adjusted_rand_index <- function(x, y) {
  contingency <- table(x, y)
  choose2 <- function(n) n * (n - 1) / 2
  index <- sum(choose2(contingency))
  expected_index <- sum(choose2(rowSums(contingency))) * sum(choose2(colSums(contingency))) / choose2(sum(contingency))
  max_index <- (sum(choose2(rowSums(contingency))) + sum(choose2(colSums(contingency)))) / 2

  if (isTRUE(all.equal(max_index, expected_index))) {
    return(NA_real_)
  }

  (index - expected_index) / (max_index - expected_index)
}

cluster_pca_point_cloud <- function(score_df, axes, k_values = 2:8) {
  score_matrix <- score_df |>
    select(all_of(axes)) |>
    as.matrix()
  score_dist <- dist(score_matrix)

  diagnostics <- map_dfr(k_values, function(k) {
    fit <- cluster::pam(score_dist, k = k, diss = TRUE)
    tibble(
      k = k,
      mean_silhouette = fit$silinfo$avg.width,
      min_cluster_n = min(table(fit$clustering)),
      min_cluster_prop = min(table(fit$clustering)) / length(fit$clustering)
    )
  })

  chosen_k <- diagnostics |>
    filter(min_cluster_prop >= 0.03) |>
    slice_max(mean_silhouette, n = 1, with_ties = FALSE) |>
    pull(k)

  if (length(chosen_k) == 0) {
    chosen_k <- diagnostics |>
      slice_max(mean_silhouette, n = 1, with_ties = FALSE) |>
      pull(k)
  }

  fit <- cluster::pam(score_dist, k = chosen_k, diss = TRUE)

  list(
    diagnostics = diagnostics,
    chosen_k = chosen_k,
    assignments = score_df |>
      mutate(
        pca_cluster = factor(paste0("PCA cluster ", fit$clustering)),
        pca_cluster_silhouette = fit$silinfo$widths[, "sil_width"]
      )
  )
}

plot_cluster_occupancy <- function(df, palette) {
  df |>
    arrange(bioregion) |>
    mutate(
      ymin = lag(cumsum(pct_sites), default = 0),
      ymax = cumsum(pct_sites),
      ymid = (ymin + ymax) / 2,
      label = scales::percent(pct_sites, accuracy = 1)
    ) |>
    ggplot(aes(x = 1, y = pct_sites, fill = bioregion)) +
    geom_col(width = 0.55, colour = "white", linewidth = 0.35) +
    geom_text(
      aes(y = ymid, label = label),
      size = 2.7,
      colour = "grey15",
      fontface = "bold"
    ) +
    scale_fill_manual(values = palette, drop = FALSE, name = "Cluster") +
    scale_y_continuous(
      breaks = NULL,
      labels = NULL,
      limits = c(0, 1),
      expand = expansion(mult = c(0, 0))
    ) +
    labs(x = NULL, y = "% grid cells") +
    coord_cartesian(clip = "off") +
    theme_minimal(base_size = 9) +
    theme(
      legend.position = "none",
      panel.grid = element_blank(),
      axis.text = element_blank(),
      axis.ticks = element_blank(),
      axis.title = element_blank(),
      plot.margin = margin(4, 2, 4, 0)
    )
}

plot_bioregion_map_with_bar <- function(map_df, occupancy_df, palette, title) {
  map_plot <- plot_bioregion_map(map_df, "bioregion", palette, title)
  bar_plot <- plot_cluster_occupancy(occupancy_df, palette)

  plot_grid(map_plot, bar_plot, nrow = 1, rel_widths = c(1, 0.18), align = "h")
}
```

```{r bioregion-analysis, cache=TRUE}
species_alignment <- assert_species_alignment(predsY)

hellinger_communities <- map(predsY, hellinger_transform)
community_frames <- imap(predsY, ~ bioregion_input_frame(.x, designs[[.y]], .y))

reference_hellinger <- hellinger_communities[[reference_atlas]]
reference_pca <- prcomp(reference_hellinger, center = TRUE, scale. = FALSE)
reference_pca_variance <- tibble(
  axis = paste0("PC", seq_along(reference_pca$sdev)),
  variance_explained = reference_pca$sdev^2 / sum(reference_pca$sdev^2),
  cumulative_variance = cumsum(variance_explained)
)
reference_selection <- choose_reference_bioregions(reference_hellinger, k_values = 3:8)

reference_clusters <- reference_selection$fit$clustering
medoid_indices <- reference_selection$fit$id.med
medoid_matrix <- reference_hellinger[medoid_indices, , drop = FALSE]
rownames(medoid_matrix) <- paste0("Cluster ", seq_len(nrow(medoid_matrix)))

bioregion_palette <- setNames(
  brewer.pal(max(3, reference_selection$chosen_k), "Set2")[seq_len(reference_selection$chosen_k)],
  paste0("Cluster ", seq_len(reference_selection$chosen_k))
)

bioregion_assignments <- imap(hellinger_communities, function(hellinger, atlas) {
  assigned <- assign_to_reference_bioregions(
    hellinger,
    community_frames[[atlas]],
    medoid_matrix
  )

  if (atlas == reference_atlas) {
    assigned <- assigned |>
      mutate(
        reference_cluster = factor(
          paste0("Cluster ", reference_clusters),
          levels = names(bioregion_palette)
        ),
        validation_match = reference_cluster == bioregion
      )
  }

  assigned
})

bioregion_assignments_df <- bind_rows(bioregion_assignments)

atlas1_reference <- bioregion_assignments[[reference_atlas]] |>
  select(site, reference_bioregion = reference_cluster)

assignments_wide <- reduce(
  imap(bioregion_assignments, function(df, atlas) {
    period <- year_lookup[[as.character(atlas)]]
    df |>
      select(site, bioregion, assignment_margin, low_confidence) |>
      rename(
        !!paste0("bioregion_", period) := bioregion,
        !!paste0("assignment_margin_", period) := assignment_margin,
        !!paste0("low_confidence_", period) := low_confidence
      )
  }),
  full_join,
  by = "site"
) |>
  left_join(atlas1_reference, by = "site")

site_match_summary <- tibble(
  comparison = c("Atlas 1 to Atlas 2", "Atlas 1 to Atlas 3", "Atlas 2 to Atlas 3"),
  n_matched_sites = c(
    sum(!is.na(assignments_wide$bioregion_1970s) & !is.na(assignments_wide$bioregion_1990s)),
    sum(!is.na(assignments_wide$bioregion_1970s) & !is.na(assignments_wide$bioregion_2010s)),
    sum(!is.na(assignments_wide$bioregion_1990s) & !is.na(assignments_wide$bioregion_2010s))
  ),
  n_unmatched_sites = c(
    sum(xor(!is.na(assignments_wide$bioregion_1970s), !is.na(assignments_wide$bioregion_1990s))),
    sum(xor(!is.na(assignments_wide$bioregion_1970s), !is.na(assignments_wide$bioregion_2010s))),
    sum(xor(!is.na(assignments_wide$bioregion_1990s), !is.na(assignments_wide$bioregion_2010s)))
  )
)

transition_1_to_2 <- assignments_wide |>
  filter(!is.na(reference_bioregion), !is.na(bioregion_1990s)) |>
  count(reference_bioregion, assigned_bioregion = bioregion_1990s, name = "n") |>
  group_by(reference_bioregion) |>
  mutate(prop = n / sum(n), target_period = "1990s") |>
  ungroup()

transition_1_to_3 <- assignments_wide |>
  filter(!is.na(reference_bioregion), !is.na(bioregion_2010s)) |>
  count(reference_bioregion, assigned_bioregion = bioregion_2010s, name = "n") |>
  group_by(reference_bioregion) |>
  mutate(prop = n / sum(n), target_period = "2010s") |>
  ungroup()

transition_tables <- bind_rows(transition_1_to_2, transition_1_to_3)

bioregion_summary <- bind_rows(
  summarise_bioregion_transitions(assignments_wide, "bioregion_1990s", "1990s"),
  summarise_bioregion_transitions(assignments_wide, "bioregion_2010s", "2010s")
)

bioregion_shift_maps <- bioregion_assignments_df |>
  left_join(atlas1_reference, by = "site") |>
  mutate(
    shift_status = case_when(
      period == "1970s" ~ "Reference",
      is.na(reference_bioregion) ~ "No Atlas 1 match",
      bioregion == reference_bioregion ~ "Retained reference region",
      low_confidence ~ "Shifted, low confidence",
      TRUE ~ "Shifted reference region"
    ),
    shift_status = factor(
      shift_status,
      levels = c("Reference", "Retained reference region", "Shifted reference region", "Shifted, low confidence", "No Atlas 1 match")
    )
  )

n_pca_axes_for_clustering <- min(which(reference_pca_variance$cumulative_variance >= 0.8)[1], 10)
pca_axes_for_clustering <- reference_pca_variance$axis[seq_len(n_pca_axes_for_clustering)]

pca_scores_full <- imap_dfr(hellinger_communities, function(hellinger, atlas) {
  scores <- predict(reference_pca, newdata = hellinger)[, pca_axes_for_clustering, drop = FALSE]
  community_frames[[atlas]] |>
    bind_cols(as_tibble(scores)) |>
    left_join(bioregion_assignments[[atlas]] |> select(survey, bioregion), by = "survey")
})

pca_scores <- pca_scores_full |>
  select(atlas, period, survey, site, X, Y, PC1, PC2, bioregion)

pca_cloud_cluster_results <- cluster_pca_point_cloud(
  pca_scores_full,
  axes = pca_axes_for_clustering,
  k_values = 2:8
)

pca_cloud_cluster_diagnostics <- pca_cloud_cluster_results$diagnostics |>
  mutate(scope = "All atlas periods")

pca_cloud_clusters <- pca_cloud_cluster_results$assignments

pca_cluster_period_diagnostics <- pca_scores_full |>
  group_split(period) |>
  map_dfr(function(df) {
    res <- cluster_pca_point_cloud(df, axes = pca_axes_for_clustering, k_values = 2:8)
    res$diagnostics |>
      mutate(scope = unique(df$period))
  })

pca_cluster_agreement <- pca_cloud_clusters |>
  group_by(period) |>
  summarise(
    n_sites = n(),
    adjusted_rand_index = adjusted_rand_index(bioregion, pca_cluster),
    mean_silhouette = mean(pca_cluster_silhouette, na.rm = TRUE),
    .groups = "drop"
  ) |>
  bind_rows(
    pca_cloud_clusters |>
      summarise(
        period = "All atlas periods",
        n_sites = n(),
        adjusted_rand_index = adjusted_rand_index(bioregion, pca_cluster),
        mean_silhouette = mean(pca_cluster_silhouette, na.rm = TRUE)
      )
  )

pca_centroids <- pca_scores |>
  group_by(period, bioregion) |>
  summarise(
    PC1 = mean(PC1, na.rm = TRUE),
    PC2 = mean(PC2, na.rm = TRUE),
    n_sites = n(),
    .groups = "drop"
  )

cluster_occupancy <- bioregion_assignments_df |>
  count(period, bioregion, name = "n_sites") |>
  complete(
    period,
    bioregion = factor(names(bioregion_palette), levels = names(bioregion_palette)),
    fill = list(n_sites = 0)
  ) |>
  group_by(period) |>
  mutate(
    pct_sites = n_sites / sum(n_sites),
    bioregion = factor(bioregion, levels = names(bioregion_palette))
  ) |>
  ungroup()

validation_summary <- bioregion_assignments[[reference_atlas]] |>
  summarise(
    n_reference_sites = n(),
    n_validation_matches = sum(validation_match, na.rm = TRUE),
    validation_match_rate = mean(validation_match, na.rm = TRUE)
  )
```

```{r bioregion-diagnostics, results='asis'}
knitr::kable(species_alignment, caption = "Species and site dimensions used for modelled bioregion analysis.")
knitr::kable(reference_selection$k_diagnostics, digits = 3, caption = "Atlas 1 PAM/k-medoids diagnostics across candidate numbers of bioregions.")
knitr::kable(site_match_summary, caption = "Matched and unmatched grid cells across atlas periods.")
knitr::kable(validation_summary, digits = 3, caption = "Validation that Atlas 1 nearest-medoid assignment reproduces the Atlas 1 reference clustering.")
knitr::kable(bioregion_summary, digits = 3, caption = "Reference cluster retention and reassignment summaries.")
```

## Manuscript Methods

We used HMSC modelled species occurrence probabilities to evaluate whether
avian community structure shifted among atlas periods. For each atlas period,
we extracted the predicted species probability vector for every grid cell and
retained only species shared across all modelled outputs. Predicted community
vectors were transformed using the Hellinger transformation by first dividing
each grid cell's predicted probabilities by its total predicted probability
across species and then taking the square root. This transformation treats the
data as community composition while reducing the influence of very common
species.

Atlas 1 was used as the reference period for defining community clusters. We
calculated Bray-Curtis dissimilarities among Hellinger-transformed Atlas 1 grid
cells and clustered them using partitioning around medoids. Candidate solutions
from three to eight clusters were compared using mean silhouette width, while
excluding solutions with very small clusters (<3% of Atlas 1 grid cells). The
selected Atlas 1 medoids were then kept fixed as reference community types. For
Atlas 2 and Atlas 3, each grid cell was assigned to the nearest Atlas 1 medoid
using Bray-Curtis dissimilarity. Assignment confidence was summarized as the
difference between the nearest and second-nearest medoid distances, with the
lowest 10% of margins flagged as low-confidence assignments.

To visualize directional changes in community composition, we fitted a PCA to
the Hellinger-transformed Atlas 1 community matrix and projected Atlas 2 and
Atlas 3 communities into the same ordination space. Cluster centroids were
calculated separately for each atlas period in this reference PCA space. We
summarized temporal change using maps of projected cluster membership, the
percentage of grid cells assigned to each cluster, transition tables from the
Atlas 1 reference clusters to later atlas assignments, and centroid movement in
ordination space. As a sensitivity check on whether the ordination itself
contained discrete structure, we also clustered the projected PCA point cloud
directly using partitioning around medoids. Candidate solutions from two to
eight PCA-space clusters were compared using mean silhouette width, and
agreement between these unsupervised PCA clusters and the fixed reference
clusters was summarized using the adjusted Rand index.

Species-level environmental-response structure was assessed separately from
the grid-cell community clustering. For each atlas period, we extracted HMSC
posterior mean beta parameters for the environmental predictors and multiplied
each coefficient by the within-atlas standard deviation of the corresponding
predictor. This converted coefficients to standardized effects, representing
the expected response to a one-standard-deviation change in the environmental
gradient available during that atlas period. Standardized effects were weighted
by posterior certainty, defined as `2 * abs(Pr(beta > 0) - 0.5)`, so uncertain
effects were shrunk towards zero. PCA was fitted to the weighted standardized
beta matrix from Atlas 1, and Atlas 2 and Atlas 3 species response profiles
were projected into that fixed reference space.

### Plain-Language Bioregion Workflow

The bioregion analysis asks whether Danish bird communities can be summarized
as repeatable composition types, and whether grid cells have moved between
those types through time. The input is the HMSC modelled species probability
matrix for each atlas period. Rows are grid cells, columns are bird species,
and each value is the predicted probability of occurrence. Thus, each grid cell
is represented by a complete modelled community vector rather than by richness
or by a single dominant species.

Before clustering, each grid cell's predicted community vector was
Hellinger-transformed. This means the predicted probabilities were converted to
relative composition within the grid cell and then square-root transformed.
This transformation keeps the focus on compositional differences among
communities and reduces the tendency for high-richness cells or very common
species to dominate the analysis.

Atlas 1 was treated as the baseline community map. Within Atlas 1, we calculated
Bray-Curtis dissimilarity among all Hellinger-transformed communities and then
used PAM/k-medoids clustering to identify discrete reference clusters. PAM was
used because each cluster centre is a real grid-cell community, called a
medoid, rather than an abstract average. Candidate values of `k` from 3 to 8
were compared with silhouette width, and the selected solution contained three
clusters. These are labelled `Cluster 1`, `Cluster 2`, and `Cluster 3`.

Atlas 2 and Atlas 3 were not clustered independently. Instead, the Atlas 1
cluster medoids were kept fixed as reference community types. Each later grid
cell was assigned to the nearest Atlas 1 medoid using Bray-Curtis dissimilarity.
This means a later grid cell assigned to `Cluster 2` is interpreted as being
more similar to the Atlas 1 `Cluster 2` community type than to the other Atlas 1
reference types. A change in cluster identity therefore means a movement
towards another baseline community type, not the discovery of a new later-period
cluster.

Assignment confidence was summarized using the distance margin between the
nearest and second-nearest Atlas 1 medoids. If those two distances were similar,
the grid cell was flagged as low confidence. These cells sit close to the
boundary between community types, so apparent shifts in those places should be
interpreted cautiously.

The ordination plots provide a visual summary of the same modelled community
composition data. PCA was fitted to the Hellinger-transformed Atlas 1
communities, and Atlas 2 and Atlas 3 were projected into that fixed reference
PCA space. The PCA does not define the clusters; it helps show whether the
clusters occupy distinct regions of composition space and whether their
centroids drift directionally over time.

The linked trait-composition analysis then asks what kinds of species make up
each cluster. For each cluster and atlas period, predicted probabilities were
summed for species in each trait group and divided by total expected richness
among species with known trait data. Those percentages are therefore
model-weighted expected community composition, not raw species counts.

## Cluster Maps And Ordination

```{r bioregion-map-panel, fig.width=13.5, fig.height=7.5}
bioregion_period_maps <- imap(bioregion_assignments, function(df, atlas) {
  period <- year_lookup[[as.character(atlas)]]
  plot_bioregion_map_with_bar(
    df,
    cluster_occupancy |> filter(.data$period == .env$period),
    bioregion_palette,
    period
  )
})

bioregion_map_panel <- plot_grid(
  plotlist = bioregion_period_maps,
  nrow = 1,
  align = "h",
  rel_widths = c(1, 1, 1)
)

bioregion_map_panel
```

```{r bioregion-shift-map, fig.width=12, fig.height=4.8}
shift_palette <- c(
  "Reference" = "grey75",
  "Retained reference region" = "#1a9850",
  "Shifted reference region" = "#d73027",
  "Shifted, low confidence" = "#fdae61",
  "No Atlas 1 match" = "grey30"
)

shift_period_maps <- bioregion_shift_maps |>
  filter(period != "1970s") |>
  (\(df) split(df, df$period))() |>
  imap(function(df, period) {
    plot_bioregion_map(df, "shift_status", shift_palette, period)
  })

shift_map_panel <- plot_grid(
  plot_grid(plotlist = shift_period_maps, nrow = 1, align = "none"),
  bioregion_legend(bioregion_shift_maps |> filter(period != "1970s"), "shift_status", shift_palette),
  ncol = 1,
  rel_heights = c(1, 0.18)
)

shift_map_panel
```

```{r bioregion-ordination, fig.width=8.5, fig.height=6.4}
bioregion_ordination_plot <- ggplot(pca_scores, aes(x = PC1, y = PC2, colour = bioregion)) +
  geom_point(alpha = 0.07, size = 0.75) +
  geom_path(
    data = pca_centroids,
    aes(group = bioregion),
    linewidth = 0.8,
    arrow = arrow(length = unit(0.15, "cm")),
    alpha = 0.8
  ) +
  geom_point(data = pca_centroids, aes(size = n_sites), alpha = 0.95) +
  geom_text(data = pca_centroids, aes(label = period), size = 3, nudge_y = 0.015, show.legend = FALSE) +
  scale_colour_manual(values = bioregion_palette, drop = FALSE, name = "Cluster") +
  scale_size_continuous(name = "Sites", range = c(2, 5)) +
  labs(
    title = "Projected movement of modelled avian clusters",
    subtitle = "Atlas 2 and 3 communities projected into the reference Hellinger-PCA space",
    x = "PCA axis 1",
    y = "PCA axis 2"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")

bioregion_ordination_plot
```

```{r bioregion-publication-figure, fig.width=13.5, fig.height=11}
cluster_legend <- bioregion_legend(
  bioregion_assignments[[reference_atlas]],
  "bioregion",
  bioregion_palette
)

bioregion_publication_figure <- plot_grid(
  bioregion_map_panel,
  bioregion_ordination_plot +
    labs(title = NULL, subtitle = NULL) +
    theme(legend.position = "none"),
  cluster_legend,
  ncol = 1,
  rel_heights = c(1, 1.02, 0.1)
)

bioregion_publication_figure
```

Low-confidence shifts are cells where the nearest and second-nearest reference
bioregions are very similar, so those changes should be treated cautiously.

## PCA Point-Cloud Separation Diagnostics

The fixed clusters above are useful because they keep Atlas 1 as the baseline.
This diagnostic asks a slightly different question: if the projected community
points are clustered directly in PCA space, do they form clearly separated
groups?

```{r pca-cloud-cluster-diagnostics, results='asis'}
knitr::kable(
  bind_rows(pca_cloud_cluster_diagnostics, pca_cluster_period_diagnostics) |>
    arrange(scope, k),
  digits = 3,
  caption = paste(
    "PAM/k-medoids diagnostics for projected community points using",
    length(pca_axes_for_clustering),
    "PCA axes."
  )
)

knitr::kable(
  pca_cluster_agreement,
  digits = 3,
  caption = "Agreement between unsupervised PCA-space clusters and fixed reference clusters."
)
```

```{r pca-cloud-cluster-plot, fig.width=11, fig.height=5.8}
pca_cluster_palette <- setNames(
  brewer.pal(max(3, nlevels(pca_cloud_clusters$pca_cluster)), "Dark2")[seq_len(nlevels(pca_cloud_clusters$pca_cluster))],
  levels(pca_cloud_clusters$pca_cluster)
)

pca_cloud_cluster_plot <- ggplot(pca_cloud_clusters, aes(x = PC1, y = PC2, colour = pca_cluster)) +
  geom_point(alpha = 0.18, size = 0.8) +
  stat_ellipse(linewidth = 0.55, alpha = 0.8, show.legend = FALSE) +
  facet_wrap(~ period, nrow = 1) +
  scale_colour_manual(values = pca_cluster_palette, name = "PCA cluster") +
  labs(
    title = "Unsupervised clustering of projected community points",
    subtitle = "PAM/k-medoids clustering in the reference PCA space",
    x = "PCA axis 1",
    y = "PCA axis 2"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    legend.position = "bottom",
    strip.text = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )

pca_silhouette_plot <- bind_rows(pca_cloud_cluster_diagnostics, pca_cluster_period_diagnostics) |>
  mutate(scope = factor(scope, levels = c("All atlas periods", unname(year_lookup)))) |>
  ggplot(aes(x = k, y = mean_silhouette, colour = scope)) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 1.8) +
  scale_x_continuous(breaks = 2:8) +
  labs(
    title = "Separation strength",
    x = "Number of PCA clusters",
    y = "Mean silhouette width",
    colour = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")

pca_cloud_separation_panel <- plot_grid(
  pca_cloud_cluster_plot,
  pca_silhouette_plot,
  nrow = 1,
  rel_widths = c(1.35, 1)
)

pca_cloud_separation_panel
```

Interpretation: high silhouette values would indicate discrete, well-separated
groups in the ordination. Low or moderate values indicate that the modelled
communities are better understood as a gradient, even if it is still useful to
summarize that gradient with fixed reference clusters.

## Species Environmental Response PCA

This section asks whether the way species respond to the environmental
gradients shifted through time. Each species is represented by its HMSC
environmental beta coefficients. The coefficients are converted to standardized
effects before PCA:

`standardized beta = posterior mean beta * within-atlas predictor SD`

The PCA therefore ordains species by their response to a one-standard-deviation
change in each predictor within that atlas period. Coefficients are also
weighted by posterior certainty, so uncertain effects are shrunk towards zero
rather than removed with a hard threshold.

```{r beta-pca-functions}
read_beta_parameter_workbook <- function(model_folder, base_dir = "HmscOutputs") {
  path <- file.path(
    base_dir,
    model_folder,
    "Results",
    paste0(model_folder, "parameter_estimates_Beta_.xlsx")
  )

  posterior_mean <- readxl::read_excel(path, sheet = "Posterior mean") |>
    pivot_longer(
      cols = -Species,
      names_to = "variable",
      values_to = "beta_mean"
    )

  posterior_positive <- readxl::read_excel(path, sheet = "Pr(x>0)") |>
    pivot_longer(
      cols = -Species,
      names_to = "variable",
      values_to = "prob_positive"
    )

  posterior_mean |>
    left_join(posterior_positive, by = c("Species", "variable")) |>
    mutate(
      atlas = as.character(sub(".*Atlas([0-9]+).*", "\\1", model_folder)),
      period = year_lookup[atlas],
      certainty_weight = 2 * abs(prob_positive - 0.5)
    )
}

predictor_sds_by_atlas <- function(models, variables) {
  imap_dfr(models, function(model, atlas) {
    tibble(
      atlas = as.character(atlas),
      variable = variables,
      predictor_sd = map_dbl(variables, ~ sd(model$XData[[.x]], na.rm = TRUE))
    )
  }) |>
    mutate(period = year_lookup[atlas])
}

beta_parameter_alignment <- function(beta_long, variables) {
  atlas_species <- beta_long |>
    distinct(atlas, Species) |>
    count(atlas, name = "n_species")

  common_species <- beta_long |>
    distinct(atlas, Species) |>
    count(Species, name = "n_atlases") |>
    filter(n_atlases == length(unique(beta_long$atlas))) |>
    pull(Species) |>
    sort()

  tibble(
    atlas = names(year_lookup),
    period = unname(year_lookup),
    n_species_in_workbook = atlas_species$n_species[match(names(year_lookup), atlas_species$atlas)],
    n_common_species_used = length(common_species),
    n_predictors_used = length(variables)
  )
}

weighted_beta_matrix <- function(beta_df, atlas_id, species, variables) {
  beta_df |>
    filter(atlas == atlas_id, Species %in% species, variable %in% variables) |>
    mutate(variable = factor(variable, levels = variables)) |>
    arrange(Species, variable) |>
    select(Species, variable, weighted_standardized_beta) |>
    pivot_wider(names_from = variable, values_from = weighted_standardized_beta) |>
    arrange(Species) |>
    column_to_rownames(var = "Species") |>
    as.matrix()
}

largest_beta_pca_movements <- function(scores, target_period, n = 15) {
  scores |>
    filter(period %in% c("1970s", target_period)) |>
    select(Species, period, PC1, PC2) |>
    pivot_wider(
      names_from = period,
      values_from = c(PC1, PC2),
      names_sep = "_"
    ) |>
    mutate(
      target_period = target_period,
      delta_PC1 = .data[[paste0("PC1_", target_period)]] - PC1_1970s,
      delta_PC2 = .data[[paste0("PC2_", target_period)]] - PC2_1970s,
      pca_distance = sqrt(delta_PC1^2 + delta_PC2^2)
    ) |>
    arrange(desc(pca_distance)) |>
    slice_head(n = n)
}
```

```{r beta-pca-analysis}
beta_predictors <- intersect(
  predictor_names(mods[[reference_atlas]]$XData),
  names(var_labels)
)

beta_parameter_long_raw <- map_dfr(
  matching_folders,
  read_beta_parameter_workbook,
  base_dir = base_dir
) |>
  filter(variable != "(Intercept)", variable %in% beta_predictors)

beta_predictor_sds <- predictor_sds_by_atlas(mods, beta_predictors)

stopifnot(all(is.finite(beta_predictor_sds$predictor_sd)))
stopifnot(all(beta_predictor_sds$predictor_sd > 0))

beta_parameter_long <- beta_parameter_long_raw |>
  left_join(beta_predictor_sds, by = c("atlas", "period", "variable")) |>
  mutate(
    standardized_beta = beta_mean * predictor_sd,
    weighted_standardized_beta = standardized_beta * certainty_weight,
    variable_label = recode(variable, !!!var_labels)
  )

common_beta_species <- beta_parameter_long |>
  distinct(atlas, Species) |>
  count(Species, name = "n_atlases") |>
  filter(n_atlases == length(unique(beta_parameter_long$atlas))) |>
  pull(Species) |>
  sort()

beta_alignment_summary <- beta_parameter_alignment(beta_parameter_long, beta_predictors)

beta_matrices <- setNames(
  map(names(year_lookup), ~ weighted_beta_matrix(
    beta_parameter_long,
    .x,
    common_beta_species,
    beta_predictors
  )),
  names(year_lookup)
)

stopifnot(all(map_lgl(beta_matrices, ~ identical(rownames(.x), common_beta_species))))
stopifnot(all(map_lgl(beta_matrices, ~ identical(colnames(.x), beta_predictors))))

beta_reference_pca <- prcomp(beta_matrices[[reference_atlas]], center = TRUE, scale. = FALSE)
beta_pca_variance <- tibble(
  axis = paste0("PC", seq_along(beta_reference_pca$sdev)),
  variance_explained = beta_reference_pca$sdev^2 / sum(beta_reference_pca$sdev^2),
  cumulative_variance = cumsum(beta_reference_pca$sdev^2 / sum(beta_reference_pca$sdev^2))
)

beta_pca_scores <- imap_dfr(beta_matrices, function(mat, atlas) {
  predict(beta_reference_pca, newdata = mat)[, 1:2, drop = FALSE] |>
    as_tibble(rownames = "Species") |>
    mutate(
      atlas = atlas,
      period = year_lookup[atlas],
      .before = 2
    )
})

atlas1_projection_check <- max(abs(
  beta_pca_scores |>
    filter(atlas == reference_atlas) |>
    arrange(Species) |>
    select(PC1, PC2) |>
    as.matrix() -
    beta_reference_pca$x[common_beta_species, 1:2, drop = FALSE]
))

stopifnot(atlas1_projection_check < 1e-10)

beta_pca_loadings <- beta_reference_pca$rotation[, 1:2, drop = FALSE] |>
  as_tibble(rownames = "variable") |>
  mutate(
    variable_label = recode(variable, !!!var_labels),
    loading_strength = sqrt(PC1^2 + PC2^2)
  ) |>
  arrange(desc(loading_strength))

beta_pca_centroids <- beta_pca_scores |>
  group_by(atlas, period) |>
  summarise(
    PC1 = mean(PC1, na.rm = TRUE),
    PC2 = mean(PC2, na.rm = TRUE),
    n_species = n(),
    .groups = "drop"
  ) |>
  arrange(as.numeric(atlas)) |>
  mutate(
    delta_PC1_from_1970s = PC1 - first(PC1),
    delta_PC2_from_1970s = PC2 - first(PC2),
    distance_from_1970s = sqrt(delta_PC1_from_1970s^2 + delta_PC2_from_1970s^2)
  )

beta_pca_centroid_labels <- beta_pca_centroids |>
  mutate(
    label_dx = c(-0.055, 0.045, 0.055),
    label_dy = c(0.045, -0.055, 0.045)
  )

beta_pca_movements <- bind_rows(
  largest_beta_pca_movements(beta_pca_scores, "1990s", n = 10),
  largest_beta_pca_movements(beta_pca_scores, "2010s", n = 10)
)

beta_pca_interpretation <- tibble(
  axis = c("PC1", "PC2"),
  strongest_positive_loading = map_chr(c("PC1", "PC2"), ~ beta_pca_loadings$variable_label[which.max(beta_pca_loadings[[.x]])]),
  strongest_negative_loading = map_chr(c("PC1", "PC2"), ~ beta_pca_loadings$variable_label[which.min(beta_pca_loadings[[.x]])]),
  variance_explained = beta_pca_variance$variance_explained[1:2]
)
```

```{r beta-pca-diagnostics, results='asis'}
knitr::kable(beta_alignment_summary, caption = "Species and predictor alignment for the HMSC beta-parameter PCA.")
knitr::kable(beta_pca_interpretation, digits = 3, caption = "Main environmental-response gradients in the Atlas 1 beta PCA.")
knitr::kable(beta_pca_centroids, digits = 3, caption = "Centroid movement of species environmental-response profiles in the Atlas 1 beta PCA space.")
```

```{r beta-pca-plot, fig.width=10, fig.height=7}
period_palette <- c(
  "1970s" = "#4575b4",
  "1990s" = "#fdae61",
  "2010s" = "#d73027"
)

loading_multiplier <- 0.7 * min(
  diff(range(beta_pca_scores$PC1, na.rm = TRUE)) / diff(range(beta_pca_loadings$PC1, na.rm = TRUE)),
  diff(range(beta_pca_scores$PC2, na.rm = TRUE)) / diff(range(beta_pca_loadings$PC2, na.rm = TRUE))
)

beta_loading_arrows <- beta_pca_loadings |>
  mutate(
    PC1_end = PC1 * loading_multiplier,
    PC2_end = PC2 * loading_multiplier,
    loading_label = case_when(
      variable == "hh" ~ "Heterogeneity",
      variable == "perc_grass_shrub" ~ "Grass/Shrub",
      TRUE ~ stringr::str_remove(variable_label, " \\(% coverage\\)")
    )
  )

beta_pca_main_plot <- ggplot(beta_pca_scores, aes(x = PC1, y = PC2, colour = period)) +
  geom_point(alpha = 0.32, size = 1.5) +
  geom_path(
    data = beta_pca_centroids,
    aes(group = 1),
    colour = "grey20",
    linewidth = 0.8,
    arrow = arrow(length = unit(0.16, "cm"))
  ) +
  geom_point(
    data = beta_pca_centroids,
    aes(size = n_species),
    colour = "grey10",
    fill = "white",
    shape = 21,
    stroke = 0.8,
    show.legend = FALSE
  ) +
  geom_label(
    data = beta_pca_centroid_labels,
    aes(x = PC1 + label_dx, y = PC2 + label_dy, label = period),
    colour = "grey10",
    fill = "white",
    label.size = 0.2,
    size = 3,
    show.legend = FALSE
  ) +
  scale_colour_manual(values = period_palette, drop = FALSE, name = "Atlas period") +
  labs(
    title = "Species environmental-response structure",
    subtitle = "Atlas 2 and 3 beta profiles projected into the Atlas 1 weighted standardized-beta PCA",
    x = paste0("PC1 (", scales::percent(beta_pca_variance$variance_explained[1], accuracy = 0.1), ")"),
    y = paste0("PC2 (", scales::percent(beta_pca_variance$variance_explained[2], accuracy = 0.1), ")")
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")

beta_loading_plot <- ggplot(beta_loading_arrows, aes(x = 0, y = 0, xend = PC1_end, yend = PC2_end)) +
  geom_segment(
    aes(colour = variable),
    arrow = arrow(length = unit(0.16, "cm")),
    linewidth = 0.8,
    show.legend = FALSE
  ) +
  geom_text(
    aes(
      x = PC1_end,
      y = PC2_end,
      label = loading_label,
      colour = variable,
      hjust = ifelse(PC1_end >= 0, -0.05, 1.05),
      vjust = ifelse(PC2_end >= 0, -0.25, 1.15)
    ),
    size = 2.7,
    show.legend = FALSE
  ) +
  scale_colour_manual(values = driver_colours[names(driver_colours) %in% beta_predictors]) +
  coord_cartesian(clip = "off") +
  labs(
    title = "Predictor loadings",
    x = "PC1 loading",
    y = "PC2 loading"
  ) +
  theme_minimal(base_size = 11) +
  theme(plot.margin = margin(8, 28, 8, 28))

beta_movement_plot <- beta_pca_movements |>
  mutate(
    species_label = stringr::str_replace_all(Species, "_", " "),
    species_period = paste(species_label, target_period, sep = "___"),
    species_period = forcats::fct_reorder(species_period, pca_distance)
  ) |>
  ggplot(aes(x = pca_distance, y = species_period, fill = target_period)) +
  geom_col(width = 0.7) +
  facet_wrap(~ target_period, ncol = 1, scales = "free_y") +
  scale_fill_manual(values = period_palette[c("1990s", "2010s")], name = "Target period") +
  scale_y_discrete(labels = ~ stringr::str_remove(.x, "___.*$")) +
  labs(
    title = "Largest species movements",
    x = "Distance from Atlas 1 position",
    y = NULL
  ) +
  theme_minimal(base_size = 9) +
  theme(
    legend.position = "none",
    strip.text = element_text(face = "bold"),
    panel.spacing.y = unit(0.5, "lines")
  )

beta_pca_panel <- beta_pca_main_plot / (beta_loading_plot | beta_movement_plot) +
  plot_layout(heights = c(1.18, 1), widths = c(1, 1.15))

beta_pca_panel
```

Interpretation: this ordination summarizes how species' environmental responses
are structured, not where species occur. Species close together have similar
weighted standardized responses to the modelled environmental predictors. The
centroid arrow shows whether the average species response profile moves away
from the Atlas 1 baseline. The loading panel gives the ecological meaning of
the axes: movement towards a predictor arrow means stronger positive association
with that predictor, while movement in the opposite direction means stronger
negative association. Because the coefficients were standardized with
atlas-specific predictor standard deviations, these shifts describe response
strength within the environmental variation available in each atlas period.

## Export Results

```{r export-results}
out_dir <- file.path("notebooks", "exploratory", "outputs", "bioregion-pca")
model_dir <- file.path(out_dir, "models")
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(model_dir, recursive = TRUE, showWarnings = FALSE)

write_csv(species_alignment, file.path(out_dir, paste0(pattern, "-bioregion-species-alignment.csv")))
write_csv(reference_selection$k_diagnostics, file.path(out_dir, paste0(pattern, "-bioregion-k-diagnostics.csv")))
write_csv(site_match_summary, file.path(out_dir, paste0(pattern, "-bioregion-site-match-summary.csv")))
write_csv(transition_tables, file.path(out_dir, paste0(pattern, "-bioregion-transition-tables.csv")))
write_csv(bioregion_summary, file.path(out_dir, paste0(pattern, "-bioregion-retention-summary.csv")))
write_csv(bioregion_assignments_df, file.path(out_dir, paste0(pattern, "-bioregion-assignments.csv")))
write_csv(bioregion_shift_maps, file.path(out_dir, paste0(pattern, "-bioregion-shift-status.csv")))
write_csv(pca_centroids, file.path(out_dir, paste0(pattern, "-bioregion-pca-centroids.csv")))
write_csv(cluster_occupancy, file.path(out_dir, paste0(pattern, "-cluster-occupancy.csv")))
write_csv(
  bind_rows(pca_cloud_cluster_diagnostics, pca_cluster_period_diagnostics),
  file.path(out_dir, paste0(pattern, "-pca-cloud-cluster-diagnostics.csv"))
)
write_csv(pca_cloud_clusters, file.path(out_dir, paste0(pattern, "-pca-cloud-cluster-assignments.csv")))
write_csv(pca_cluster_agreement, file.path(out_dir, paste0(pattern, "-pca-cloud-cluster-agreement.csv")))

write_csv(beta_parameter_long, file.path(out_dir, paste0(pattern, "-species-beta-weighted-standardized-effects.csv")))
write_csv(beta_pca_scores, file.path(out_dir, paste0(pattern, "-species-beta-pca-scores.csv")))
write_csv(beta_pca_loadings, file.path(out_dir, paste0(pattern, "-species-beta-pca-loadings.csv")))
write_csv(beta_pca_centroids, file.path(out_dir, paste0(pattern, "-species-beta-pca-centroids.csv")))
write_csv(beta_pca_movements, file.path(out_dir, paste0(pattern, "-species-beta-pca-largest-movements.csv")))

ggsave(
  filename = file.path(out_dir, paste0(pattern, "-bioregion-reference-assignment-maps.png")),
  plot = bioregion_map_panel,
  width = 12,
  height = 4.8,
  units = "in",
  dpi = 300,
  bg = "white"
)
ggsave(
  filename = file.path(out_dir, paste0(pattern, "-bioregion-shift-maps.png")),
  plot = shift_map_panel,
  width = 12,
  height = 4.8,
  units = "in",
  dpi = 300,
  bg = "white"
)
ggsave(
  filename = file.path(out_dir, paste0(pattern, "-bioregion-ordination-drift.png")),
  plot = bioregion_ordination_plot,
  width = 8.5,
  height = 6.4,
  units = "in",
  dpi = 300,
  bg = "white"
)
ggsave(
  filename = file.path(out_dir, paste0(pattern, "-avian-cluster-map-pca-publication-figure.png")),
  plot = bioregion_publication_figure,
  width = 13.5,
  height = 11,
  units = "in",
  dpi = 300,
  bg = "white"
)
ggsave(
  filename = file.path(out_dir, paste0(pattern, "-pca-cloud-cluster-separation.png")),
  plot = pca_cloud_separation_panel,
  width = 11,
  height = 5.8,
  units = "in",
  dpi = 300,
  bg = "white"
)
ggsave(
  filename = file.path(out_dir, paste0(pattern, "-species-beta-pca-drift.png")),
  plot = beta_pca_panel,
  width = 10,
  height = 7,
  units = "in",
  dpi = 300,
  bg = "white"
)

saveRDS(
  list(
    species_alignment = species_alignment,
    hellinger_communities = hellinger_communities,
    reference_pca = reference_pca,
    reference_selection = reference_selection,
    medoid_matrix = medoid_matrix,
    bioregion_palette = bioregion_palette,
    bioregion_assignments = bioregion_assignments,
    assignments_wide = assignments_wide,
    transition_tables = transition_tables,
    bioregion_summary = bioregion_summary,
    cluster_occupancy = cluster_occupancy,
    pca_scores = pca_scores,
    pca_centroids = pca_centroids,
    pca_axes_for_clustering = pca_axes_for_clustering,
    pca_cloud_clusters = pca_cloud_clusters,
    pca_cloud_cluster_diagnostics = pca_cloud_cluster_diagnostics,
    pca_cluster_period_diagnostics = pca_cluster_period_diagnostics,
    pca_cluster_agreement = pca_cluster_agreement,
    beta_parameter_long = beta_parameter_long,
    beta_predictor_sds = beta_predictor_sds,
    beta_matrices = beta_matrices,
    beta_reference_pca = beta_reference_pca,
    beta_pca_variance = beta_pca_variance,
    beta_pca_scores = beta_pca_scores,
    beta_pca_loadings = beta_pca_loadings,
    beta_pca_centroids = beta_pca_centroids,
    beta_pca_movements = beta_pca_movements
  ),
  file.path(model_dir, paste0(pattern, "-avian-bioregion-pca-analysis.rds"))
)
```
