---
title: "z-breeding-vs-annual-climate"
output: html_document
---


```{r setup, include=TRUE,echo=FALSE}
# Global chunk options
if(!require("pacman")) install.packages("pacman")
p_load(readxl, tidyverse, cowplot, Hmsc, cli,knitr,sf)
opts_chunk$set(
	fig.align = "center",
	fig.height = 10,
	fig.width = 10,
	inclue=T,echo=FALSE,
	warning=FALSE, message=FALSE
)

```

```{r species-environment-df}
### load environment / species data 
X <- read.csv('Data/data/1_preprocessing/X_environmental/X_Environmental.csv') 
Y <- read.csv('Data/data/1_preprocessing/Y_occurrences/Y_occurrences.csv') 
# load design and merge 
master_df <- read.csv('Data/data/1_preprocessing/design/studyDesign.csv') %>% 
  left_join(X, by='survey') %>% 
  left_join(Y, by = 'survey') %>% 
  column_to_rownames('survey') %>%
  rename(X = lon) %>% 
  rename(Y = lat)

### Main Plotting Function
species_spatial_comparison <- function(master_df, species, english_name, env_var,
                                       col_low, col_high, legend_name,
                                       relative = FALSE) { 
  
  v_env <- sym(env_var)
  
  # 1. Prep data
  subs_cols <- c('atlas', 'X', 'Y', env_var, species) 
  subs <- master_df[, colnames(master_df) %in% subs_cols]
  
  # Coordinate Transformation
  if(max(subs$Y, na.rm = TRUE) < 180){
    subs_sf <- st_as_sf(subs, coords = c("X", "Y"), crs = 4326)
    subs_sf <- st_transform(subs_sf, 23032)
    coords <- st_coordinates(subs_sf)
    
    subs <- st_drop_geometry(subs_sf) %>%
      mutate(X = coords[,1], Y = coords[,2]) %>% 
      relocate(X, Y, .after = atlas)
  }
  
  # 2. Handle Relative Transformation
  legend_title <- legend_name
  if(relative) {
    subs <- subs %>%
      group_by(atlas) %>%
      mutate(!!v_env := !!v_env - mean(!!v_env, na.rm = TRUE)) %>%
      ungroup()
    legend_title <- paste(legend_name, "(deviation from mean)")
  }
  
  # 3. Global limits
  xlims <- range(subs$X, na.rm = TRUE)
  ylims <- range(subs$Y, na.rm = TRUE)
  
  if(relative) {
    max_val <- max(abs(range(subs %>% pull(!!v_env), na.rm = TRUE)))
    env_lims <- c(-max_val, max_val)
  } else {
    env_lims <- subs %>% pull(!!v_env) %>% range(na.rm = TRUE)
  }
  
  # 4. Create Atlas Plots
  all_plots <- lapply(sort(unique(subs$atlas)), function(x) {
    subset_data <- subs[subs$atlas == x, ]
    subset_data$presence_val <- subset_data[[species]]
    
    p_top <- single_spatial_plot(subset_data, v_env, xlims, ylims, env_lims, 
                                 title = paste0("Atlas ", x, ": Presence Only"), 
                                 only_presences = TRUE,
                                 col_low, col_high)
    
    p_bottom <- single_spatial_plot(subset_data, v_env, xlims, ylims, env_lims, 
                                    title = paste0("Atlas ", x, ": Full Gradient"), 
                                    only_presences = FALSE,
                                    col_low, col_high)
    
    list(top = p_top, bottom = p_bottom)
  })
  
  # 5. Extraction and Grid Assembly
  top_row <- lapply(all_plots, `[[`, "top")
  bottom_row <- lapply(all_plots, `[[`, "bottom")
  
  final_grid <- plot_grid(
    plot_grid(plotlist = top_row, ncol = 3, align = 'h'),
    plot_grid(plotlist = bottom_row, ncol = 3, align = 'h'),
    ncol = 1, rel_heights = c(1, 1)
  )
  
  # Create Legend (using fill for consistency with top row logic)
  dummy_plot <- ggplot(subs, aes(x=X, y=Y, fill=!!v_env)) + 
    geom_point(shape = 21, color = "transparent") +
    scale_fill_gradient2(low = col_low, mid = "grey90", high = col_high, 
                         midpoint = 0, name = legend_title)
  
  if(!relative) {
    dummy_plot <- ggplot(subs, aes(x=X, y=Y, fill=!!v_env)) + 
      geom_point(shape = 21, color = "transparent") +
      scale_fill_gradient(low = col_low, high = col_high, name = legend_title)
  }

  leg <- get_legend(dummy_plot + theme(legend.position = "bottom"))
  title_theme <- ggdraw() + draw_label(sprintf('%s - %s', english_name, legend_name), 
                                       fontface = 'bold', x = 0.5, size = 18)
  
  layout <- plot_grid(title_theme, final_grid, leg, ncol = 1, rel_heights = c(0.1, 1, 0.1))  
  return(layout)
}

### Helper Plotting Function
single_spatial_plot <- function(subset_data, v_env, xlims, ylims, env_lims, title, 
                                only_presences, col_low, col_high) {
  
  is_relative <- env_lims[1] < 0 && env_lims[2] > 0
  
  if(only_presences) {
    # --- TOP ROW: Presence sites with black outlines ---
    plot_data <- subset_data %>% filter(presence_val == 1)
    
    # Use fill for the environment color and color for the stroke
    color_scale <- if(is_relative) {
      scale_fill_gradient2(limits = env_lims, low = col_low, mid = "grey90", high = col_high, midpoint = 0)
    } else {
      scale_fill_gradient(limits = env_lims, low = col_low, high = col_high)
    }
    
    p <- ggplot(plot_data, aes(x = X, y = Y)) +
      geom_point(aes(fill = !!v_env), shape = 21, color = "black", size = 1.2, stroke = 0.3) +
      color_scale
      
  } else {
    # --- BOTTOM ROW: Full background gradient ---
    color_scale <- if(is_relative) {
      scale_color_gradient2(limits = env_lims, low = col_low, mid = "grey90", high = col_high, midpoint = 0)
    } else {
      scale_color_gradient(limits = env_lims, low = col_low, high = col_high)
    }
    
    p <- ggplot(subset_data, aes(x = X, y = Y)) +
      geom_point(aes(color = !!v_env), size = 0.7, alpha = 0.8) +
      color_scale
  }
  
  p + 
    coord_fixed(xlim = xlims, ylim = ylims) +
    theme_minimal() +
    theme(
      axis.text = element_blank(), axis.title = element_blank(),
      panel.grid = element_blank(), legend.position = "none",
      plot.subtitle = element_text(size = 10, hjust = 0.5)
    ) +
    labs(subtitle = title)
}

```

# Oystercatchers presences over seasonal/annual climate variables 
## Temperature 
I've made some new plots showing the spatial distribution of oystercatchers through time and associated climatic variables. Beforehand I just showed annual variables, but now I first plot winter minimum temperature, breeding season average temperature, and annual average temperature. The values you see are deviations from the mean for that variable in that atlas. 

```{r}
species <- 'Haematopus_ostralegus'
english_name <- 'Oystercatcher'

master_df <- na.omit(master_df)
species_spatial_comparison(master_df,
                           species=species,
                           english_name=english_name,
                           env_var='tmean_winter',
                           col_low = 'dodgerblue3',col_high = 'firebrick3',
                           legend_name = 'Winter minimum temperature',
                           relative=T)
species_spatial_comparison(master_df,
                           species=species,
                           english_name=english_name,
                           env_var='tmean_breeding',
                           col_low = 'dodgerblue3',col_high = 'firebrick3',
                           legend_name = 'Breeding season temperature',
                           relative=T)
species_spatial_comparison(master_df,
                           species=species,
                           english_name=english_name,
                           env_var='tmean_year',
                           col_low = 'dodgerblue3',col_high = 'firebrick3',
                           legend_name = 'Annual temperature',
                           relative=T)
```

## Precipitation 
I repeat this exercise with precipitation! 

```{r}
species_spatial_comparison(master_df,
                           species=species,
                           english_name=english_name,
                           env_var='prec_winter',
                           col_low = 'burlywood',col_high = 'cadetblue',
                           legend_name = 'Winter season precipitation',
                           relative=T)
species_spatial_comparison(master_df,
                           species=species,
                           english_name=english_name,
                           env_var='prec_breeding',
                           col_low = 'burlywood',col_high = 'cadetblue',
                           legend_name = 'Breeding season precipitation',
                           relative=T)
species_spatial_comparison(master_df,
                           species=species,
                           english_name=english_name,
                           env_var='prec_year',
                           col_low = 'burlywood',col_high = 'cadetblue',
                           legend_name = 'Annual precipitation',
                           relative=T)
```

## Correlations 
I also add the pearson correlations between all these climatic variables. The values are quite different from what I sent around previously, that's probably because I bunched all atlases together in one correlation matrix in that one, versus here where I separate by atlas. 

```{r echo=FALSE, fig.height=5, fig.width=10}
library(corrplot)
library(Hmisc)
library(dplyr)
library(tidyr)

# 1. Define variables
target_vars <- c('tmean_winter', 'tmean_breeding', 'tmean_year', 
                 'prec_winter', 'prec_breeding', 'prec_year')

# 2. Setup plotting grid (1 row, 3 columns for your 3 atlases)
par(mfrow = c(1, 3)) 

# 3. Define color palette
col_palette <- colorRampPalette(rev(c("#BB4444", "#EE9988", "#FFFFFF", "#77AADD", "#4477AA")))

# 4. Loop through each atlas and plot
atlases <- sort(unique(master_df$atlas))

t <- lapply(atlases, function(a) {
  
  # Filter data for specific atlas
  atlas_data <- master_df %>%
    filter(atlas == a) %>%
    select(all_of(target_vars)) %>%
    drop_na()
  
  # Calculate Spearman correlations
  # Note: Spearman is excellent here as it captures non-linear monotonic relationships
  cor_results <- rcorr(as.matrix(atlas_data), type = 'pearson')
  
  # Plot
  corrplot(cor_results$r, 
           method = "color", 
           col = col_palette(200),  
           type = 'lower',        # Using 'lower' often makes side-by-side comparison cleaner
           addCoef.col = "black", 
           number.cex = 0.8,      # Smaller text for coefficients to fit
           tl.col = "black", 
           tl.srt = 45, 
           diag = FALSE,
           title = paste("Atlas", a),
           mar = c(0,0,2,0))      # Adjusted margin for the title
})

# Reset plotting parameters to default
par(mfrow = c(1, 1))
```



## Breeding climates
```{r}
# Species presences in environmental space 
species_environment_plots <- function(master_df,species, english_name, env_vars,
                                      x_variable,y_variable) {
  
  ### NOW WE GET TO PLOTTING 
  subs_names <- c('atlas', env_vars, species)
  subs <- master_df[, colnames(master_df) %in% subs_names]
  
  v1 <- sym(env_vars[1])
  v2 <- sym(env_vars[2])
  
  xlims <- master_df %>% pull(!!v1) %>% range(na.rm = TRUE)
  ylims <- master_df %>% pull(!!v2) %>% range(na.rm = TRUE)
  
  # A. Generate a "dummy" plot just to grab the shared legend
  # We make a simple ggplot with the same alpha mapping
  legend_plot <- ggplot(subs, aes(x = !!v1, y = !!v2, alpha = factor(!!sym(species)))) +
    geom_point() +
    scale_alpha_manual(values = c("0" = 0.1, "1" = 1.0), 
                       labels = c("Absence", "Presence"),
                       name = "Status") +
    theme_minimal() +
    theme(legend.position = "bottom")
  
  shared_legend <- get_legend(legend_plot)
  
  # B. Generate the 3 actual plots (with legend = FALSE)
  colors <- c('darkblue','darkgreen','darkred')
  plots <- lapply(c(1, 2, 3), function(x) {
    single_plot(x, subs, color = colors[x], env_vars, xlims, ylims, show_legend = FALSE,x_variable,y_variable)
  })
  
  # C. Combine into a grid
  plot_row <- plot_grid(plotlist = plots, ncol = 3, labels = "AUTO", align = 'h')
  
  # D. Add the legend at the bottom
  final_combined <- plot_grid(plot_row, shared_legend, ncol = 1, rel_heights = c(1, 0.1))
  
  # Add a super title (optional)
  title_theme <- ggdraw() + draw_label(paste(english_name), fontface='bold')
  
  return(plot_grid(title_theme, final_combined, ncol = 1, rel_heights = c(0.1, 1)))
}
single_plot <- function(atlas_nr, subs, color, env_vars, xlims, ylims, show_legend = FALSE,x_variable,y_variable) {
  
  subset_data <- subs[subs$atlas == atlas_nr, ]
  colnames(subset_data)[ncol(subset_data)] <- "presenceabsence"
  subset_data$presenceabsence <- factor(subset_data$presenceabsence, levels = c("0", "1"))
  
  v1 <- sym(env_vars[1])
  v2 <- sym(env_vars[2])
  
  # MAIN PLOT
  pmain <- ggplot(data = subset_data, aes(x = !!v1, y = !!v2, alpha = presenceabsence)) +
    geom_point(col = color) +
    scale_alpha_manual(values = c("0" = 0.1, "1" = 1.0), 
                       labels = c("Absence", "Presence"),
                       name = "Status") +
    scale_x_continuous(limits = xlims) + 
    scale_y_continuous(limits = ylims) +
    theme_minimal() +
    labs(title = paste0("Atlas Period: ", atlas_nr),
         x = x_variable,
         y = y_variable) +
    # Toggle legend based on function call
    theme(legend.position = if(show_legend) "bottom" else "none")
  
  # MARGINAL DENSITIES
  xdens <- axis_canvas(pmain, axis = "x") +
    geom_density(data = subset_data, aes(x = !!v1), color = color, linetype = 3) +
    geom_density(data = subset_data[subset_data$presenceabsence == 1, ], aes(x = !!v1), linetype = 1, color = color)
  
  ydens <- axis_canvas(pmain, axis = "y", coord_flip = TRUE) +
    geom_density(data = subset_data, aes(x = !!v2), linetype = 3, color = color) +
    geom_density(data = subset_data[subset_data$presenceabsence == 1, ], aes(x = !!v2), linetype = 1, color = color) +
    coord_flip()
  
  # COMBINE
  p1 <- insert_xaxis_grob(pmain, xdens, grid::unit(.2, "null"), position = "top")
  p2 <- insert_yaxis_grob(p1, ydens, grid::unit(.2, "null"), position = "right")
  
  return(p2)
}


```
```{r}
species <- 'Haematopus_ostralegus'
english_name <- 'Oystercatcher'
env_var <- 'tmean_year'
env_vars <- c('tmean_breeding','prec_breeding')
title <- 'Annual temperature'


species_environment_plots(master_df = master_df,
                          species=species,
                          english_name=english_name,
                          env_vars=env_vars,
                          x_variable = 'Temperature (breeding season)',
                          y_variable = 'Precipitation (breeding season)')


```

