---
title: "Grid-level environmental maps"
author: "WM"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  results = "hide",
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  dev = "png"
)
knitr::opts_knit$set(root.dir = normalizePath("../.."))
```

## Aim

These maps and distribution plots show the grid-level environmental variables
used by the HMSC analysis. The map panels focus on Atlas 1 and Atlas 3,
labelled here as the 1970s and 2010s. The violin plots include all three atlas
periods so temporal shifts in the land-use distributions are visible.

The map style follows the avian bioregion cluster notebooks: each grid cell is
drawn at its model coordinate, with mainland Denmark as the main panel and
Bornholm shown as an inset.

## Load Data

```{r load-data}
suppressPackageStartupMessages({
  library(tidyverse)
  library(cowplot)
  library(scales)
  library(knitr)
})

source(file.path("support_scripts", "data_helpers.R"))

input_path <- file.path(
  "Data",
  "preprocessed_data_min_occs_is_5_coverage_is_good_and_average.RData"
)
if (!file.exists(input_path)) {
  stop("Missing preprocessed model input file: ", input_path, call. = FALSE)
}

load(input_path)
if (!all(c("X", "Design") %in% ls())) {
  stop("Expected objects `X` and `Design` in: ", input_path, call. = FALSE)
}

output_prefix <- format(Sys.Date(), "%Y-%m-%d")
out_dir <- file.path("notebooks", "exploratory", "outputs", "grid-environment-maps")
plot_dir <- file.path(out_dir, "plots")
dir.create(plot_dir, recursive = TRUE, showWarnings = FALSE)

year_lookup <- c("1" = "1970s", "2" = "1990s", "3" = "2010s")
map_year_lookup <- year_lookup[c("1", "3")]
focal_atlases <- names(year_lookup)

# These are the land-use predictors retained by the fitted HMSC models. The
# freshwater/saltwater percentage is deliberately excluded here because it was
# also excluded from the model formula.
landuse_vars <- c(
  "perc_urban",
  "perc_cropland",
  "perc_pasture",
  "perc_forest",
  "perc_grass_shrub"
)

landuse_relative_vars <- c("hh", landuse_vars)
climate_relative_vars <- c("tmean_breeding", "prec_breeding")
relative_vars <- c(landuse_relative_vars, climate_relative_vars)

var_labels <- c(
  "tmean_breeding" = "Breeding temperature",
  "prec_breeding" = "Breeding precipitation",
  "hh" = "Land-use heterogeneity",
  "perc_urban" = "Urban",
  "perc_cropland" = "Cropland",
  "perc_pasture" = "Pasture",
  "perc_forest" = "Forest",
  "perc_grass_shrub" = "Grass/Shrubland"
)

landuse_labels <- c(
  "perc_urban" = "Urban",
  "perc_cropland" = "Cropland",
  "perc_pasture" = "Pasture",
  "perc_forest" = "Forest",
  "perc_grass_shrub" = "Grass/Shrubland"
)

landuse_palette <- c(
  "Urban" = "#666666",
  "Cropland" = "#D6A11F",
  "Pasture" = "#E68A2E",
  "Forest" = "#2F7D32",
  "Grass/Shrubland" = "#79B75B"
)
```

```{r prepare-data}
design_df <- Design |>
  as.data.frame() |>
  rownames_to_column("survey") |>
  transmute(
    survey,
    site,
    atlas = as.character(.data$atlas),
    X = .data$lon,
    Y = .data$lat
  )

env_df <- X |>
  as.data.frame() |>
  rownames_to_column("survey") |>
  mutate(atlas = stringr::str_extract(.data$survey, "[123]$")) |>
  filter(.data$atlas %in% focal_atlases) |>
  left_join(design_df, by = c("survey", "atlas")) |>
  filter(!is.na(.data$X), !is.na(.data$Y)) |>
  rowwise() |>
  mutate(
    dominant_model_land_use = landuse_vars[
      which.max(c_across(all_of(landuse_vars)))
    ],
    dominant_model_land_use = landuse_labels[[dominant_model_land_use]]
  ) |>
  ungroup() |>
  mutate(
    period = factor(year_lookup[.data$atlas], levels = unname(year_lookup)),
    dominant_model_land_use = factor(
      .data$dominant_model_land_use,
      levels = unname(landuse_labels)
    )
  )

scaling_summary <- env_df |>
  pivot_longer(
    cols = all_of(relative_vars),
    names_to = "variable",
    values_to = "value"
  ) |>
  group_by(.data$period, .data$variable) |>
  summarise(
    variable_label = var_labels[.data$variable[[1]]],
    mean_value = mean(.data$value, na.rm = TRUE),
    sd_value = sd(.data$value, na.rm = TRUE),
    q10 = quantile(.data$value, 0.10, na.rm = TRUE),
    q50 = quantile(.data$value, 0.50, na.rm = TRUE),
    q90 = quantile(.data$value, 0.90, na.rm = TRUE),
    .groups = "drop"
  )

relative_long <- env_df |>
  select(
    survey,
    site,
    atlas,
    period,
    X,
    Y,
    all_of(relative_vars)
  ) |>
  pivot_longer(
    cols = all_of(relative_vars),
    names_to = "variable",
    values_to = "value"
  ) |>
  left_join(
    scaling_summary |> select(period, variable, mean_value, sd_value),
    by = c("period", "variable")
  ) |>
  mutate(
    variable_label = var_labels[.data$variable],
    relative_z = (.data$value - .data$mean_value) / .data$sd_value,
    relative_z = if_else(is.finite(.data$relative_z), .data$relative_z, NA_real_)
  )

dominant_summary <- env_df |>
  count(.data$period, .data$dominant_model_land_use, name = "n_cells") |>
  group_by(.data$period) |>
  mutate(prop_cells = .data$n_cells / sum(.data$n_cells)) |>
  ungroup()

landuse_distribution <- env_df |>
  select(
    survey,
    site,
    atlas,
    period,
    all_of(landuse_vars)
  ) |>
  pivot_longer(
    cols = all_of(landuse_vars),
    names_to = "variable",
    values_to = "proportion"
  ) |>
  mutate(
    land_use_type = factor(var_labels[.data$variable], levels = var_labels[landuse_vars]),
    percentage = .data$proportion * 100,
    has_cover = .data$percentage > 0
  )

safe_quantile <- function(x, prob) {
  x <- x[is.finite(x)]
  if (length(x) == 0) {
    return(NA_real_)
  }
  as.numeric(quantile(x, prob, na.rm = TRUE))
}

landuse_distribution_summary <- landuse_distribution |>
  group_by(.data$period, .data$land_use_type) |>
  summarise(
    n_cells = n(),
    mean_percentage = mean(.data$percentage, na.rm = TRUE),
    median_percentage = median(.data$percentage, na.rm = TRUE),
    q10_percentage = quantile(.data$percentage, 0.10, na.rm = TRUE),
    q90_percentage = quantile(.data$percentage, 0.90, na.rm = TRUE),
    .groups = "drop"
  )

landuse_zero_inflated_summary <- landuse_distribution |>
  group_by(.data$period, .data$land_use_type) |>
  summarise(
    n_cells = n(),
    n_present = sum(.data$has_cover, na.rm = TRUE),
    prop_present = .data$n_present / .data$n_cells,
    mean_positive_percentage = mean(.data$percentage[.data$has_cover], na.rm = TRUE),
    median_positive_percentage = median(.data$percentage[.data$has_cover], na.rm = TRUE),
    q10_positive_percentage = safe_quantile(.data$percentage[.data$has_cover], 0.10),
    q90_positive_percentage = safe_quantile(.data$percentage[.data$has_cover], 0.90),
    .groups = "drop"
  )

write_csv(
  env_df |>
    select(
      survey,
      site,
      atlas,
      period,
      X,
      Y,
      dominant_model_land_use,
      all_of(relative_vars)
    ),
  file.path(out_dir, paste0(output_prefix, "-environment-map-data-all-atlas-periods.csv"))
)
write_csv(
  scaling_summary,
  file.path(out_dir, paste0(output_prefix, "-environment-map-scaling-summary.csv"))
)
write_csv(
  dominant_summary,
  file.path(out_dir, paste0(output_prefix, "-dominant-modelled-land-use-summary.csv"))
)
write_csv(
  landuse_distribution_summary,
  file.path(out_dir, paste0(output_prefix, "-land-use-percentage-distribution-summary.csv"))
)
write_csv(
  landuse_zero_inflated_summary,
  file.path(out_dir, paste0(output_prefix, "-land-use-zero-inflated-distribution-summary.csv"))
)
```

## Map Helpers

```{r map-helpers}
mainland_bbox <- c(xmin = 400000, xmax = 750000, ymin = 6000000, ymax = 6450000)
bornholm_bbox <- c(xmin = 855000, xmax = 905000, ymin = 6100000, ymax = 6160000)

mainland_width <- mainland_bbox[["xmax"]] - mainland_bbox[["xmin"]]
mainland_height <- mainland_bbox[["ymax"]] - mainland_bbox[["ymin"]]
bornholm_width <- bornholm_bbox[["xmax"]] - bornholm_bbox[["xmin"]]
bornholm_height <- bornholm_bbox[["ymax"]] - bornholm_bbox[["ymin"]]
bornholm_inset_width <- bornholm_width / mainland_width
bornholm_inset_height <- bornholm_height / mainland_height

denmark_map_theme <- function(legend_position = "bottom") {
  theme_minimal(base_size = 10) +
    theme(
      legend.position = legend_position,
      legend.title = element_text(face = "bold"),
      axis.text = element_blank(),
      axis.title = element_blank(),
      panel.grid = element_blank(),
      plot.title = element_text(face = "bold", hjust = 0.5),
      plot.background = element_rect(fill = "white", colour = NA),
      plot.margin = margin(0, 0, 0, 0)
    )
}

dominant_landuse_scale <- function() {
  scale_colour_manual(
    values = landuse_palette,
    drop = FALSE,
    name = "Dominant modelled\nland-use class"
  )
}

relative_value_scale <- function() {
  scale_colour_gradient2(
    low = "#2166AC",
    mid = "grey95",
    high = "#B2182B",
    midpoint = 0,
    limits = c(-2.5, 2.5),
    oob = squish,
    name = "Relative value\n(z-score)",
    breaks = c(-2, 0, 2),
    labels = c("Lower", "Average", "Higher")
  )
}

plot_environment_base <- function(df, colour_col, scale_layer, title = NULL,
                                  bbox = mainland_bbox, show_legend = FALSE,
                                  border = FALSE) {
  ggplot(df) +
    geom_point(
      aes(x = .data$X, y = .data$Y, colour = .data[[colour_col]]),
      size = 1.25,
      alpha = 0.95
    ) +
    scale_layer +
    coord_fixed(
      xlim = c(bbox[["xmin"]], bbox[["xmax"]]),
      ylim = c(bbox[["ymin"]], bbox[["ymax"]]),
      expand = FALSE
    ) +
    denmark_map_theme(if (show_legend) "bottom" else "none") +
    theme(
      panel.border = if (border) {
        element_rect(colour = "grey35", fill = NA, linewidth = 0.45)
      } else {
        element_blank()
      }
    ) +
    labs(title = title)
}

plot_denmark_inset_map <- function(df, colour_col, scale_layer, title = NULL,
                                   bornholm_label = FALSE) {
  p_main <- plot_environment_base(
    df,
    colour_col = colour_col,
    scale_layer = scale_layer,
    title = title,
    bbox = mainland_bbox
  )

  p_inset <- plot_environment_base(
    df,
    colour_col = colour_col,
    scale_layer = scale_layer,
    bbox = bornholm_bbox,
    border = TRUE
  ) +
    theme_void() +
    theme(
      legend.position = "none",
      plot.background = element_rect(fill = "white", colour = NA),
      panel.border = element_rect(colour = "grey35", fill = NA, linewidth = 0.7)
    )

  map_plot <- ggdraw(p_main) +
    draw_plot(
      p_inset,
      x = 0.66,
      y = 0.60,
      width = 0.24,
      height = 0.22
    )

  if (isTRUE(bornholm_label)) {
    map_plot <- map_plot +
      draw_label("Bornholm", x = 0.78, y = 0.84, size = 7, colour = "grey30")
  }

  map_plot
}

period_maps <- function(df, colour_col, scale_fn, bornholm_label = FALSE) {
  map(unname(map_year_lookup), function(period_name) {
    plot_denmark_inset_map(
      df |> filter(.data$period == period_name),
      colour_col = colour_col,
      scale_layer = scale_fn(),
      title = period_name,
      bornholm_label = bornholm_label
    )
  })
}

legend_for_map <- function(df, colour_col, scale_fn) {
  plot_environment_base(
    df,
    colour_col = colour_col,
    scale_layer = scale_fn(),
    show_legend = TRUE
  ) |>
    get_legend()
}

relative_variable_row <- function(variable_name) {
  variable_label <- var_labels[[variable_name]]
  maps <- map(unname(map_year_lookup), function(period_name) {
    relative_long |>
      filter(.data$variable == variable_name, .data$period == period_name) |>
      plot_denmark_inset_map(
        colour_col = "relative_z",
        scale_layer = relative_value_scale(),
        title = period_name
      )
  })

  row_label <- ggdraw() +
    draw_label(
      variable_label,
      x = 0.5,
      y = 0.5,
      angle = 90,
      fontface = "bold",
      size = 9
    )

  plot_grid(
    row_label,
    plot_grid(plotlist = maps, nrow = 1, align = "none"),
    nrow = 1,
    rel_widths = c(0.08, 1)
  )
}

relative_variable_panel <- function(variable_names) {
  plot_grid(
    plotlist = map(variable_names, relative_variable_row),
    ncol = 1,
    align = "v"
  )
}
```

## Dominant Land-Use Class

Dominance here is calculated only among the land-use predictors retained by the
HMSC models: urban, cropland, pasture, forest, and grass/shrubland. The
freshwater/saltwater percentage is omitted to match the model formula.

```{r dominant-land-use-map, fig.width=9, fig.height=5}
dominant_panel <- plot_grid(
  plot_grid(
    plotlist = period_maps(
      env_df,
      colour_col = "dominant_model_land_use",
      scale_fn = dominant_landuse_scale,
      bornholm_label = TRUE
    ),
    nrow = 1,
    align = "none"
  ),
  legend_for_map(
    env_df,
    colour_col = "dominant_model_land_use",
    scale_fn = dominant_landuse_scale
  ),
  ncol = 1,
  rel_heights = c(1, 0.16)
)

dominant_panel

ggsave(
  file.path(plot_dir, paste0(output_prefix, "-dominant-modelled-land-use-maps.png")),
  dominant_panel,
  width = 9,
  height = 5,
  dpi = 300,
  bg = "white"
)
```

```{r dominant-land-use-summary, results='asis'}
dominant_summary |>
  mutate(prop_cells = percent(.data$prop_cells, accuracy = 0.1)) |>
  arrange(.data$period, .data$dominant_model_land_use) |>
  kable()
```

## Land-Use Percentage Distributions Through Time

Each violin shows the distribution of grid-cell percentage cover for one
land-use type within each atlas period. The y axis is the raw percentage cover
of that land-use type in the grid cell.

```{r land-use-percentage-violins, fig.width=10, fig.height=6.5}
landuse_violin_plot <- landuse_distribution |>
  ggplot(aes(
    x = .data$period,
    y = .data$percentage,
    fill = .data$land_use_type
  )) +
  geom_violin(
    trim = FALSE,
    scale = "width",
    colour = "grey35",
    linewidth = 0.25,
    alpha = 0.78
  ) +
  geom_boxplot(
    width = 0.13,
    outlier.shape = NA,
    fill = "white",
    colour = "grey25",
    linewidth = 0.25,
    alpha = 0.82
  ) +
  facet_wrap(~land_use_type, ncol = 3) +
  scale_fill_manual(values = landuse_palette, guide = "none") +
  scale_y_continuous(
    limits = c(0, 100),
    labels = label_percent(scale = 1),
    expand = expansion(mult = c(0, 0.04))
  ) +
  labs(
    x = "Atlas period",
    y = "Land-use cover (%)"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid.minor = element_blank(),
    strip.text = element_text(face = "bold"),
    axis.title = element_text(face = "bold"),
    plot.background = element_rect(fill = "white", colour = NA)
  )

landuse_violin_plot

ggsave(
  file.path(plot_dir, paste0(output_prefix, "-land-use-percentage-violins.png")),
  landuse_violin_plot,
  width = 10,
  height = 6.5,
  dpi = 300,
  bg = "white"
)
```

```{r land-use-percentage-summary, results='asis'}
landuse_distribution_summary |>
  mutate(across(ends_with("_percentage"), ~ round(.x, 2))) |>
  arrange(.data$period, .data$land_use_type) |>
  kable()
```

## Zero-Inflated Land-Use Distribution View

The plot below separates the two parts of the distribution. The top panels show
the percentage of grid cells where each land-use type is present at all. The
bottom panels then show the percentage cover distribution only for cells where
that land-use type is present. The bottom panels use free y axes because the
sparse land-use types otherwise disappear against cropland's 0-100% range.

```{r zero-inflated-land-use-view, fig.width=10, fig.height=9.5}
percentage_axis_breaks <- function(limits) {
  axis_limits <- c(max(0, limits[[1]]), min(100, limits[[2]]))
  breaks <- pretty(axis_limits, n = 4)
  breaks[breaks >= 0 & breaks <= 100]
}

presence_plot <- landuse_zero_inflated_summary |>
  ggplot(aes(
    x = .data$period,
    y = .data$prop_present,
    fill = .data$land_use_type
  )) +
  geom_col(
    width = 0.66,
    colour = "grey35",
    linewidth = 0.25,
    alpha = 0.82
  ) +
  facet_wrap(~land_use_type, ncol = 3) +
  scale_fill_manual(values = landuse_palette, guide = "none") +
  scale_y_continuous(
    limits = c(0, 1),
    labels = label_percent(accuracy = 1),
    expand = expansion(mult = c(0, 0.04))
  ) +
  labs(
    x = NULL,
    y = "Cells with >0% cover"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid.minor = element_blank(),
    strip.text = element_text(face = "bold"),
    axis.title = element_text(face = "bold"),
    plot.background = element_rect(fill = "white", colour = NA),
    plot.margin = margin(5.5, 8, 5.5, 12)
  )

positive_cover_plot <- landuse_distribution |>
  filter(.data$has_cover) |>
  ggplot(aes(
    x = .data$period,
    y = .data$percentage,
    fill = .data$land_use_type
  )) +
  geom_violin(
    trim = FALSE,
    scale = "width",
    colour = "grey35",
    linewidth = 0.25,
    alpha = 0.78
  ) +
  geom_boxplot(
    width = 0.13,
    outlier.shape = NA,
    fill = "white",
    colour = "grey25",
    linewidth = 0.25,
    alpha = 0.82
  ) +
  facet_wrap(~land_use_type, ncol = 3, scales = "free_y") +
  scale_fill_manual(values = landuse_palette, guide = "none") +
  scale_y_continuous(
    labels = label_percent(scale = 1),
    breaks = percentage_axis_breaks,
    expand = expansion(mult = c(0.03, 0.08))
  ) +
  labs(
    x = "Atlas period",
    y = "Cover where present (%)"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid.minor = element_blank(),
    strip.text = element_text(face = "bold"),
    axis.title = element_text(face = "bold"),
    plot.background = element_rect(fill = "white", colour = NA),
    plot.margin = margin(5.5, 8, 5.5, 12)
  )

zero_inflated_landuse_plot <- plot_grid(
  presence_plot,
  positive_cover_plot,
  ncol = 1,
  rel_heights = c(0.9, 1.35),
  align = "v"
)

zero_inflated_landuse_plot

ggsave(
  file.path(plot_dir, paste0(output_prefix, "-land-use-zero-inflated-hurdle-plot.png")),
  zero_inflated_landuse_plot,
  width = 10,
  height = 9.5,
  dpi = 300,
  bg = "white"
)
```

```{r zero-inflated-land-use-summary, results='asis'}
landuse_zero_inflated_summary |>
  mutate(
    prop_present = percent(.data$prop_present, accuracy = 0.1),
    across(ends_with("_percentage"), ~ round(.x, 2))
  ) |>
  arrange(.data$period, .data$land_use_type) |>
  kable()
```

## Land-Use Relative High-Low Maps

The maps below show whether each grid cell is lower or higher than the
same-period distribution for that variable. Values are standardized within each
`variable x period` combination before plotting, so the colours emphasize
spatial high-low regions rather than raw units or period-level shifts.

```{r land-use-relative-maps, fig.width=9, fig.height=13.5}
landuse_relative_panel <- plot_grid(
  relative_variable_panel(landuse_relative_vars),
  legend_for_map(
    relative_long |> filter(.data$variable == "hh"),
    colour_col = "relative_z",
    scale_fn = relative_value_scale
  ),
  ncol = 1,
  rel_heights = c(1, 0.07)
)

landuse_relative_panel

ggsave(
  file.path(plot_dir, paste0(output_prefix, "-land-use-relative-high-low-maps.png")),
  landuse_relative_panel,
  width = 9,
  height = 13.5,
  dpi = 300,
  bg = "white"
)
```

## Climate Relative High-Low Maps

These climate panels use the breeding-season temperature and precipitation
variables included in the fitted models. As above, colours are relative to each
period's own spatial distribution.

```{r climate-relative-maps, fig.width=9, fig.height=5.4}
climate_relative_panel <- plot_grid(
  relative_variable_panel(climate_relative_vars),
  legend_for_map(
    relative_long |> filter(.data$variable == "tmean_breeding"),
    colour_col = "relative_z",
    scale_fn = relative_value_scale
  ),
  ncol = 1,
  rel_heights = c(1, 0.14)
)

climate_relative_panel

ggsave(
  file.path(plot_dir, paste0(output_prefix, "-climate-relative-high-low-maps.png")),
  climate_relative_panel,
  width = 9,
  height = 5.4,
  dpi = 300,
  bg = "white"
)
```

## Scaling Summary

```{r scaling-summary, results='asis'}
scaling_summary |>
  mutate(variable_label = factor(.data$variable_label, levels = var_labels[relative_vars])) |>
  arrange(.data$period, .data$variable_label) |>
  select(
    period,
    variable_label,
    mean_value,
    sd_value,
    q10,
    q50,
    q90
  ) |>
  kable(digits = 3)
```
