---
title: "Ecologically informed GWR on modelled avian richness"
author: "WM"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center"
)

input_path <- knitr::current_input(dir = TRUE)
candidate_roots <- unique(c(
  normalizePath(".", mustWork = TRUE),
  if (!is.null(input_path) && nzchar(input_path)) {
    normalizePath(file.path(dirname(input_path), "..", ".."), mustWork = FALSE)
  },
  normalizePath("../..", mustWork = FALSE)
))
project_root <- candidate_roots[
  file.exists(file.path(candidate_roots, "hmsc-hpc-cv-pipeline.Rproj"))
][[1]]
knitr::opts_knit$set(root.dir = project_root)

required_packages <- c("dplyr", "readr", "tidyr", "knitr")
missing_packages <- required_packages[
  !vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)
]
if (length(missing_packages) > 0) {
  stop(
    "Install these R packages before rendering this notebook: ",
    paste(missing_packages, collapse = ", "),
    call. = FALSE
  )
}

library(dplyr)
library(readr)
library(tidyr)
```

## Aim

This notebook reads the cached geographically weighted regression outputs created
by `misc-figures/scripts/fig10-gwr-richness-turnover.R` and
`misc-figures/scripts/fig11-gwr-richness-univariate.R`. The expensive GWR fitting
now lives in those plotting scripts so this report can be re-rendered quickly
without re-running the regressions.

All response variables are derived from HMSC modelled outputs rather than raw
observations:

- predicted richness: summed species occurrence probabilities per grid cell
- all-site beta turnover: mean Bray-Curtis dissimilarity from each grid cell to
  all other atlas grid cells in the same period

The focal GWR uses a fixed 100 km bisquare bandwidth. The mapped driver at each
grid cell is the environmental predictor with the largest absolute local
t-value; cells with `|t| < 1.96` are labelled as having no significant driver.
The richness-only univariate workflow fits the same richness response one
predictor at a time, plus an intercept-only baseline model, to compare how
single-predictor support changes through time. That workflow is restricted to
grid cells present in all three atlas periods, so every mapped period is based on
the same spatial footprint.

## Cached Outputs

```{r load-cached-gwr}
pattern <- "2026-03-13"
gwr_out_dir <- file.path("notebooks", "exploratory", "outputs", "gwr")
gwr_model_dir <- file.path(gwr_out_dir, "models")
figure_out_dir <- file.path("misc-figures", "outputs", "main")

gwr_fit_path <- file.path(gwr_model_dir, paste0(pattern, "-gwr-fits-richness-beta-100km.rds"))
gwr_inputs_path <- file.path(gwr_model_dir, paste0(pattern, "-gwr-analysis-inputs-100km.rds"))
gwr_summary_path <- file.path(gwr_out_dir, paste0(pattern, "-gwr-100km-summary.csv"))
gwr_local_path <- file.path(gwr_out_dir, paste0(pattern, "-gwr-local-coefficients-100km.csv"))
gwr_dominant_path <- file.path(gwr_out_dir, paste0(pattern, "-gwr-dominant-drivers-100km.csv"))
figure_path <- file.path(figure_out_dir, paste0(pattern, "-fig10-gwr-richness-turnover.png"))
gwr_effect_ridge_summary_path <- file.path(gwr_out_dir, paste0(pattern, "-gwr-multivariate-effect-ridge-summary-100km.csv"))
gwr_effect_ridge_density_path <- file.path(gwr_out_dir, paste0(pattern, "-gwr-multivariate-effect-ridge-density-100km.csv"))
gwr_effect_ridge_figure_path <- file.path(figure_out_dir, paste0(pattern, "-fig10b-gwr-multivariate-effect-ridges.png"))
gwr_effect_forest_summary_path <- file.path(gwr_out_dir, paste0(pattern, "-gwr-multivariate-effect-forest-interval-summary-100km.csv"))
gwr_effect_forest_richness_figure_path <- file.path(figure_out_dir, paste0(pattern, "-fig10e1-gwr-richness-effect-forest-intervals.png"))
gwr_effect_forest_turnover_figure_path <- file.path(figure_out_dir, paste0(pattern, "-fig10e2-gwr-turnover-effect-forest-intervals.png"))
gwr_effect_forest_combined_figure_path <- file.path(figure_out_dir, paste0(pattern, "-fig10e-gwr-richness-turnover-effect-forest-intervals.png"))
gwr_richness_line_ridge_summary_path <- file.path(gwr_out_dir, paste0(pattern, "-gwr-richness-multivariate-effect-ridge-line-summary-100km.csv"))
gwr_richness_line_ridge_density_path <- file.path(gwr_out_dir, paste0(pattern, "-gwr-richness-multivariate-effect-ridge-line-density-100km.csv"))
gwr_richness_line_ridge_figure_path <- file.path(figure_out_dir, paste0(pattern, "-fig10c-gwr-richness-multivariate-effect-ridge-lines.png"))

richness_site_filter <- "common-sites"
richness_univariate_fit_path <- file.path(gwr_model_dir, paste0(pattern, "-richness-univariate-gwr-", richness_site_filter, "-fits-100km.rds"))
richness_univariate_inputs_path <- file.path(gwr_model_dir, paste0(pattern, "-richness-univariate-gwr-", richness_site_filter, "-analysis-inputs-100km.rds"))
richness_univariate_summary_path <- file.path(gwr_out_dir, paste0(pattern, "-richness-univariate-gwr-", richness_site_filter, "-summary-100km.csv"))
richness_univariate_local_path <- file.path(gwr_out_dir, paste0(pattern, "-richness-univariate-gwr-", richness_site_filter, "-local-results-100km.csv"))
richness_univariate_comparison_path <- file.path(gwr_out_dir, paste0(pattern, "-richness-univariate-gwr-", richness_site_filter, "-model-comparison-100km.csv"))
richness_univariate_strongest_path <- file.path(gwr_out_dir, paste0(pattern, "-richness-univariate-gwr-", richness_site_filter, "-strongest-associations-100km.csv"))
richness_univariate_comparison_figure_path <- file.path(figure_out_dir, paste0(pattern, "-fig11-gwr-richness-univariate-", richness_site_filter, "-model-comparison.png"))
richness_univariate_strongest_figure_path <- file.path(figure_out_dir, paste0(pattern, "-fig11-gwr-richness-univariate-", richness_site_filter, "-strongest-associations.png"))
richness_univariate_map_paths <- sort(Sys.glob(file.path(figure_out_dir, paste0(pattern, "-fig11-gwr-richness-univariate-", richness_site_filter, "-map-*.png"))))

cwm_sti_site_filter <- "common-sites"
cwm_sti_univariate_fit_path <- file.path(gwr_model_dir, paste0(pattern, "-cwm-sti-univariate-gwr-", cwm_sti_site_filter, "-fits-100km.rds"))
cwm_sti_univariate_inputs_path <- file.path(gwr_model_dir, paste0(pattern, "-cwm-sti-univariate-gwr-", cwm_sti_site_filter, "-analysis-inputs-100km.rds"))
cwm_sti_univariate_summary_path <- file.path(gwr_out_dir, paste0(pattern, "-cwm-sti-univariate-gwr-", cwm_sti_site_filter, "-summary-100km.csv"))
cwm_sti_univariate_local_path <- file.path(gwr_out_dir, paste0(pattern, "-cwm-sti-univariate-gwr-", cwm_sti_site_filter, "-local-results-100km.csv"))
cwm_sti_univariate_comparison_path <- file.path(gwr_out_dir, paste0(pattern, "-cwm-sti-univariate-gwr-", cwm_sti_site_filter, "-model-comparison-100km.csv"))
cwm_sti_univariate_strongest_path <- file.path(gwr_out_dir, paste0(pattern, "-cwm-sti-univariate-gwr-", cwm_sti_site_filter, "-strongest-associations-100km.csv"))
cwm_sti_univariate_comparison_figure_path <- file.path(figure_out_dir, paste0(pattern, "-fig12-gwr-cwm-sti-univariate-", cwm_sti_site_filter, "-model-comparison.png"))
cwm_sti_univariate_strongest_figure_path <- file.path(figure_out_dir, paste0(pattern, "-fig12-gwr-cwm-sti-univariate-", cwm_sti_site_filter, "-strongest-associations.png"))
cwm_sti_univariate_map_paths <- sort(Sys.glob(file.path(figure_out_dir, paste0(pattern, "-fig12-gwr-cwm-sti-univariate-", cwm_sti_site_filter, "-map-*.png"))))
cwm_sti_univariate_effect_map_paths <- sort(Sys.glob(file.path(figure_out_dir, paste0(pattern, "-fig12-gwr-cwm-sti-univariate-", cwm_sti_site_filter, "-effect-map-*.png"))))

required_outputs <- c(
  gwr_fit_path,
  gwr_inputs_path,
  gwr_summary_path,
  gwr_local_path,
  gwr_dominant_path,
  figure_path,
  gwr_effect_ridge_summary_path,
  gwr_effect_ridge_density_path,
  gwr_effect_ridge_figure_path,
  gwr_effect_forest_summary_path,
  gwr_effect_forest_richness_figure_path,
  gwr_effect_forest_turnover_figure_path,
  gwr_effect_forest_combined_figure_path,
  gwr_richness_line_ridge_summary_path,
  gwr_richness_line_ridge_density_path,
  gwr_richness_line_ridge_figure_path,
  richness_univariate_fit_path,
  richness_univariate_inputs_path,
  richness_univariate_summary_path,
  richness_univariate_local_path,
  richness_univariate_comparison_path,
  richness_univariate_strongest_path,
  richness_univariate_comparison_figure_path,
  richness_univariate_strongest_figure_path,
  cwm_sti_univariate_fit_path,
  cwm_sti_univariate_inputs_path,
  cwm_sti_univariate_summary_path,
  cwm_sti_univariate_local_path,
  cwm_sti_univariate_comparison_path,
  cwm_sti_univariate_strongest_path,
  cwm_sti_univariate_comparison_figure_path,
  cwm_sti_univariate_strongest_figure_path
)
missing_outputs <- required_outputs[!file.exists(required_outputs)]

if (length(missing_outputs) > 0 || length(richness_univariate_map_paths) == 0 || length(cwm_sti_univariate_map_paths) == 0 || length(cwm_sti_univariate_effect_map_paths) == 0) {
  stop(
    "Cached GWR outputs are missing. Run `Rscript misc-figures/scripts/fig10-gwr-richness-turnover.R`, `Rscript misc-figures/scripts/fig11-gwr-richness-univariate.R`, and `Rscript misc-figures/scripts/fig12-gwr-cwm-sti-univariate.R` first.\nMissing: ",
    paste(missing_outputs, collapse = ", "),
    if (length(richness_univariate_map_paths) == 0) "\nNo richness-univariate predictor map PNGs were found." else "",
    if (length(cwm_sti_univariate_map_paths) == 0) "\nNo CWM-STI univariate predictor map PNGs were found." else "",
    if (length(cwm_sti_univariate_effect_map_paths) == 0) "\nNo CWM-STI univariate local-effect map PNGs were found." else "",
    call. = FALSE
  )
}

period_col_types <- cols(period = col_character(), .default = col_guess())

gwr_summary <- read_csv(gwr_summary_path, col_types = period_col_types, show_col_types = FALSE)
gwr_local <- read_csv(gwr_local_path, col_types = period_col_types, show_col_types = FALSE)
gwr_dominant <- read_csv(gwr_dominant_path, col_types = period_col_types, show_col_types = FALSE)
gwr_effect_ridge_summary <- read_csv(gwr_effect_ridge_summary_path, col_types = period_col_types, show_col_types = FALSE)
gwr_effect_forest_summary <- read_csv(gwr_effect_forest_summary_path, col_types = period_col_types, show_col_types = FALSE)
gwr_richness_line_ridge_summary <- read_csv(gwr_richness_line_ridge_summary_path, col_types = period_col_types, show_col_types = FALSE)
richness_univariate_summary <- read_csv(richness_univariate_summary_path, col_types = period_col_types, show_col_types = FALSE)
richness_univariate_comparison <- read_csv(richness_univariate_comparison_path, col_types = period_col_types, show_col_types = FALSE)
richness_univariate_strongest <- read_csv(richness_univariate_strongest_path, col_types = period_col_types, show_col_types = FALSE)
richness_univariate_inputs <- readRDS(richness_univariate_inputs_path)
cwm_sti_univariate_summary <- read_csv(cwm_sti_univariate_summary_path, col_types = period_col_types, show_col_types = FALSE)
cwm_sti_univariate_comparison <- read_csv(cwm_sti_univariate_comparison_path, col_types = period_col_types, show_col_types = FALSE)
cwm_sti_univariate_strongest <- read_csv(cwm_sti_univariate_strongest_path, col_types = period_col_types, show_col_types = FALSE)
cwm_sti_univariate_inputs <- readRDS(cwm_sti_univariate_inputs_path)
figure_path_absolute <- normalizePath(figure_path, mustWork = TRUE)
gwr_effect_ridge_figure_path_absolute <- normalizePath(gwr_effect_ridge_figure_path, mustWork = TRUE)
gwr_effect_forest_combined_figure_path_absolute <- normalizePath(gwr_effect_forest_combined_figure_path, mustWork = TRUE)
gwr_effect_forest_turnover_figure_path_absolute <- normalizePath(gwr_effect_forest_turnover_figure_path, mustWork = TRUE)
gwr_richness_line_ridge_figure_path_absolute <- normalizePath(gwr_richness_line_ridge_figure_path, mustWork = TRUE)
richness_univariate_comparison_figure_path_absolute <- normalizePath(richness_univariate_comparison_figure_path, mustWork = TRUE)
richness_univariate_strongest_figure_path_absolute <- normalizePath(richness_univariate_strongest_figure_path, mustWork = TRUE)
richness_univariate_map_paths_absolute <- normalizePath(richness_univariate_map_paths, mustWork = TRUE)
cwm_sti_univariate_comparison_figure_path_absolute <- normalizePath(cwm_sti_univariate_comparison_figure_path, mustWork = TRUE)
cwm_sti_univariate_strongest_figure_path_absolute <- normalizePath(cwm_sti_univariate_strongest_figure_path, mustWork = TRUE)
cwm_sti_univariate_map_paths_absolute <- normalizePath(cwm_sti_univariate_map_paths, mustWork = TRUE)
cwm_sti_univariate_effect_map_paths_absolute <- normalizePath(cwm_sti_univariate_effect_map_paths, mustWork = TRUE)
```

The cached model object and analysis inputs are stored at:

```{r cached-paths, results='asis'}
knitr::kable(
  tibble::tibble(
    output = c(
      "Fitted multivariable GWR models",
      "Multivariable analysis inputs",
      "Multivariable local coefficients",
      "Multivariable dominant drivers",
      "Multivariable figure",
      "Multivariable effect-ridge summary",
      "Multivariable effect-ridge densities",
      "Multivariable effect-ridge figure",
      "Multivariable forest-interval summary",
      "Richness forest-interval figure",
      "Turnover forest-interval figure",
      "Combined richness-turnover forest-interval figure",
      "Richness multivariable ridge-line summary",
      "Richness multivariable ridge-line densities",
      "Richness multivariable ridge-line figure",
      "Fitted richness-univariate GWR models",
      "Richness-univariate analysis inputs",
      "Richness-univariate local results",
      "Richness-univariate model comparison",
      "Richness-univariate strongest associations",
      "Richness-univariate comparison figure",
      "Richness-univariate strongest-association map",
      "Fitted CWM-STI univariate GWR models",
      "CWM-STI univariate analysis inputs",
      "CWM-STI univariate local results",
      "CWM-STI univariate model comparison",
      "CWM-STI univariate strongest associations",
      "CWM-STI univariate comparison figure",
      "CWM-STI univariate strongest-association map"
    ),
    path = c(
      gwr_fit_path,
      gwr_inputs_path,
      gwr_local_path,
      gwr_dominant_path,
      figure_path,
      gwr_effect_ridge_summary_path,
      gwr_effect_ridge_density_path,
      gwr_effect_ridge_figure_path,
      gwr_effect_forest_summary_path,
      gwr_effect_forest_richness_figure_path,
      gwr_effect_forest_turnover_figure_path,
      gwr_effect_forest_combined_figure_path,
      gwr_richness_line_ridge_summary_path,
      gwr_richness_line_ridge_density_path,
      gwr_richness_line_ridge_figure_path,
      richness_univariate_fit_path,
      richness_univariate_inputs_path,
      richness_univariate_local_path,
      richness_univariate_comparison_path,
      richness_univariate_strongest_path,
      richness_univariate_comparison_figure_path,
      richness_univariate_strongest_figure_path,
      cwm_sti_univariate_fit_path,
      cwm_sti_univariate_inputs_path,
      cwm_sti_univariate_local_path,
      cwm_sti_univariate_comparison_path,
      cwm_sti_univariate_strongest_path,
      cwm_sti_univariate_comparison_figure_path,
      cwm_sti_univariate_strongest_figure_path
    )
  ),
  caption = "Local cached outputs used by this notebook."
)
```

## Figure

```{r fig10-gwr, out.width='100%'}
knitr::include_graphics(figure_path_absolute, rel_path = FALSE)
```

## Multivariate Effect Forest Intervals

This forest-style panel summarises the local t-value distributions from the
multivariate GWRs for both predicted richness and all-site beta turnover. Thin
intervals are the 5-95% spread across grid cells, thick intervals are the
25-75% spread, and points are medians.

```{r fig10e-gwr-effect-forest-intervals, out.width='100%'}
knitr::include_graphics(gwr_effect_forest_combined_figure_path_absolute, rel_path = FALSE)
```

```{r gwr-effect-forest-counts, results='asis'}
gwr_effect_forest_summary |>
  transmute(
    response_label,
    period,
    driver_label,
    n_cells,
    n_supported_negative,
    n_supported_positive,
    percent_supported_negative,
    percent_supported_positive,
    local_t_median
  ) |>
  arrange(.data$response_label, .data$period, .data$driver_label) |>
  knitr::kable(
    digits = 1,
    caption = "Supported positive and negative local effects for the multivariate GWR forest-interval panel."
  )
```

## Richness Effect Ridges

This ridge-line plot uses the local t-values from the multivariate richness GWR.
Each environmental layer keeps the Figure 2 variable-importance colour, while
the three atlas periods are overlaid with increasing opacity from oldest to
newest.

```{r fig10c-richness-gwr-effect-ridge-lines, out.width='100%'}
knitr::include_graphics(gwr_richness_line_ridge_figure_path_absolute, rel_path = FALSE)
```

```{r richness-gwr-effect-ridge-counts, results='asis'}
gwr_richness_line_ridge_summary |>
  transmute(
    period,
    driver_label,
    n_cells,
    n_supported_negative,
    n_supported_positive,
    percent_supported_negative,
    percent_supported_positive,
    local_t_median
  ) |>
  arrange(.data$period, .data$driver_label) |>
  knitr::kable(
    digits = 1,
    caption = "Supported positive and negative local effects for the multivariate richness GWR."
  )
```

## Multivariate Effect Ridges

This ridge plot uses the local t-values from the multivariate GWRs, not the
single-predictor models. Blue density sits on the negative side of the local
effect, red density sits on the positive side, and the side labels count grid
cells with supported effects at `|t| >= 1.96`.

```{r fig10b-gwr-effect-ridges, out.width='100%'}
knitr::include_graphics(gwr_effect_ridge_figure_path_absolute, rel_path = FALSE)
```

```{r gwr-effect-ridge-counts, results='asis'}
gwr_effect_ridge_summary |>
  mutate(
    max_supported_direction = if_else(
      .data$n_supported_positive >= .data$n_supported_negative,
      "Positive",
      "Negative"
    ),
    max_supported_cells = pmax(.data$n_supported_positive, .data$n_supported_negative),
    max_supported_percent = pmax(.data$percent_supported_positive, .data$percent_supported_negative)
  ) |>
  group_by(.data$response_label, .data$period) |>
  slice_max(.data$max_supported_cells, n = 3, with_ties = FALSE) |>
  ungroup() |>
  select(
    response_label,
    period,
    driver_label,
    max_supported_direction,
    max_supported_cells,
    max_supported_percent,
    n_supported_negative,
    n_supported_positive
  ) |>
  arrange(.data$response_label, .data$period, desc(.data$max_supported_cells)) |>
  knitr::kable(
    digits = 1,
    caption = "Environmental layers with the largest number of supported local effects in the multivariate GWRs."
  )
```

## Model Summary

```{r model-summary, results='asis'}
gwr_summary |>
  mutate(
    response = recode(
      .data$response,
      predicted_richness = "Predicted richness",
      beta_turnover = "All-site beta turnover"
    ),
    bandwidth_km = .data$bandwidth_value / 1000
  ) |>
  select(
    response,
    period,
    bandwidth_km,
    aicc,
    rss,
    local_r2_mean,
    local_r2_median,
    local_r2_min,
    local_r2_max,
    local_cn_max,
    local_cn_over_30
  ) |>
  arrange(response, period) |>
  knitr::kable(
    digits = 4,
    caption = "Summary of fixed 100 km GWR fits for predicted richness and all-site beta turnover."
  )
```

## Dominant Drivers

```{r dominant-driver-counts, results='asis'}
gwr_dominant |>
  mutate(
    response = recode(
      .data$response,
      predicted_richness = "Predicted richness",
      beta_turnover = "All-site beta turnover"
    )
  ) |>
  count(.data$response, .data$period, .data$dominant_driver, name = "n_cells") |>
  group_by(.data$response, .data$period) |>
  mutate(percent_cells = 100 * .data$n_cells / sum(.data$n_cells)) |>
  ungroup() |>
  arrange(.data$response, .data$period, desc(.data$n_cells)) |>
  knitr::kable(
    digits = 1,
    caption = "Cell counts and percentages for dominant local GWR drivers."
  )
```

## Richness-only univariate models

This cache fits predicted richness against one local predictor at a time, plus an
intercept-only baseline, after keeping only grid cells found in all three atlas
periods. These models are not conditional on the other environmental predictors,
so the maps should be read as local univariate associations rather than
multivariable dominant drivers.

```{r richness-univariate-common-site-summary, results='asis'}
richness_univariate_inputs$common_site_summary |>
  knitr::kable(
    caption = "Grid-cell filtering for the common-site richness-only GWR run."
  )
```

```{r fig11-richness-univariate-comparison, out.width='100%'}
knitr::include_graphics(richness_univariate_comparison_figure_path_absolute, rel_path = FALSE)
```

```{r richness-univariate-model-comparison-table, results='asis'}
richness_univariate_comparison |>
  mutate(
    bandwidth_km = .data$bandwidth_value / 1000
  ) |>
  filter(.data$period_rank <= 3 | .data$model_id == "intercept_only") |>
  select(
    period,
    model_label,
    bandwidth_km,
    period_rank,
    delta_aicc,
    aicc_weight,
    local_r2_mean,
    gw_r2,
    beats_intercept
  ) |>
  arrange(.data$period, .data$period_rank, .data$model_label) |>
  knitr::kable(
    digits = 4,
    caption = "Top richness-only univariate GWR models by period, with the intercept-only baseline retained for comparison."
  )
```

```{r fig11-richness-univariate-strongest, out.width='100%'}
knitr::include_graphics(richness_univariate_strongest_figure_path_absolute, rel_path = FALSE)
```

```{r richness-univariate-strongest-counts, results='asis'}
richness_univariate_strongest |>
  count(.data$period, .data$dominant_driver, name = "n_cells") |>
  group_by(.data$period) |>
  mutate(percent_cells = 100 * .data$n_cells / sum(.data$n_cells)) |>
  ungroup() |>
  arrange(.data$period, desc(.data$n_cells)) |>
  knitr::kable(
    digits = 1,
    caption = "Cell counts and percentages for the strongest supported univariate richness association."
  )
```

### Predictor-specific richness maps

Each map shows the local t-value for a single-predictor richness GWR. Grey cells
have `|t| < 1.96`.

```{r fig11-richness-univariate-predictor-maps, out.width='100%'}
knitr::include_graphics(richness_univariate_map_paths_absolute, rel_path = FALSE)
```

## CWM-STI univariate models

This cache rebuilds the response as the community weighted mean of species
thermal index from HMSC posterior site predictions. Like the richness-only
workflow, it keeps only grid cells found in all three atlas periods and fits one
predictor at a time plus an intercept-only baseline.

```{r cwm-sti-univariate-common-site-summary, results='asis'}
cwm_sti_univariate_inputs$common_site_summary |>
  knitr::kable(
    caption = "Grid-cell filtering for the common-site CWM-STI GWR run."
  )
```

```{r cwm-sti-univariate-sti-summary, results='asis'}
cwm_sti_univariate_inputs$sti_summary |>
  knitr::kable(
    digits = 3,
    caption = "Species thermal index summary for species used in CWM-STI."
  )
```

```{r fig12-cwm-sti-univariate-comparison, out.width='100%'}
knitr::include_graphics(cwm_sti_univariate_comparison_figure_path_absolute, rel_path = FALSE)
```

```{r cwm-sti-univariate-model-comparison-table, results='asis'}
cwm_sti_univariate_comparison |>
  mutate(
    bandwidth_km = .data$bandwidth_value / 1000
  ) |>
  filter(.data$period_rank <= 3 | .data$model_id == "intercept_only") |>
  select(
    period,
    model_label,
    bandwidth_km,
    period_rank,
    delta_aicc,
    aicc_weight,
    local_r2_mean,
    gw_r2,
    beats_intercept
  ) |>
  arrange(.data$period, .data$period_rank, .data$model_label) |>
  knitr::kable(
    digits = 4,
    caption = "Top CWM-STI univariate GWR models by period, with the intercept-only baseline retained for comparison."
  )
```

```{r fig12-cwm-sti-univariate-strongest, out.width='100%'}
knitr::include_graphics(cwm_sti_univariate_strongest_figure_path_absolute, rel_path = FALSE)
```

```{r cwm-sti-univariate-strongest-counts, results='asis'}
cwm_sti_univariate_strongest |>
  count(.data$period, .data$dominant_driver, name = "n_cells") |>
  group_by(.data$period) |>
  mutate(percent_cells = 100 * .data$n_cells / sum(.data$n_cells)) |>
  ungroup() |>
  arrange(.data$period, desc(.data$n_cells)) |>
  knitr::kable(
    digits = 1,
    caption = "Cell counts and percentages for the strongest supported univariate CWM-STI association."
  )
```

### Predictor-specific CWM-STI maps

Each map shows the local t-value for a single-predictor CWM-STI GWR. Grey cells
have `|t| < 1.96`.

```{r fig12-cwm-sti-univariate-predictor-maps, out.width='100%'}
knitr::include_graphics(cwm_sti_univariate_map_paths_absolute, rel_path = FALSE)
```

### Individual local-effect CWM-STI maps

Each map shows the local coefficient from the single-predictor CWM-STI GWR.
Predictors were scaled within atlas period, so the mapped value is the expected
change in CWM-STI per one standard deviation increase in that predictor. Grey
cells have `|t| < 1.96`.

```{r fig12-cwm-sti-univariate-effect-maps, out.width='100%'}
knitr::include_graphics(cwm_sti_univariate_effect_map_paths_absolute, rel_path = FALSE)
```

## Interpretation Notes

The richness GWR asks which environmental variables explain spatial variation in
the number of modelled species. The beta-turnover GWR asks which environmental
variables explain where modelled community composition is most distinct from
the full set of atlas grid cells in the same period. These are related but not
equivalent: richness can be spatially smooth while composition turns over.

To refresh the cached results and figure, run:

```{r refresh-command, eval=FALSE, echo=TRUE}
Rscript misc-figures/scripts/fig10-gwr-richness-turnover.R --force-refit
Rscript misc-figures/scripts/fig11-gwr-richness-univariate.R --force-refit
Rscript misc-figures/scripts/fig12-gwr-cwm-sti-univariate.R --force-refit
```
