---
title: "Empirical Atlas Richness and Turnover"
author: "WM"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  dev = "png"
)
knitr::opts_knit$set(root.dir = normalizePath("../.."))
```

```{r libraries}
suppressPackageStartupMessages({
  library(cowplot)
  library(dplyr)
  library(ggplot2)
  library(sf)
})
```

## Overview

This report uses empirical atlas presences from the raw occurrence matrix, not
HMSC predictions. The saved model-input files are used only to define the
survey rows and retained species for each coverage filter.

First, the analysis is shown for the `Good/Fair` coverage set, which matches
the current model input. Then the same analysis is repeated for `Good` coverage
only, so we can see whether stricter sampling coverage changes the spatial
richness, delta richness, species-level turnover, and guild-level patterns.

Sampling-effort estimates are only available for Atlas 1 and Atlas 2. For Atlas
3, the visualisations therefore include all available Atlas 3 sites from the
preprocessed atlas data rather than applying a `Good` or `Fair` effort
filter. In the shared-site analyses, a site is retained only if it appears in
all three atlas periods after applying the Atlas 1/2 coverage filter and the
unfiltered Atlas 3 site set.

All richness maps count observed atlas presences for retained species. Delta
maps and turnover summaries use only grid cells present in all three atlas
periods within each coverage filter.

Species-level change classes are based on the change in the proportion of
shared grid cells occupied by each species. A species is classified as
`Increase` when the later atlas occupies at least `0.10` more of the shared
grid-cell set than the earlier atlas, `Decrease` when it occupies at least
`0.10` less, and `Stable` otherwise. In plain terms, this is a 10
percentage-point threshold in occupied shared grid cells.

```{r constants-paths}
grid_threshold_path <- file.path(
  "Data", "data", "1_preprocessing", "atlas-grids",
  "grids-ocean-thresholds", "grids_ocean_thresholds.shp"
)

occurrence_path <- file.path(
  "Data", "data", "1_preprocessing", "Y_occurrences", "Y_occurrences.csv"
)

model_input_good_average_path <- file.path(
  "Data", "preprocessed_data_min_occs_is_5_coverage_is_good_and_average.RData"
)

model_input_good_path <- file.path(
  "Data", "preprocessed_data_min_occs_is_5_coverage_is_good.RData"
)

traits_path <- file.path(
  "Data", "data", "1_preprocessing", "Tr_aits", "traits-guild_migration.csv"
)

traits_common_name_path <- file.path(
  "Data", "data", "0_traits", "birdTraits-merged-subset.csv"
)

sti_common_name_path <- file.path("Data", "sti", "STI_Devictor.csv")

out_dir <- file.path(
  "notebooks", "exploratory", "outputs", "shared-site-richness-turnover"
)
plot_dir <- file.path(out_dir, "plots")

dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(plot_dir, recursive = TRUE, showWarnings = FALSE)

atlas_lookup <- c(
  "1" = "1970s",
  "2" = "1990s",
  "3" = "2010s"
)

comparisons <- data.frame(
  comparison_id = c("1990s_minus_1970s", "2010s_minus_1990s"),
  comparison_label = c("1990s minus 1970s", "2010s minus 1990s"),
  start_atlas = c("1", "2"),
  end_atlas = c("2", "3"),
  stringsAsFactors = FALSE
)

coverage_definitions <- list(
  good_average = list(
    coverage_id = "good_average",
    coverage_label = "Good/Fair",
    model_input_path = model_input_good_average_path
  ),
  good = list(
    coverage_id = "good",
    coverage_label = "Good",
    model_input_path = model_input_good_path
  )
)

period_levels <- unname(atlas_lookup)
species_change_levels <- c("Decrease", "Stable", "Increase")
species_change_threshold <- 0.10

richness_palette <- rev(grDevices::rainbow(14, start = 0, end = 0.78))
richness_quantile_probs <- c(0, 0.10, 0.25, 0.50, 0.75, 0.90, 1)
class_colours <- c(
  "Decrease" = "#B2182B",
  "Stable" = "grey72",
  "Increase" = "#1A9850"
)

mainland_bbox <- c(
  xmin = 400000,
  xmax = 750000,
  ymin = 6000000,
  ymax = 6450000
)

bornholm_bbox <- c(
  xmin = 855000,
  xmax = 905000,
  ymin = 6100000,
  ymax = 6160000
)

mainland_width <- mainland_bbox["xmax"] - mainland_bbox["xmin"]
mainland_height <- mainland_bbox["ymax"] - mainland_bbox["ymin"]
bornholm_width <- bornholm_bbox["xmax"] - bornholm_bbox["xmin"]
bornholm_height <- bornholm_bbox["ymax"] - bornholm_bbox["ymin"]

bornholm_inset_width <- bornholm_width / mainland_width
bornholm_inset_height <- bornholm_height / mainland_height
```

```{r helpers}
survey_ids <- function(sites, atlas) {
  paste(sites, atlas, sep = "_")
}

percent_label <- function(x, accuracy = 1) {
  paste0(round(100 * x, accuracy), "%")
}

plot_sf_panel_with_bornholm <- function(panel_data, fill_col, fill_scale,
                                        panel_title) {
  # Match the publication map layout: mainland is the main map and Bornholm is
  # shown as a small bordered inset using the same fill scale.
  p_main <- ggplot(panel_data) +
    geom_sf(
      aes(fill = .data[[fill_col]]),
      colour = "white",
      linewidth = 0.04
    ) +
    fill_scale() +
    coord_sf(
      xlim = c(mainland_bbox["xmin"], mainland_bbox["xmax"]),
      ylim = c(mainland_bbox["ymin"], mainland_bbox["ymax"]),
      expand = FALSE
    ) +
    labs(title = panel_title) +
    theme_void(base_size = 11) +
    theme(
      legend.position = "none",
      plot.title = element_text(
        hjust = 0.5,
        face = "bold",
        colour = "grey15",
        size = 12
      ),
      plot.background = element_rect(fill = "white", colour = NA)
    )

  p_inset <- ggplot(panel_data) +
    geom_sf(
      aes(fill = .data[[fill_col]]),
      colour = "white",
      linewidth = 0.04
    ) +
    fill_scale() +
    coord_sf(
      xlim = c(bornholm_bbox["xmin"], bornholm_bbox["xmax"]),
      ylim = c(bornholm_bbox["ymin"], bornholm_bbox["ymax"]),
      expand = FALSE
    ) +
    theme_void(base_size = 9) +
    theme(
      legend.position = "none",
      plot.background = element_rect(fill = "white", colour = NA),
      panel.border = element_rect(colour = "grey35", fill = NA, linewidth = 0.5)
    )

  ggdraw(p_main) +
    draw_plot(
      p_inset,
      x = 1 - bornholm_inset_width - 0.2,
      y = 1 - bornholm_inset_height - 0.2,
      width = bornholm_inset_width,
      height = bornholm_inset_height
    )
}

plot_sf_series_with_bornholm <- function(plot_data, panel_col, fill_col,
                                         fill_scale, plot_title,
                                         plot_subtitle = "") {
  if (is.factor(plot_data[[panel_col]])) {
    panel_values <- levels(plot_data[[panel_col]])
    panel_values <- panel_values[panel_values %in% as.character(plot_data[[panel_col]])]
  } else {
    panel_values <- unique(plot_data[[panel_col]])
  }
  panel_values <- panel_values[!is.na(panel_values)]

  panel_plots <- lapply(panel_values, function(panel_value) {
    panel_data <- plot_data[plot_data[[panel_col]] == panel_value, ]
    plot_sf_panel_with_bornholm(
      panel_data = panel_data,
      fill_col = fill_col,
      fill_scale = fill_scale,
      panel_title = as.character(panel_value)
    )
  })

  legend_plot <- ggplot(plot_data) +
    geom_sf(aes(fill = .data[[fill_col]]), colour = "white", linewidth = 0.04) +
    fill_scale() +
    theme_void(base_size = 10) +
    theme(
      legend.position = "bottom",
      legend.key.width = unit(1.7, "cm")
    )

  title_block <- ggdraw() +
    draw_label(
      plot_title,
      x = 0.02,
      y = ifelse(plot_subtitle == "", 0.5, 0.72),
      hjust = 0,
      fontface = "bold",
      size = 13,
      colour = "grey15"
    ) +
    draw_label(
      plot_subtitle,
      x = 0.02,
      y = 0.32,
      hjust = 0,
      size = 9.5,
      colour = "grey30"
    )

  plot_grid(
    title_block,
    plot_grid(plotlist = panel_plots, nrow = 1),
    get_legend(legend_plot),
    ncol = 1,
    rel_heights = c(0.12, 0.82, 0.11)
  )
}

load_model_input_definition <- function(coverage_definition) {
  model_env <- new.env(parent = emptyenv())
  load(coverage_definition$model_input_path, envir = model_env)

  if (!all(c("Y", "Design", "Tr") %in% ls(model_env))) {
    stop(
      "Expected Y, Design, and Tr in ",
      coverage_definition$model_input_path,
      call. = FALSE
    )
  }

  if (!setequal(rownames(model_env$Y), rownames(model_env$Design))) {
    stop(
      "Y and Design do not contain the same survey rows in ",
      coverage_definition$model_input_path,
      call. = FALSE
    )
  }

  if (!setequal(colnames(model_env$Y), rownames(model_env$Tr))) {
    stop(
      "Y species columns and Tr rows do not contain the same species in ",
      coverage_definition$model_input_path,
      call. = FALSE
    )
  }

  design <- model_env$Design[rownames(model_env$Y), , drop = FALSE]

  list(
    coverage_id = coverage_definition$coverage_id,
    coverage_label = coverage_definition$coverage_label,
    surveys = rownames(model_env$Y),
    species = colnames(model_env$Y),
    design = data.frame(
      survey = rownames(model_env$Y),
      site = as.character(design$site),
      atlas = as.character(design$atlas),
      stringsAsFactors = FALSE
    )
  )
}

build_richness_frame <- function(y, sites, species, atlas, species_pool_size) {
  y_atlas <- y[
    survey_ids(sites, atlas),
    species,
    drop = FALSE
  ]

  data.frame(
    site = sites,
    survey = survey_ids(sites, atlas),
    atlas = atlas,
    period = unname(atlas_lookup[[atlas]]),
    species_richness = rowSums(y_atlas, na.rm = TRUE),
    retained_species_pool_size = species_pool_size,
    stringsAsFactors = FALSE
  )
}

build_comparison_delta <- function(comparison, richness_df) {
  start_period <- unname(atlas_lookup[[comparison$start_atlas]])
  end_period <- unname(atlas_lookup[[comparison$end_atlas]])

  start_df <- richness_df |>
    filter(.data$atlas == .env$comparison$start_atlas) |>
    arrange(.data$site)

  end_df <- richness_df |>
    filter(.data$atlas == .env$comparison$end_atlas) |>
    arrange(.data$site)

  if (!identical(start_df$site, end_df$site)) {
    stop("Site order mismatch for ", comparison$comparison_label, call. = FALSE)
  }

  data.frame(
    comparison_id = comparison$comparison_id,
    comparison_label = comparison$comparison_label,
    start_period = start_period,
    end_period = end_period,
    site = start_df$site,
    start_richness = start_df$species_richness,
    end_richness = end_df$species_richness,
    delta_richness = end_df$species_richness - start_df$species_richness,
    stringsAsFactors = FALSE
  )
}

build_species_change <- function(comparison, y, shared_sites, retained_species,
                                 species_counts_by_atlas, traits) {
  start_period <- unname(atlas_lookup[[comparison$start_atlas]])
  end_period <- unname(atlas_lookup[[comparison$end_atlas]])

  start_y <- y[
    survey_ids(shared_sites, comparison$start_atlas),
    retained_species,
    drop = FALSE
  ]

  end_y <- y[
    survey_ids(shared_sites, comparison$end_atlas),
    retained_species,
    drop = FALSE
  ]

  data.frame(
    comparison_id = comparison$comparison_id,
    comparison_label = comparison$comparison_label,
    start_period = start_period,
    end_period = end_period,
    species = retained_species,
    occupied_grid_cells_start = as.integer(colSums(start_y > 0, na.rm = TRUE)),
    occupied_grid_cells_end = as.integer(colSums(end_y > 0, na.rm = TRUE)),
    n_shared_grid_cells = length(shared_sites),
    records_1970s = as.integer(species_counts_by_atlas[retained_species, "1"]),
    records_1990s = as.integer(species_counts_by_atlas[retained_species, "2"]),
    records_2010s = as.integer(species_counts_by_atlas[retained_species, "3"]),
    stringsAsFactors = FALSE
  ) |>
    mutate(
      prop_occupied_start = .data$occupied_grid_cells_start / .data$n_shared_grid_cells,
      prop_occupied_end = .data$occupied_grid_cells_end / .data$n_shared_grid_cells,
      delta_occupied_grid_cells = .data$occupied_grid_cells_end - .data$occupied_grid_cells_start,
      delta_occupancy_proportion = .data$prop_occupied_end - .data$prop_occupied_start,
      delta_occupancy_percentage_points = 100 * .data$delta_occupancy_proportion,
      change_class = case_when(
        .data$delta_occupancy_proportion >= species_change_threshold ~ "Increase",
        .data$delta_occupancy_proportion <= -species_change_threshold ~ "Decrease",
        TRUE ~ "Stable"
      ),
      change_class = factor(.data$change_class, levels = species_change_levels)
    ) |>
    left_join(traits, by = "species") |>
    mutate(
      foraging_guild_consensus = ifelse(
        is.na(.data$foraging_guild_consensus) | .data$foraging_guild_consensus == "",
        "Unknown",
        .data$foraging_guild_consensus
      )
    ) |>
    arrange(
      .data$comparison_id,
      .data$change_class,
      desc(abs(.data$delta_occupancy_percentage_points)),
      .data$species
    )
}

build_coverage_analysis <- function(coverage_definition, raw_y, traits, grid_thresholds) {
  definition <- load_model_input_definition(coverage_definition)

  missing_surveys <- setdiff(definition$surveys, rownames(raw_y))
  if (length(missing_surveys) > 0) {
    stop(
      "Some retained surveys are absent from the raw atlas occurrence matrix: ",
      paste(head(missing_surveys, 10), collapse = ", "),
      call. = FALSE
    )
  }

  missing_species <- setdiff(definition$species, colnames(raw_y))
  if (length(missing_species) > 0) {
    stop(
      "Some retained species are absent from the raw atlas occurrence matrix: ",
      paste(head(missing_species, 10), collapse = ", "),
      call. = FALSE
    )
  }

  # This is the key step for keeping the report empirical: counts are
  # recomputed from Y_occurrences.csv after applying the model-input survey and
  # species definitions.
  empirical_y <- raw_y[definition$surveys, definition$species, drop = FALSE]

  all_richness <- definition$design |>
    mutate(
      period = unname(atlas_lookup[.data$atlas]),
      species_richness = rowSums(empirical_y, na.rm = TRUE),
      retained_species_pool_size = length(definition$species),
      period = factor(.data$period, levels = period_levels)
    ) |>
    arrange(.data$atlas, .data$site)

  sites_by_atlas <- lapply(names(atlas_lookup), function(atlas) {
    all_richness |>
      filter(.data$atlas == .env$atlas) |>
      pull(.data$site)
  })
  names(sites_by_atlas) <- names(atlas_lookup)

  grid_sites <- as.character(grid_thresholds$kvdrtkd)
  shared_sites <- grid_sites[grid_sites %in% Reduce(intersect, sites_by_atlas)]

  shared_surveys <- unlist(
    lapply(names(atlas_lookup), function(atlas) survey_ids(shared_sites, atlas)),
    use.names = FALSE
  )

  missing_shared_surveys <- setdiff(shared_surveys, rownames(empirical_y))
  if (length(missing_shared_surveys) > 0) {
    stop(
      "Some shared-site surveys are absent after coverage filtering: ",
      paste(head(missing_shared_surveys, 10), collapse = ", "),
      call. = FALSE
    )
  }

  shared_richness <- bind_rows(
    lapply(names(atlas_lookup), function(atlas) {
      build_richness_frame(
        y = empirical_y,
        sites = shared_sites,
        species = definition$species,
        atlas = atlas,
        species_pool_size = length(definition$species)
      )
    })
  ) |>
    mutate(period = factor(.data$period, levels = period_levels)) |>
    arrange(.data$atlas, match(.data$site, shared_sites))

  delta_df <- bind_rows(
    lapply(seq_len(nrow(comparisons)), function(i) {
      build_comparison_delta(comparisons[i, ], shared_richness)
    })
  ) |>
    mutate(comparison_label = factor(.data$comparison_label, levels = comparisons$comparison_label))

  species_counts_by_atlas <- sapply(
    names(atlas_lookup),
    function(atlas) {
      colSums(
        empirical_y[survey_ids(shared_sites, atlas), definition$species, drop = FALSE],
        na.rm = TRUE
      )
    }
  )

  species_occupancy <- bind_rows(
    lapply(seq_len(nrow(comparisons)), function(i) {
      build_species_change(
        comparison = comparisons[i, ],
        y = empirical_y,
        shared_sites = shared_sites,
        retained_species = definition$species,
        species_counts_by_atlas = species_counts_by_atlas,
        traits = traits
      )
    })
  )

  species_top_change <- species_occupancy |>
    filter(.data$change_class != "Stable") |>
    arrange(.data$comparison_id, desc(abs(.data$delta_occupancy_percentage_points)))

  species_class_proportions <- data.frame(
    comparison_id = rep(comparisons$comparison_id, each = length(species_change_levels)),
    change_class = rep(species_change_levels, times = nrow(comparisons)),
    stringsAsFactors = FALSE
  ) |>
    left_join(
      species_occupancy |>
        count(.data$comparison_id, .data$change_class, name = "n_species") |>
        mutate(change_class = as.character(.data$change_class)),
      by = c("comparison_id", "change_class")
    ) |>
    left_join(
      comparisons |>
        select("comparison_id", "comparison_label"),
      by = "comparison_id"
    ) |>
    group_by(.data$comparison_id, .data$comparison_label) |>
    mutate(
      n_species = ifelse(is.na(.data$n_species), 0L, .data$n_species),
      total_species = sum(.data$n_species),
      proportion_species = .data$n_species / .data$total_species,
      change_class = factor(.data$change_class, levels = species_change_levels)
    ) |>
    ungroup()

  guild_class_grid <- expand.grid(
    comparison_id = comparisons$comparison_id,
    foraging_guild_consensus = sort(unique(species_occupancy$foraging_guild_consensus)),
    change_class = species_change_levels,
    stringsAsFactors = FALSE
  )

  guild_class_proportions <- guild_class_grid |>
    left_join(
      species_occupancy |>
        count(
          .data$comparison_id,
          .data$foraging_guild_consensus,
          .data$change_class,
          name = "n_species"
        ) |>
        mutate(change_class = as.character(.data$change_class)),
      by = c("comparison_id", "foraging_guild_consensus", "change_class")
    ) |>
    left_join(
      comparisons |>
        select("comparison_id", "comparison_label"),
      by = "comparison_id"
    ) |>
    mutate(n_species = ifelse(is.na(.data$n_species), 0L, .data$n_species)) |>
    group_by(.data$comparison_id, .data$comparison_label, .data$foraging_guild_consensus) |>
    mutate(
      guild_species_count = sum(.data$n_species),
      proportion_within_guild = .data$n_species / .data$guild_species_count
    ) |>
    ungroup() |>
    mutate(change_class = factor(.data$change_class, levels = species_change_levels))

  guild_order <- guild_class_proportions |>
    group_by(.data$foraging_guild_consensus) |>
    summarise(
      mean_increase = mean(
        ifelse(.data$change_class == "Increase", .data$proportion_within_guild, NA),
        na.rm = TRUE
      ),
      mean_decrease = mean(
        ifelse(.data$change_class == "Decrease", .data$proportion_within_guild, NA),
        na.rm = TRUE
      ),
      net_change_signal = .data$mean_increase - .data$mean_decrease,
      .groups = "drop"
    ) |>
    arrange(.data$net_change_signal, .data$foraging_guild_consensus) |>
    pull(.data$foraging_guild_consensus)

  guild_class_proportions <- guild_class_proportions |>
    mutate(foraging_guild_consensus = factor(.data$foraging_guild_consensus, levels = guild_order))

  guild_label_order <- guild_class_proportions |>
    distinct(.data$foraging_guild_consensus, .data$guild_species_count) |>
    arrange(.data$foraging_guild_consensus) |>
    mutate(
      foraging_guild_label = paste0(
        as.character(.data$foraging_guild_consensus),
        " (n=",
        .data$guild_species_count,
        ")"
      )
    )

  guild_class_proportions <- guild_class_proportions |>
    left_join(
      guild_label_order,
      by = c("foraging_guild_consensus", "guild_species_count")
    ) |>
    mutate(
      foraging_guild_label = factor(
        .data$foraging_guild_label,
        levels = guild_label_order$foraging_guild_label
      )
    )

  guild_sum_check <- guild_class_proportions |>
    group_by(.data$comparison_id, .data$foraging_guild_consensus) |>
    summarise(total = sum(.data$proportion_within_guild), .groups = "drop")

  if (any(abs(guild_sum_check$total - 1) > 1e-8)) {
    stop("At least one guild-normalized class proportion does not sum to 1.", call. = FALSE)
  }

  list(
    coverage_id = definition$coverage_id,
    coverage_label = definition$coverage_label,
    retained_species = definition$species,
    all_richness = all_richness,
    shared_sites = shared_sites,
    shared_richness = shared_richness,
    delta_df = delta_df,
    species_occupancy = species_occupancy,
    species_top_change = species_top_change,
    species_class_proportions = species_class_proportions,
    guild_class_proportions = guild_class_proportions
  )
}

write_analysis_tables <- function(analysis) {
  prefix <- file.path(
    out_dir,
    paste0("empirical-atlas-", analysis$coverage_id)
  )

  write.csv(
    analysis$all_richness,
    paste0(prefix, "-richness-all-sites.csv"),
    row.names = FALSE,
    na = ""
  )
  write.csv(
    analysis$shared_richness,
    paste0(prefix, "-richness-shared-sites.csv"),
    row.names = FALSE,
    na = ""
  )
  write.csv(
    analysis$delta_df,
    paste0(prefix, "-delta-richness-shared-sites.csv"),
    row.names = FALSE,
    na = ""
  )
  write.csv(
    analysis$species_occupancy,
    paste0(prefix, "-species-occupancy-change.csv"),
    row.names = FALSE,
    na = ""
  )
  write.csv(
    analysis$species_top_change,
    paste0(prefix, "-species-top-gains-losses.csv"),
    row.names = FALSE,
    na = ""
  )
  write.csv(
    analysis$species_class_proportions,
    paste0(prefix, "-species-change-class-proportions.csv"),
    row.names = FALSE,
    na = ""
  )
  write.csv(
    analysis$guild_class_proportions,
    paste0(prefix, "-guild-normalized-change-class-proportions.csv"),
    row.names = FALSE,
    na = ""
  )
}

build_richness_scale <- function(plot_data, reference_period = "1970s") {
  # Allocate the colour gradient according to the reference-period
  # distribution, then preserve that exact mapping for every later period.
  reference_values <- plot_data$species_richness[
    as.character(plot_data$period) == reference_period
  ]
  all_values <- plot_data$species_richness

  reference_breaks <- reference_values |>
    quantile(
      probs = richness_quantile_probs,
      na.rm = TRUE,
      names = FALSE
    ) |>
    round() |>
    unique() |>
    sort()

  scale_limits <- range(all_values, na.rm = TRUE)
  colour_anchors <- sort(unique(c(scale_limits, reference_breaks)))
  anchor_colours <- grDevices::colorRampPalette(richness_palette)(
    length(colour_anchors)
  )

  scale_fill_gradientn(
    colours = anchor_colours,
    values = scales::rescale(colour_anchors, from = scale_limits),
    limits = scale_limits,
    breaks = reference_breaks,
    labels = reference_breaks,
    oob = scales::squish,
    name = paste0("Observed richness\n(", reference_period, " quantiles)")
  )
}

plot_empirical_richness <- function(plot_data, title, subtitle) {
  plot_sf_series_with_bornholm(
    plot_data = plot_data,
    panel_col = "period",
    fill_col = "species_richness",
    fill_scale = function() {
      build_richness_scale(plot_data)
    },
    plot_title = title,
    plot_subtitle = subtitle
  )
}

plot_delta_richness <- function(plot_data, title, subtitle, delta_limit) {
  plot_sf_series_with_bornholm(
    plot_data = plot_data,
    panel_col = "comparison_label",
    fill_col = "delta_richness",
    fill_scale = function() {
      scale_fill_gradient2(
        low = "#2C7BB6",
        mid = "white",
        high = "#D7191C",
        midpoint = 0,
        limits = c(-delta_limit, delta_limit),
        name = "Delta\nrichness"
      )
    },
    plot_title = title,
    plot_subtitle = subtitle
  )
}

plot_species_class_proportions <- function(analysis) {
  ggplot(
    analysis$species_class_proportions,
    aes(x = .data$change_class, y = .data$proportion_species, fill = .data$change_class)
  ) +
    geom_col(width = 0.68, colour = "grey25", linewidth = 0.25) +
    geom_text(
      aes(label = paste0(.data$n_species, "\n", percent_label(.data$proportion_species))),
      vjust = -0.25,
      size = 3.3,
      lineheight = 0.9
    ) +
    facet_wrap(vars(.data$comparison_label), nrow = 1) +
    scale_fill_manual(values = class_colours, guide = "none") +
    scale_y_continuous(
      labels = percent_label,
      limits = c(0, max(analysis$species_class_proportions$proportion_species) * 1.20),
      expand = c(0, 0)
    ) +
    labs(
      x = NULL,
      y = "Proportion of retained species",
      title = paste0(
        "Species occupancy-change classes: ",
        analysis$coverage_label,
        " coverage"
      )
    ) +
    theme_minimal(base_size = 11) +
    theme(
      panel.grid.major.x = element_blank(),
      panel.grid.minor = element_blank(),
      plot.title = element_text(face = "bold"),
      strip.text = element_text(face = "bold"),
      axis.text.x = element_text(colour = "grey20")
    )
}

plot_species_top_changes <- function(analysis) {
  species_top_change_plot_df <- analysis$species_occupancy |>
    filter(.data$change_class != "Stable") |>
    group_by(.data$comparison_label, .data$change_class) |>
    arrange(desc(abs(.data$delta_occupancy_percentage_points)), .by_group = TRUE) |>
    slice_head(n = 12) |>
    ungroup() |>
    group_by(.data$comparison_label) |>
    arrange(.data$delta_occupancy_percentage_points, .by_group = TRUE) |>
    mutate(
      species_label = .data$species_english,
      species_panel_label = paste(.data$comparison_label, .data$species_label, sep = " | ")
    ) |>
    ungroup()

  species_top_change_plot_df$species_panel_label <- factor(
    species_top_change_plot_df$species_panel_label,
    levels = species_top_change_plot_df$species_panel_label
  )

  ggplot(
    species_top_change_plot_df,
    aes(
      x = .data$delta_occupancy_percentage_points,
      y = .data$species_panel_label,
      fill = .data$change_class
    )
  ) +
    geom_vline(xintercept = 0, colour = "grey45", linewidth = 0.4) +
    geom_col(width = 0.72) +
    facet_wrap(vars(.data$comparison_label), scales = "free_y", nrow = 1) +
    scale_fill_manual(values = class_colours, name = NULL) +
    scale_x_continuous(labels = function(x) paste0(x, "%")) +
    scale_y_discrete(labels = function(x) sub("^.* \\| ", "", x)) +
    labs(
      x = "Change in occupied shared sites",
      y = NULL,
      title = paste0(
        "Largest species-level occupancy changes: ",
        analysis$coverage_label,
        " coverage"
      )
    ) +
    theme_minimal(base_size = 10.2) +
    theme(
      panel.grid.major.y = element_blank(),
      panel.grid.minor = element_blank(),
      plot.title = element_text(face = "bold"),
      strip.text = element_text(face = "bold"),
      legend.position = "bottom",
      axis.text.y = element_text(colour = "grey20", size = 7.2)
    )
}

plot_guild_class_proportions <- function(analysis) {
  ggplot(
    analysis$guild_class_proportions,
    aes(
      x = .data$foraging_guild_label,
      y = .data$proportion_within_guild,
      fill = .data$change_class
    )
  ) +
    geom_col(width = 0.72, colour = "white", linewidth = 0.15) +
    coord_flip() +
    facet_wrap(vars(.data$comparison_label), nrow = 1) +
    scale_fill_manual(values = class_colours, name = NULL) +
    scale_y_continuous(labels = percent_label, expand = expansion(mult = c(0, 0.02))) +
    labs(
      x = NULL,
      y = "Within-guild proportion of retained species",
      title = paste0(
        "Occupancy-change classes by foraging guild: ",
        analysis$coverage_label,
        " coverage"
      )
    ) +
    theme_minimal(base_size = 10.2) +
    theme(
      panel.grid.major.y = element_blank(),
      panel.grid.minor = element_blank(),
      panel.spacing.x = unit(1.2, "cm"),
      plot.title = element_text(face = "bold"),
      strip.text = element_text(face = "bold"),
      legend.position = "bottom",
      axis.text.y = element_text(colour = "grey20", size = 7.2),
      plot.margin = margin(10, 28, 10, 10)
    )
}

build_plot_data <- function(analysis, grid_thresholds) {
  all_richness_sf <- grid_thresholds |>
    select("kvdrtkd") |>
    inner_join(analysis$all_richness, by = c("kvdrtkd" = "site")) |>
    mutate(period = factor(.data$period, levels = period_levels))

  shared_grid_sf <- grid_thresholds |>
    select("kvdrtkd") |>
    filter(.data$kvdrtkd %in% analysis$shared_sites) |>
    mutate(kvdrtkd = factor(.data$kvdrtkd, levels = analysis$shared_sites)) |>
    arrange(.data$kvdrtkd)

  if (!identical(as.character(shared_grid_sf$kvdrtkd), analysis$shared_sites)) {
    stop("Shared grid polygons are not ordered like the shared site vector.", call. = FALSE)
  }

  shared_richness_sf <- shared_grid_sf |>
    inner_join(analysis$shared_richness, by = c("kvdrtkd" = "site"))

  delta_sf <- shared_grid_sf |>
    inner_join(analysis$delta_df, by = c("kvdrtkd" = "site")) |>
    mutate(comparison_label = factor(.data$comparison_label, levels = comparisons$comparison_label))

  list(
    all_richness_sf = all_richness_sf,
    shared_richness_sf = shared_richness_sf,
    delta_sf = delta_sf
  )
}

save_analysis_plots <- function(analysis, plot_data, delta_limit) {
  plot_prefix <- file.path(
    plot_dir,
    paste0("empirical-atlas-", analysis$coverage_id)
  )

  all_sites_map <- plot_empirical_richness(
    plot_data = plot_data$all_richness_sf,
    title = paste0("Empirical atlas richness: ", analysis$coverage_label, " sites"),
    subtitle = paste(
      "Observed atlas presences from the raw occurrence matrix;",
      "colour spacing follows 1970s richness quantiles in every panel."
    )
  )

  shared_sites_map <- plot_empirical_richness(
    plot_data = plot_data$shared_richness_sf,
    title = paste0("Empirical atlas richness: shared ", analysis$coverage_label, " sites"),
    subtitle = paste0(
      length(analysis$shared_sites),
      " grid cells present in all three atlas periods; ",
      "colour spacing follows 1970s richness quantiles in every panel."
    )
  )

  delta_map <- plot_delta_richness(
    plot_data = plot_data$delta_sf,
    title = paste0("Delta empirical richness: shared ", analysis$coverage_label, " sites"),
    subtitle = "Positive values indicate higher observed richness in the later atlas period.",
    delta_limit = delta_limit
  )

  species_class_plot <- plot_species_class_proportions(analysis)
  species_top_change_plot <- plot_species_top_changes(analysis)
  guild_class_plot <- plot_guild_class_proportions(analysis)

  ggsave(
    filename = paste0(plot_prefix, "-richness-all-sites.png"),
    plot = all_sites_map,
    width = 11.5,
    height = 5.4,
    dpi = 300,
    bg = "white"
  )
  ggsave(
    filename = paste0(plot_prefix, "-richness-shared-sites.png"),
    plot = shared_sites_map,
    width = 11.5,
    height = 5.4,
    dpi = 300,
    bg = "white"
  )
  ggsave(
    filename = paste0(plot_prefix, "-delta-richness-shared-sites.png"),
    plot = delta_map,
    width = 10.2,
    height = 5.4,
    dpi = 300,
    bg = "white"
  )
  ggsave(
    filename = paste0(plot_prefix, "-species-change-class-proportions.png"),
    plot = species_class_plot,
    width = 9.4,
    height = 4.8,
    dpi = 300,
    bg = "white"
  )
  ggsave(
    filename = paste0(plot_prefix, "-species-top-gains-losses.png"),
    plot = species_top_change_plot,
    width = 12.5,
    height = 7.4,
    dpi = 300,
    bg = "white"
  )
  ggsave(
    filename = paste0(plot_prefix, "-guild-normalized-change-class-proportions.png"),
    plot = guild_class_plot,
    width = 13.5,
    height = max(
      7.6,
      0.25 * length(unique(analysis$species_occupancy$foraging_guild_consensus)) + 2.4
    ),
    dpi = 300,
    bg = "white"
  )

  list(
    all_sites_map = all_sites_map,
    shared_sites_map = shared_sites_map,
    delta_map = delta_map,
    species_class_plot = species_class_plot,
    species_top_change_plot = species_top_change_plot,
    guild_class_plot = guild_class_plot
  )
}
```

```{r load-data}
grid_thresholds <- st_read(grid_threshold_path, quiet = TRUE)

raw_y <- read.csv(
  occurrence_path,
  row.names = 1,
  check.names = FALSE,
  stringsAsFactors = FALSE
)

traits_raw <- read.csv(
  traits_path,
  row.names = 2,
  check.names = FALSE,
  stringsAsFactors = FALSE
)

species_common_names <- read.csv(
  traits_common_name_path,
  check.names = FALSE,
  stringsAsFactors = FALSE
) |>
  transmute(
    species = gsub(" ", "_", .data$latin_DOF),
    species_english = .data$english_DOF
  ) |>
  filter(!is.na(.data$species), .data$species != "") |>
  distinct(.data$species, .keep_all = TRUE)

sti_common_names <- read.csv(
  sti_common_name_path,
  check.names = FALSE,
  stringsAsFactors = FALSE
) |>
  transmute(
    species = gsub(" ", "_", .data$SPECIES),
    species_english_sti = .data[["ENGL NAME"]]
  ) |>
  filter(!is.na(.data$species), .data$species != "") |>
  distinct(.data$species, .keep_all = TRUE)

traits <- data.frame(
  species = rownames(traits_raw),
  foraging_guild_consensus = traits_raw$foraging_guild_consensus,
  stringsAsFactors = FALSE
) |>
  left_join(species_common_names, by = "species") |>
  left_join(sti_common_names, by = "species") |>
  mutate(
    species_english = coalesce(.data$species_english, .data$species_english_sti),
    species_english = case_when(
      .data$species == "Acanthis_flammea" ~ "Lesser Redpoll",
      .data$species == "Corvus_corax" ~ "Common Raven",
      TRUE ~ .data$species_english
    ),
    species_english = ifelse(
      is.na(.data$species_english) | .data$species_english == "",
      gsub("_", " ", .data$species),
      .data$species_english
    )
  ) |>
  select(-"species_english_sti")
```

```{r build-analyses}
analyses <- lapply(
  coverage_definitions,
  build_coverage_analysis,
  raw_y = raw_y,
  traits = traits,
  grid_thresholds = grid_thresholds
)

plot_data <- lapply(analyses, build_plot_data, grid_thresholds = grid_thresholds)

delta_limit <- max(
  abs(unlist(lapply(plot_data, function(x) x$delta_sf$delta_richness))),
  na.rm = TRUE
)

invisible(lapply(analyses, write_analysis_tables))

plots <- Map(
  save_analysis_plots,
  analysis = analyses,
  plot_data = plot_data,
  MoreArgs = list(
    delta_limit = delta_limit
  )
)
```

## Good/Fair Coverage

This section uses the `Good/Fair` retained survey rows and species pool from
the current model input. Richness and turnover values are recalculated from the
raw atlas occurrence matrix.

For Atlas 3, this section includes all available Atlas 3 sites because there is
no Atlas 3 sampling-effort table to separate `Good` from `Fair` coverage.

### Richness Across Good/Fair Sites

```{r good-average-all-sites-map-display, out.width="100%"}
plots$good_average$all_sites_map
```

### Richness On Shared Good/Fair Sites

```{r good-average-shared-sites-map-display, out.width="100%"}
plots$good_average$shared_sites_map
```

### Delta Richness On Shared Good/Fair Sites

```{r good-average-delta-map-display, out.width="100%"}
plots$good_average$delta_map
```

### Species-Level Changes

For this `Good/Fair` analysis, the 10 percentage-point rule corresponds to
about `r ceiling(species_change_threshold * length(analyses$good_average$shared_sites))`
shared grid cells.

```{r good-average-species-class-display, out.width="100%"}
plots$good_average$species_class_plot
```

```{r good-average-species-top-display, out.width="100%"}
plots$good_average$species_top_change_plot
```

### Foraging-Guild Changes

```{r good-average-guild-display, out.width="100%"}
plots$good_average$guild_class_plot
```

## Good Coverage Only

This section repeats the same empirical atlas analysis using only the `Good`
coverage retained survey rows and species pool. This is the stricter coverage
check against the `Good/Fair` patterns above.

As above, the `Good` coverage filter applies to Atlas 1 and Atlas 2 only. Atlas
3 includes all available Atlas 3 sites because sampling-effort estimates are not
available for that period.

### Richness Across Good Sites

```{r good-all-sites-map-display, out.width="100%"}
plots$good$all_sites_map
```

### Richness On Shared Good Sites

```{r good-shared-sites-map-display, out.width="100%"}
plots$good$shared_sites_map
```

### Delta Richness On Shared Good Sites

```{r good-delta-map-display, out.width="100%"}
plots$good$delta_map
```

### Species-Level Changes

For this `Good` analysis, the 10 percentage-point rule corresponds to about
`r ceiling(species_change_threshold * length(analyses$good$shared_sites))`
shared grid cells.

```{r good-species-class-display, out.width="100%"}
plots$good$species_class_plot
```

```{r good-species-top-display, out.width="100%"}
plots$good$species_top_change_plot
```

### Foraging-Guild Changes

```{r good-guild-display, out.width="100%"}
plots$good$guild_class_plot
```

```{r output-checks}
non_png_plot_outputs <- list.files(
  plot_dir,
  pattern = "\\.(pdf|svg|jpg|jpeg|tif|tiff)$",
  ignore.case = TRUE,
  full.names = TRUE
)

if (length(non_png_plot_outputs) > 0) {
  stop(
    "Unexpected non-PNG plot output(s): ",
    paste(non_png_plot_outputs, collapse = ", "),
    call. = FALSE
  )
}
```
