---
title: "Species association Omega exploration"
author: "WM"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    df_print: paged
---

```{r setup, include=FALSE}
# This notebook is exploratory, but the outputs are intended to be reusable.
# All rendered and manually saved plots are PNGs, following the project rule.
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  dev = "png",
  dpi = 300
)
project_root <- normalizePath(file.path("..", ".."))
knitr::opts_knit$set(root.dir = project_root)

required_packages <- c(
  "tidyverse", "readxl", "knitr", "patchwork", "viridis", "scales"
)
missing_packages <- required_packages[
  !vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)
]
if (length(missing_packages) > 0) {
  stop(
    "Install required packages before rendering: ",
    paste(missing_packages, collapse = ", "),
    call. = FALSE
  )
}

suppressPackageStartupMessages({
  library(tidyverse)
  library(readxl)
  library(knitr)
  library(patchwork)
  library(viridis)
  library(scales)
})

old_wd <- setwd(project_root)
source(file.path("support_scripts", "figure_data_helpers.R"))
setwd(old_wd)

theme_set(theme_minimal(base_size = 11))
```

## Aim

This notebook explores residual species associations from the HMSC `Omega`
estimates. The values are not raw co-occurrence counts. They are residual
associations after the fitted environmental, trait, phylogenetic, and spatial
structure in each atlas-period model.

The main questions are:

- Which species pairs show supported positive or negative residual association?
- Do those pairwise association patterns change across atlas periods?
- Do foraging guilds show stronger or weaker residual association than expected
  from the atlas-wide average, or from a guild-label permutation null?

## Load Omega Estimates

```{r configuration}
# The March 2026 models are the current publication-facing atlas comparison
# model family documented in the project README.
base_dir <- "HmscOutputs"
pattern <- "2026-03-13"
omega_random_level <- "site"
support_level <- 0.95
negative_support_cutoff <- 0.05
expected_species_count <- 157

# This is high enough for a stable exploratory null while still rendering
# comfortably on a laptop. Increase if the guild null becomes manuscript-facing.
n_permutations <- 999
set.seed(20260518)

period_order <- c("1970s", "1990s", "2010s")
year_lookup <- c("1" = "1970s", "2" = "1990s", "3" = "2010s")
support_levels <- c("Supported positive", "Supported negative", "Unsupported")
min_pairs_for_stable_guild_summary <- 10

out_dir <- file.path(
  "notebooks", "exploratory", "outputs", "species-associations-omega"
)
plot_dir <- file.path(out_dir, "plots")
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(plot_dir, recursive = TRUE, showWarnings = FALSE)

matching_folders <- figure_model_folders(pattern = pattern, base_dir = base_dir)
model_metadata <- tibble(
  model_folder = matching_folders,
  atlas = as.character(atlas_numbers(matching_folders)),
  period = unname(year_lookup[atlas])
) |>
  arrange(as.numeric(.data$atlas))

kable(model_metadata, caption = "Model folders used for the Omega exploration.")
```

```{r omega-reader}
read_numeric_matrix_sheet <- function(path, sheet) {
  # The S7 Omega workbooks use a blank first column for row species names.
  # readxl imports that column as ...1, so we rename it immediately.
  df <- readxl::read_excel(path, sheet = sheet)
  names(df)[1] <- "species"

  matrix_values <- as.matrix(df[, -1])
  storage.mode(matrix_values) <- "numeric"

  list(
    species = df$species,
    column_species = colnames(df)[-1],
    matrix = matrix_values
  )
}

read_omega_pairs <- function(model_folder, atlas, period) {
  omega_path <- file.path(
    base_dir, model_folder, "Results",
    paste0("parameter_estimates_Omega_", omega_random_level, ".xlsx")
  )
  if (!file.exists(omega_path)) {
    stop(
      "Missing Omega workbook: ", omega_path,
      "\nRun S7_Show_parameter_estimates.R for this model first.",
      call. = FALSE
    )
  }

  mean_sheet <- read_numeric_matrix_sheet(omega_path, "Posterior mean")
  positive_sheet <- read_numeric_matrix_sheet(omega_path, "Pr(x>0)")
  negative_sheet <- read_numeric_matrix_sheet(omega_path, "Pr(x<0)")

  species_names <- mean_sheet$species
  if (!identical(species_names, mean_sheet$column_species)) {
    stop("Row and column species names do not match in: ", omega_path, call. = FALSE)
  }
  if (!identical(species_names, positive_sheet$species) ||
      !identical(species_names, positive_sheet$column_species) ||
      !identical(species_names, negative_sheet$species) ||
      !identical(species_names, negative_sheet$column_species)) {
    stop("Omega workbook sheets have inconsistent species names: ", omega_path, call. = FALSE)
  }

  pair_index <- which(upper.tri(mean_sheet$matrix), arr.ind = TRUE)

  tibble(
    model_folder = model_folder,
    atlas = atlas,
    period = period,
    species_a = species_names[pair_index[, "row"]],
    species_b = species_names[pair_index[, "col"]],
    omega_mean = as.numeric(mean_sheet$matrix[pair_index]),
    pr_positive = as.numeric(positive_sheet$matrix[pair_index]),
    pr_negative = as.numeric(negative_sheet$matrix[pair_index])
  ) |>
    mutate(
      # Rounding avoids Excel floating-point boundary artifacts where values
      # that print as 0.05 can otherwise behave as just below 0.05.
      pr_positive = round(.data$pr_positive, 10),
      pr_negative = round(.data$pr_negative, 10),
      pair_id = paste(.data$species_a, .data$species_b, sep = "__"),
      support_class = case_when(
        .data$pr_positive > support_level ~ "Supported positive",
        .data$pr_positive < negative_support_cutoff ~ "Supported negative",
        TRUE ~ "Unsupported"
      ),
      support_class = factor(.data$support_class, levels = support_levels),
      is_supported_positive = .data$support_class == "Supported positive",
      is_supported_negative = .data$support_class == "Supported negative",
      period = factor(.data$period, levels = period_order)
    )
}

omega_pairs_raw <- pmap_dfr(
  model_metadata,
  function(model_folder, atlas, period) {
    read_omega_pairs(
      model_folder = model_folder,
      atlas = atlas,
      period = period
    )
  }
)

validation_checks <- omega_pairs_raw |>
  group_by(.data$atlas, .data$period, .data$model_folder) |>
  summarise(
    n_species = n_distinct(c(.data$species_a, .data$species_b)),
    n_pairs = n(),
    expected_pairs = .data$n_species * (.data$n_species - 1) / 2,
    pair_count_ok = .data$n_pairs == .data$expected_pairs,
    .groups = "drop"
  )

stopifnot(
  all(validation_checks$n_species == expected_species_count),
  all(validation_checks$pair_count_ok)
)

kable(
  validation_checks,
  caption = "Omega workbook validation. Each atlas should have 157 species and 12,246 unique non-self pairs."
)
```

## Add Species Traits

```{r trait-joins}
load(file.path("Data", "preprocessed_data_min_occs_is_5_coverage_is_good_and_average.RData"))
if (!exists("Tr")) {
  stop("Expected object `Tr` in the preprocessed data file.", call. = FALSE)
}

species_traits <- Tr |>
  as.data.frame() |>
  rownames_to_column(var = "species") |>
  transmute(
    species = .data$species,
    guild = .data$foraging_guild_consensus,
    migration_group = .data$Migration_a3_DOF,
    species_thermal_index = as.numeric(.data$species_thermal_index)
  )

guild_sizes <- species_traits |>
  count(.data$guild, name = "n_species") |>
  arrange(desc(.data$n_species), .data$guild)

traits_a <- species_traits |>
  rename(
    species_a = species,
    guild_a = guild,
    migration_group_a = migration_group,
    species_thermal_index_a = species_thermal_index
  )
traits_b <- species_traits |>
  rename(
    species_b = species,
    guild_b = guild,
    migration_group_b = migration_group,
    species_thermal_index_b = species_thermal_index
  )

omega_pairs <- omega_pairs_raw |>
  left_join(traits_a, by = "species_a") |>
  left_join(traits_b, by = "species_b") |>
  mutate(
    guild_1 = pmin(.data$guild_a, .data$guild_b),
    guild_2 = pmax(.data$guild_a, .data$guild_b),
    guild_pair = if_else(
      .data$guild_1 == .data$guild_2,
      .data$guild_1,
      paste(.data$guild_1, .data$guild_2, sep = " | ")
    ),
    guild_pair_type = if_else(
      .data$guild_1 == .data$guild_2,
      "Within guild",
      "Between guild"
    )
  )

if (any(is.na(omega_pairs$guild_a)) || any(is.na(omega_pairs$guild_b))) {
  missing_species <- omega_pairs |>
    filter(is.na(.data$guild_a) | is.na(.data$guild_b)) |>
    summarise(species = paste(unique(c(.data$species_a, .data$species_b)), collapse = ", "))
  stop("Some Omega species were not found in the trait table: ", missing_species$species, call. = FALSE)
}

kable(guild_sizes, caption = "Foraging guild sizes in the modelled species set.")
```

## Overall Association Patterns

```{r overall-summaries}
support_summary <- omega_pairs |>
  count(.data$atlas, .data$period, .data$support_class, name = "n_pairs") |>
  complete(
    atlas,
    period,
    support_class = factor(support_levels, levels = support_levels),
    fill = list(n_pairs = 0)
  ) |>
  group_by(.data$atlas, .data$period) |>
  mutate(prop_pairs = .data$n_pairs / sum(.data$n_pairs)) |>
  ungroup()

top_supported_pairs <- omega_pairs |>
  filter(.data$support_class != "Unsupported") |>
  group_by(.data$atlas, .data$period, .data$support_class) |>
  arrange(desc(abs(.data$omega_mean)), .by_group = TRUE) |>
  slice_head(n = 25) |>
  ungroup() |>
  arrange(.data$atlas, .data$support_class, desc(abs(.data$omega_mean)))

temporal_change <- omega_pairs |>
  select(
    pair_id, species_a, species_b,
    guild_a, guild_b, guild_1, guild_2,
    guild_pair, guild_pair_type,
    period, omega_mean, pr_positive,
    pr_negative, support_class
  ) |>
  pivot_wider(
    names_from = .data$period,
    values_from = c(
      omega_mean, pr_positive, pr_negative,
      support_class
    ),
    names_glue = "{.value}_{period}"
  ) |>
  mutate(
    delta_1990s_minus_1970s = .data$omega_mean_1990s - .data$omega_mean_1970s,
    delta_2010s_minus_1970s = .data$omega_mean_2010s - .data$omega_mean_1970s,
    abs_delta_2010s_minus_1970s = abs(.data$delta_2010s_minus_1970s),
    raw_sign_flip_1970s_to_2010s =
      sign(.data$omega_mean_1970s) != sign(.data$omega_mean_2010s),
    support_transition_1970s_to_2010s = paste(
      .data$support_class_1970s, "->", .data$support_class_2010s
    ),
    newly_supported_2010s =
      .data$support_class_1970s == "Unsupported" &
      .data$support_class_2010s != "Unsupported",
    lost_supported_2010s =
      .data$support_class_1970s != "Unsupported" &
      .data$support_class_2010s == "Unsupported",
    supported_positive_to_negative_2010s =
      .data$support_class_1970s == "Supported positive" &
      .data$support_class_2010s == "Supported negative",
    supported_negative_to_positive_2010s =
      .data$support_class_1970s == "Supported negative" &
      .data$support_class_2010s == "Supported positive"
  ) |>
  arrange(desc(.data$abs_delta_2010s_minus_1970s))

support_transitions <- temporal_change |>
  count(
    .data$support_transition_1970s_to_2010s,
    name = "n_pairs"
  ) |>
  mutate(prop_pairs = .data$n_pairs / sum(.data$n_pairs)) |>
  arrange(desc(.data$n_pairs))

kable(support_summary, digits = 3, caption = "Supported Omega association classes by atlas period.")
```

```{r write-main-csvs}
write_csv(
  omega_pairs,
  file.path(out_dir, paste0(pattern, "-omega-species-pairs.csv"))
)
write_csv(
  top_supported_pairs,
  file.path(out_dir, paste0(pattern, "-omega-top-supported-pairs.csv"))
)
write_csv(
  temporal_change,
  file.path(out_dir, paste0(pattern, "-omega-temporal-change.csv"))
)
write_csv(
  support_transitions,
  file.path(out_dir, paste0(pattern, "-omega-support-transitions.csv"))
)
```

## Guild-Level Observed Summaries

```{r guild-observed-summary}
atlas_averages <- omega_pairs |>
  group_by(.data$atlas, .data$period) |>
  summarise(
    atlas_mean_omega = mean(.data$omega_mean, na.rm = TRUE),
    atlas_median_omega = median(.data$omega_mean, na.rm = TRUE),
    atlas_prop_supported_positive = mean(.data$is_supported_positive, na.rm = TRUE),
    atlas_prop_supported_negative = mean(.data$is_supported_negative, na.rm = TRUE),
    atlas_net_support_prop =
      .data$atlas_prop_supported_positive - .data$atlas_prop_supported_negative,
    .groups = "drop"
  )

guild_pair_summary <- omega_pairs |>
  group_by(
    .data$atlas, .data$period, .data$guild_pair_type,
    .data$guild_1, .data$guild_2, .data$guild_pair
  ) |>
  summarise(
    n_pairs = n(),
    mean_omega = mean(.data$omega_mean, na.rm = TRUE),
    median_omega = median(.data$omega_mean, na.rm = TRUE),
    mean_abs_omega = mean(abs(.data$omega_mean), na.rm = TRUE),
    prop_supported_positive = mean(.data$is_supported_positive, na.rm = TRUE),
    prop_supported_negative = mean(.data$is_supported_negative, na.rm = TRUE),
    net_support_prop =
      .data$prop_supported_positive - .data$prop_supported_negative,
    .groups = "drop"
  ) |>
  left_join(atlas_averages, by = c("atlas", "period")) |>
  left_join(
    guild_sizes |> rename(guild_1 = guild, n_species_guild_1 = n_species),
    by = "guild_1"
  ) |>
  left_join(
    guild_sizes |> rename(guild_2 = guild, n_species_guild_2 = n_species),
    by = "guild_2"
  ) |>
  mutate(
    mean_omega_minus_atlas_average = .data$mean_omega - .data$atlas_mean_omega,
    prop_positive_minus_atlas_average =
      .data$prop_supported_positive - .data$atlas_prop_supported_positive,
    prop_negative_minus_atlas_average =
      .data$prop_supported_negative - .data$atlas_prop_supported_negative,
    net_support_minus_atlas_average =
      .data$net_support_prop - .data$atlas_net_support_prop,
    unstable_small_pair_count =
      .data$n_pairs < min_pairs_for_stable_guild_summary
  ) |>
  arrange(.data$atlas, .data$guild_pair_type, desc(abs(.data$mean_omega_minus_atlas_average)))

within_guild_summary <- guild_pair_summary |>
  filter(.data$guild_pair_type == "Within guild") |>
  arrange(.data$atlas, desc(abs(.data$mean_omega_minus_atlas_average)))

write_csv(
  guild_pair_summary,
  file.path(out_dir, paste0(pattern, "-omega-guild-pair-observed-summary.csv"))
)
write_csv(
  within_guild_summary,
  file.path(out_dir, paste0(pattern, "-omega-within-guild-observed-summary.csv"))
)

kable(
  within_guild_summary |>
    select(
      period, guild = guild_1, n_species_guild_1,
      n_pairs, mean_omega, mean_omega_minus_atlas_average,
      prop_supported_positive, prop_supported_negative,
      unstable_small_pair_count
    ),
  digits = 3,
  caption = "Within-guild observed Omega summaries. Small pair counts are flagged because they are unstable."
)
```

## Guild Permutation Null

The null keeps the Omega matrix fixed and randomly shuffles guild labels among
species. This preserves the observed guild-size distribution, while asking
whether a named guild or guild pair has stronger or weaker association than
expected if guild identity were unrelated to the Omega structure.

```{r guild-permutation-null}
observed_guild_counts <- sort(table(species_traits$guild))
permuted_guild_counts <- sort(table(sample(species_traits$guild)))
stopifnot(identical(observed_guild_counts, permuted_guild_counts))

guild_size_check <- tibble(
  guild = names(sort(table(species_traits$guild), decreasing = TRUE)),
  observed_n_species = as.integer(sort(table(species_traits$guild), decreasing = TRUE)),
  null_preserves_size = TRUE
)

omega_pair_core <- omega_pairs |>
  select(
    atlas, period, species_a, species_b,
    omega_mean, is_supported_positive,
    is_supported_negative
  )

species_for_permutation <- species_traits$species
guild_labels_for_permutation <- species_traits$guild

summarise_permuted_guild_pairs <- function(iteration) {
  # Sampling the guild-label vector without replacement preserves the exact
  # number of species in every guild for each null iteration.
  permuted_guilds <- sample(guild_labels_for_permutation)
  names(permuted_guilds) <- species_for_permutation

  permuted_guild_a <- unname(permuted_guilds[omega_pair_core$species_a])
  permuted_guild_b <- unname(permuted_guilds[omega_pair_core$species_b])
  permuted_guild_1 <- pmin(permuted_guild_a, permuted_guild_b)
  permuted_guild_2 <- pmax(permuted_guild_a, permuted_guild_b)

  omega_pair_core |>
    transmute(
      permutation = iteration,
      atlas = .data$atlas,
      period = .data$period,
      guild_1 = permuted_guild_1,
      guild_2 = permuted_guild_2,
      guild_pair = if_else(
        .data$guild_1 == .data$guild_2,
        .data$guild_1,
        paste(.data$guild_1, .data$guild_2, sep = " | ")
      ),
      guild_pair_type = if_else(
        .data$guild_1 == .data$guild_2,
        "Within guild",
        "Between guild"
      ),
      omega_mean = .data$omega_mean,
      is_supported_positive = .data$is_supported_positive,
      is_supported_negative = .data$is_supported_negative
    ) |>
    group_by(
      .data$permutation, .data$atlas, .data$period,
      .data$guild_pair_type, .data$guild_1, .data$guild_2, .data$guild_pair
    ) |>
    summarise(
      n_pairs = n(),
      mean_omega = mean(.data$omega_mean, na.rm = TRUE),
      prop_supported_positive = mean(.data$is_supported_positive, na.rm = TRUE),
      prop_supported_negative = mean(.data$is_supported_negative, na.rm = TRUE),
      net_support_prop =
        .data$prop_supported_positive - .data$prop_supported_negative,
      .groups = "drop"
    )
}

permutation_summaries <- map_dfr(
  seq_len(n_permutations),
  summarise_permuted_guild_pairs
)

observed_guild_metrics <- guild_pair_summary |>
  select(
    atlas, period, guild_pair_type,
    guild_1, guild_2, guild_pair,
    n_pairs, unstable_small_pair_count,
    mean_omega, prop_supported_positive,
    prop_supported_negative, net_support_prop
  ) |>
  pivot_longer(
    cols = c(
      mean_omega, prop_supported_positive,
      prop_supported_negative, net_support_prop
    ),
    names_to = "metric",
    values_to = "observed_value"
  )

permuted_guild_metrics <- permutation_summaries |>
  # The observed pair count is the one we want to report and flag, so the
  # permutation-side count is dropped before joining to avoid .x/.y suffixes.
  select(-n_pairs) |>
  pivot_longer(
    cols = c(
      mean_omega, prop_supported_positive,
      prop_supported_negative, net_support_prop
    ),
    names_to = "metric",
    values_to = "permuted_value"
  )

guild_null_results <- permuted_guild_metrics |>
  inner_join(
    observed_guild_metrics,
    by = c(
      "atlas", "period", "guild_pair_type", "guild_1",
      "guild_2", "guild_pair", "metric"
    )
  ) |>
  group_by(
    .data$atlas, .data$period, .data$guild_pair_type,
    .data$guild_1, .data$guild_2, .data$guild_pair,
    .data$n_pairs, .data$unstable_small_pair_count,
    .data$metric, .data$observed_value
  ) |>
  summarise(
    expected_null_mean = mean(.data$permuted_value, na.rm = TRUE),
    expected_null_sd = sd(.data$permuted_value, na.rm = TRUE),
    null_lower_95 = quantile(.data$permuted_value, 0.025, na.rm = TRUE),
    null_upper_95 = quantile(.data$permuted_value, 0.975, na.rm = TRUE),
    empirical_p_upper =
      (sum(.data$permuted_value >= .data$observed_value, na.rm = TRUE) + 1) /
      (n() + 1),
    empirical_p_lower =
      (sum(.data$permuted_value <= .data$observed_value, na.rm = TRUE) + 1) /
      (n() + 1),
    .groups = "drop"
  ) |>
  mutate(
    observed_minus_expected = .data$observed_value - .data$expected_null_mean,
    permutation_z = if_else(
      .data$expected_null_sd > 0,
      .data$observed_minus_expected / .data$expected_null_sd,
      NA_real_
    ),
    empirical_p_two_sided = pmin(
      1,
      2 * pmin(.data$empirical_p_lower, .data$empirical_p_upper)
    )
  ) |>
  group_by(.data$metric) |>
  mutate(
    fdr_p = p.adjust(.data$empirical_p_two_sided, method = "fdr"),
    fdr_0_05 = .data$fdr_p < 0.05
  ) |>
  ungroup() |>
  arrange(.data$metric, .data$atlas, .data$fdr_p)

write_csv(
  guild_null_results,
  file.path(out_dir, paste0(pattern, "-omega-guild-permutation-null-results.csv"))
)
write_csv(
  guild_size_check,
  file.path(out_dir, paste0(pattern, "-omega-guild-null-size-check.csv"))
)

kable(
  guild_null_results |>
    filter(.data$metric == "mean_omega", .data$guild_pair_type == "Within guild") |>
    arrange(.data$period, .data$fdr_p) |>
    select(
      period, guild = guild_1, n_pairs,
      observed_value, expected_null_mean,
      observed_minus_expected, permutation_z,
      empirical_p_two_sided, fdr_p,
      unstable_small_pair_count
    ),
  digits = 3,
  caption = "Within-guild mean Omega compared with the guild-label permutation null."
)
```

## Plots

```{r plotting-helpers}
save_png <- function(plot, filename, width, height, dpi = 300) {
  # This guard prevents accidental PDF/SVG output from this notebook.
  if (tolower(tools::file_ext(filename)) != "png") {
    stop("Plot filename must end in .png: ", filename, call. = FALSE)
  }

  path <- file.path(plot_dir, filename)
  ggsave(
    filename = path,
    plot = plot,
    width = width,
    height = height,
    dpi = dpi,
    bg = "white"
  )
  path
}

species_names <- omega_pairs |>
  filter(.data$atlas == "1") |>
  summarise(species = list(unique(c(.data$species_a, .data$species_b)))) |>
  pull(species) |>
  pluck(1)

omega_matrix_from_pairs <- function(df, species_order) {
  mat <- matrix(
    NA_real_,
    nrow = length(species_order),
    ncol = length(species_order),
    dimnames = list(species_order, species_order)
  )
  mat[cbind(df$species_a, df$species_b)] <- df$omega_mean
  mat[cbind(df$species_b, df$species_a)] <- df$omega_mean
  diag(mat) <- 1
  mat
}

atlas1_matrix <- omega_matrix_from_pairs(
  omega_pairs |> filter(.data$atlas == "1"),
  species_names
)
species_plot_order <- rownames(atlas1_matrix)[hclust(dist(atlas1_matrix))$order]

make_symmetric_species_tiles <- function(df, value_col, value_name = "value") {
  value_col <- rlang::ensym(value_col)

  bind_rows(
    df |>
      transmute(
        atlas = .data$atlas,
        period = .data$period,
        species_row = .data$species_a,
        species_col = .data$species_b,
        value = !!value_col
      ),
    df |>
      transmute(
        atlas = .data$atlas,
        period = .data$period,
        species_row = .data$species_b,
        species_col = .data$species_a,
        value = !!value_col
      )
  ) |>
    mutate(
      species_row = factor(.data$species_row, levels = rev(species_plot_order)),
      species_col = factor(.data$species_col, levels = species_plot_order)
    ) |>
    rename(!!value_name := value)
}

guild_order <- guild_sizes$guild

make_symmetric_guild_tiles <- function(df, value_col, value_name = "value") {
  value_col <- rlang::ensym(value_col)

  bind_rows(
    df |>
      transmute(
        atlas = .data$atlas,
        period = .data$period,
        guild_row = .data$guild_1,
        guild_col = .data$guild_2,
        value = !!value_col
      ),
    df |>
      filter(.data$guild_1 != .data$guild_2) |>
      transmute(
        atlas = .data$atlas,
        period = .data$period,
        guild_row = .data$guild_2,
        guild_col = .data$guild_1,
        value = !!value_col
      )
  ) |>
    mutate(
      guild_row = factor(.data$guild_row, levels = rev(guild_order)),
      guild_col = factor(.data$guild_col, levels = guild_order)
    ) |>
    rename(!!value_name := value)
}

support_colours <- c(
  "Supported positive" = "#b2182b",
  "Supported negative" = "#2166ac",
  "Unsupported" = "grey80"
)
```

```{r species-omega-heatmaps, fig.width=12, fig.height=4}
omega_heatmap_data <- make_symmetric_species_tiles(
  omega_pairs,
  omega_mean,
  value_name = "omega_mean"
)

p_omega_heatmaps <- ggplot(
  omega_heatmap_data,
  aes(x = .data$species_col, y = .data$species_row, fill = .data$omega_mean)
) +
  geom_tile() +
  facet_wrap(~ period, nrow = 1) +
  scale_fill_gradient2(
    low = "#2166ac",
    mid = "white",
    high = "#b2182b",
    midpoint = 0,
    limits = c(-1, 1),
    oob = squish,
    name = "Mean Omega"
  ) +
  coord_equal(expand = FALSE) +
  labs(x = NULL, y = NULL) +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    panel.grid = element_blank()
  )

p_omega_heatmaps
```

```{r species-difference-heatmaps, fig.width=9, fig.height=4}
difference_pairs <- temporal_change |>
  select(
    pair_id, species_a, species_b,
    delta_1990s_minus_1970s, delta_2010s_minus_1970s
  ) |>
  pivot_longer(
    cols = starts_with("delta_"),
    names_to = "comparison",
    values_to = "omega_delta"
  ) |>
  mutate(
    period = recode(
      .data$comparison,
      "delta_1990s_minus_1970s" = "1990s - 1970s",
      "delta_2010s_minus_1970s" = "2010s - 1970s"
    ),
    atlas = .data$period
  )

difference_heatmap_data <- make_symmetric_species_tiles(
  difference_pairs,
  omega_delta,
  value_name = "omega_delta"
)
delta_limit <- max(abs(difference_heatmap_data$omega_delta), na.rm = TRUE)

p_difference_heatmaps <- ggplot(
  difference_heatmap_data,
  aes(x = .data$species_col, y = .data$species_row, fill = .data$omega_delta)
) +
  geom_tile() +
  facet_wrap(~ period, nrow = 1) +
  scale_fill_gradient2(
    low = "#2166ac",
    mid = "white",
    high = "#b2182b",
    midpoint = 0,
    limits = c(-delta_limit, delta_limit),
    oob = squish,
    name = "Delta mean Omega"
  ) +
  coord_equal(expand = FALSE) +
  labs(x = NULL, y = NULL) +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    panel.grid = element_blank()
  )

p_difference_heatmaps
```

```{r support-class-bars, fig.width=7, fig.height=4}
p_support_bars <- ggplot(
  support_summary,
  aes(x = .data$period, y = .data$prop_pairs, fill = .data$support_class)
) +
  geom_col(width = 0.75, colour = "white", linewidth = 0.2) +
  scale_fill_manual(values = support_colours, name = NULL) +
  scale_y_continuous(labels = percent_format()) +
  labs(x = NULL, y = "Species-pair proportion") +
  theme(legend.position = "bottom")

p_support_bars
```

```{r top-changing-trajectories, fig.width=12, fig.height=8}
top_changing_pair_ids <- temporal_change |>
  slice_max(.data$abs_delta_2010s_minus_1970s, n = 20, with_ties = FALSE) |>
  pull(pair_id)

top_changing_long <- omega_pairs |>
  filter(.data$pair_id %in% top_changing_pair_ids) |>
  mutate(
    pair_label = paste(.data$species_a, .data$species_b, sep = " - "),
    pair_label = fct_reorder(
      .data$pair_label,
      abs(.data$omega_mean),
      .fun = max,
      .desc = TRUE
    )
  )

p_top_changing <- ggplot(
  top_changing_long,
  aes(x = .data$period, y = .data$omega_mean, group = .data$pair_id)
) +
  geom_hline(yintercept = 0, colour = "grey75", linewidth = 0.3) +
  geom_line(colour = "grey35", linewidth = 0.5) +
  geom_point(aes(fill = .data$support_class), shape = 21, size = 2.4, colour = "grey20") +
  facet_wrap(~ pair_label, ncol = 4) +
  scale_fill_manual(values = support_colours, name = NULL) +
  labs(x = NULL, y = "Mean Omega") +
  theme(
    legend.position = "bottom",
    strip.text = element_text(size = 7)
  )

p_top_changing
```

```{r guild-heatmaps, fig.width=13, fig.height=10}
guild_mean_heatmap_data <- make_symmetric_guild_tiles(
  guild_pair_summary,
  mean_omega,
  value_name = "mean_omega"
)

p_guild_mean_heatmaps <- ggplot(
  guild_mean_heatmap_data,
  aes(x = .data$guild_col, y = .data$guild_row, fill = .data$mean_omega)
) +
  geom_tile(colour = "white", linewidth = 0.1) +
  facet_wrap(~ period, nrow = 1) +
  scale_fill_gradient2(
    low = "#2166ac",
    mid = "white",
    high = "#b2182b",
    midpoint = 0,
    name = "Mean Omega"
  ) +
  coord_equal(expand = FALSE) +
  labs(x = NULL, y = NULL) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 6),
    axis.text.y = element_text(size = 6),
    panel.grid = element_blank()
  )

p_guild_mean_heatmaps
```

```{r guild-null-z-heatmaps, fig.width=13, fig.height=10}
guild_z_heatmap_data <- guild_null_results |>
  filter(.data$metric == "mean_omega") |>
  make_symmetric_guild_tiles(permutation_z, value_name = "permutation_z")

z_limit <- quantile(abs(guild_z_heatmap_data$permutation_z), 0.99, na.rm = TRUE)

p_guild_z_heatmaps <- ggplot(
  guild_z_heatmap_data,
  aes(x = .data$guild_col, y = .data$guild_row, fill = .data$permutation_z)
) +
  geom_tile(colour = "white", linewidth = 0.1) +
  facet_wrap(~ period, nrow = 1) +
  scale_fill_gradient2(
    low = "#2166ac",
    mid = "white",
    high = "#b2182b",
    midpoint = 0,
    limits = c(-z_limit, z_limit),
    oob = squish,
    name = "Null z"
  ) +
  coord_equal(expand = FALSE) +
  labs(x = NULL, y = NULL) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 6),
    axis.text.y = element_text(size = 6),
    panel.grid = element_blank()
  )

p_guild_z_heatmaps
```

```{r within-guild-lollipop, fig.width=10, fig.height=9}
within_guild_null <- guild_null_results |>
  filter(
    .data$metric == "mean_omega",
    .data$guild_pair_type == "Within guild"
  ) |>
  mutate(
    guild = .data$guild_1,
    guild = fct_reorder(.data$guild, .data$observed_minus_expected),
    fdr_flag = if_else(.data$fdr_0_05, "FDR < 0.05", "Not FDR flagged")
  )

p_within_lollipop <- ggplot(
  within_guild_null,
  aes(y = .data$guild)
) +
  geom_vline(xintercept = 0, colour = "grey80", linewidth = 0.3) +
  geom_segment(
    aes(
      x = .data$expected_null_mean,
      xend = .data$observed_value,
      yend = .data$guild
    ),
    colour = "grey55",
    linewidth = 0.4
  ) +
  geom_point(
    aes(x = .data$expected_null_mean),
    shape = 21,
    fill = "white",
    colour = "grey30",
    size = 2
  ) +
  geom_point(
    aes(x = .data$observed_value, colour = .data$fdr_flag),
    size = 2.4
  ) +
  facet_wrap(~ period, nrow = 1) +
  scale_colour_manual(
    values = c("FDR < 0.05" = "#b2182b", "Not FDR flagged" = "grey35"),
    name = NULL
  ) +
  labs(x = "Mean Omega: open point = null expectation, filled point = observed", y = NULL) +
  theme(legend.position = "bottom")

p_within_lollipop
```

```{r save-plots-and-checks}
saved_plot_paths <- c(
  save_png(
    p_omega_heatmaps,
    paste0(pattern, "-omega-heatmaps-by-atlas.png"),
    width = 12,
    height = 4
  ),
  save_png(
    p_difference_heatmaps,
    paste0(pattern, "-omega-difference-heatmaps.png"),
    width = 9,
    height = 4
  ),
  save_png(
    p_support_bars,
    paste0(pattern, "-omega-support-class-bars.png"),
    width = 7,
    height = 4
  ),
  save_png(
    p_top_changing,
    paste0(pattern, "-omega-top-changing-trajectories.png"),
    width = 12,
    height = 8
  ),
  save_png(
    p_guild_mean_heatmaps,
    paste0(pattern, "-omega-guild-mean-heatmaps.png"),
    width = 13,
    height = 10
  ),
  save_png(
    p_guild_z_heatmaps,
    paste0(pattern, "-omega-guild-null-z-heatmaps.png"),
    width = 13,
    height = 10
  ),
  save_png(
    p_within_lollipop,
    paste0(pattern, "-omega-within-guild-observed-vs-null.png"),
    width = 10,
    height = 9
  )
)

plot_output_check <- tibble(
  path = saved_plot_paths
) |>
  mutate(
    exists = file.exists(.data$path),
    extension = tools::file_ext(.data$path),
    is_png = tolower(.data$extension) == "png"
  )

stopifnot(all(plot_output_check$exists), all(plot_output_check$is_png))

kable(
  plot_output_check,
  caption = "Saved plot check. Every manually saved plot from this notebook is a PNG."
)
```

## Reusable Outputs

```{r output-listing}
output_files <- tibble(
  path = list.files(out_dir, recursive = TRUE, full.names = TRUE)
) |>
  mutate(
    file = basename(.data$path),
    extension = tools::file_ext(.data$path)
  ) |>
  arrange(.data$extension, .data$file)

kable(output_files, caption = "Reusable CSV and PNG outputs written by this notebook.")
```
